\section{Availability}\label{s:availability}

The CVODES package has been released under a BSD open source license and is 
freely available at the web site
{\bf www.llnl.gov/CASC/sundials},
or through the DOE ACTS web site at
{\bf acts.nersc.gov/sundials/main.html}.

\section{Conclusions}\label{s:conclusions}

CVODES is the first in a series of new additions to SUNDIALS. The new codes,
IDAS and KINSOLS, together with CVODES, will provide sensitivity analysis
for all the classes of problems addressed by the basic SUNDIALS solvers.
These new capabilities extend the versatility and functionality of the SUNDIALS
solvers in addressing new classes of applications, such as dynamically-constrained
optimization, inversion, and uncertainty quantification.

Like all of SUNDIALS, CVODES is under active development.
An area of particular interest is in the automatic generation of the
sensitivity equations. A parser and code generator for the automatic
generation of derivative approximations using the complex step method
is underway.
Automatic differentiation (AD) tools will be incorporated as they
become available; we are especially interested in adding reverse AD
capabilities to the SUNDIALS adjoint sensitivity solvers.
We are currently investigating alternatives to checkpointing within
the adjoint solver in CVODES: one direction is in using reduced order
models of the forward problem, while another is in storing the
complete decision history on the first forward pass and re-using it on
the second pass.
Finally, to address language interoperability issues and thus
facilitate the use of the SUNDIALS solvers for users of other
programming languages, we plan to generate Babel~\cite{KKPR:01}
wrappers for them.
 
