\documentstyle[12pt]{letter}
\batchmode
\begin{document}
\begin{letter}


\reference{Manuscript TOMS-2003-0041}

\opening{Dear Dr. Boisvert,}

blah, blah, blah

\closing{Sincerely,}

Radu Serban and Alan C. Hindmarsh\\
Center for Applied Scientific Computing\\
Lawrence Livermore National Laboratory

\end{letter}

\newpage
\section{Answers to Referee 1 Comments}

This paper describes an ODE solver with sensitivity analysis 
capabilities. Sensitivity analysis is an extremely important
subject that can be applied in a number of situations such as
optimum design, experimental design, uncertainty analysis etc. 

This paper is very well written and can be decomposed into two
main parts: algorithms and code description. It is difficult to
find something new in the algorithm part. The authors should
give proper citations on the things they have used. For example, on
page 4, the quadrature computation has been proposed and
documented in DASPK3.0 package (see Li and Petzold (2000 and
2001), Cao et al (2002)). 

The evaluation of the sensitivity right-hand side is the key
part for users who want to use the sensitivity analysis
capabilities. I suggest the authors describe and compare all the
approaches, not only the finite-difference approaches, and give
appropriate citations. 

{\em Response:}
The mechanism provided in CVODES for user-defined routines
for the evaluation of derivative information (Jacobian-related data
and/or sensitivity right-hand sides) can be used to wrap AD-generated
routines.
We have included a paragraph discussing our own efforts 
in developing complex-step derivative approximations through automatic 
code generation and a short discussion on the challenges in 
designing general interfaces to automatic differentiation tools
for C codes.

In section 2.3, the authors ignore that the check-pointing
scheme has also been used in other adjoint packages (e.g.,
Giering, TAMC) and DAE solver DASPKADJOINT code (see Cao et
al.(2002), Li and Petzold(2001)). 

{\em Response:} 
we have changed this paragraph to mention that we are 
using a checkpointing scheme similar to the one in DASPKADJOINT
and a justification why truly optimal checkpointing (as described
by Griewank and Walther) cannot be employed.

The authors do not mention how to evaluate the adjoint sensitivity 
right-hand side (15) or (17) in the paper. This is the key for the users. 
I find that in the code, the analytic input by hand is used, which is impossible
for many of users and applications. 

The last sentence about the adjoint PDE on page 10 is wrong. 
You cannot replace the adjoint of the discretized PDE with the 
discretized adjoint PDE system directly without modifying the 
sensitivity evaluation formula (14). The boundary conditions, 
which is absent in ODE solver and formula (14), must be incorporated 
(see Li and Petzold (2003) about adjoint PDE). 

{\em Response:}
 We are aware that these two formulations are different and this
is what we were trying to say. Exactly because of this, we designed 
CVODES such that it does not assume one formulation or the other.
We have now explicitely stated that the two formulations are different
(with two citations).

The code description part is quite clear. I would like to see
the following points to be addressed. 

1) Since there are many FORTRAN ODE/DAE codes available, the authors need to 
state explicitly what the advantages are to rewrite the code to C or C++
language. 

{\em Response:}
We have included a paragraph in the introduction that lists
some of the advantages of using C: portability, availability, standard
dynamic memory allocation and data structures, etc.

2) Many numerical users are still using Fortran in writing their right-hand 
side. The authors need to give a clear and easy-used interface between the 
Fortran user and CVODES package. 

{\em Response:}
In the introduction, we point to the Fortran-C interface provided
with CVODE as an example of using a SUNDIALS solver with Fortran user code.
Moreover, as mentioned in the Conlcusions section, we are planning on
providing interlanguage operability for all of the SUNDIALS codes by using
Babel technology.

3) The authors need to describe how to incorporate efficiently the existing 
nonlinear or linear solvers. For example, many PDE packages now use the sparse 
data structure and multi-grid preconditioners. How to use the modern solver
techniques  in CVODES is an interesting field. The author need
to give some related examples in the code. 

4) The authors do not state much about the parallelization. I would like to 
see more about that topic in the paper and numerical efficiency and speed-up
comparison with the serial code in the packages.

{\em Response:}
We have included a section on parallel speedup on a sensitivity problem.


\newpage
\section{Answers to Referee 2 Comments}

This paper describes a stiff and nonstiff ODE IVP solver with forward and adjoint sensitivity
analysis capabilities. The presented software, which is a part of SUNDIALS software suite, is an
updated version of the VODE solver written in C and extended to sensitivity analysis capabilities.
All the algorithms presented have been published elsewhere by other workers, and cannot be seen
as novel contributions of the current paper. Therefore the first 10 pages of the paper can be
summarized with appropriate citations. The rest of the paper describes the code organization and
its usage briefly. There are no numerical examples demonstrating any advantages of using this
particular software. There is no mention of existing forward/adjoint sensitivity analysis tool(s)
with virtually identical capabilities, i.e., DASPKADJOINT.

Considering the editorial charter which states that "..., papers that are essentially user manuals
for a computer program are also not acceptable" and noting that most of the material presented
in the paper can be found in [1], a significant rewrite of the paper is recommended emphasizing
the advantages and novelties of the software with specific examples. However, it is not clear to this
reviewer that there are sufficient advantages and novelties in the software to satisfy the requirements
of the editorial charter.

Detailed Comments
* p. 7 paragraph 3: while it is indeed an important observation that the staggered corrector
method combined with an iterative linear solver is equivalent to the staggered direct method,
this observation has already been made in [2].

{\em Response:}
We have added the citation.

* p. 14 paragraph 3 from bottom of the page: SPGMR is not a direct linear solver.

{\em Response:}
We have rephrased this sentence.

* p. 15 paragraph 2:"sensitvity" should be "sensitivity".

{\em Response:}
fixed the typo.

* the Maly and Petzold reference is from 1996.

{\em Response: fixed biblio entry.}

Experience with the Code
The code is user friendly and has obvious advantages inherited from the usage of the C programming
language. We did not undertake performance comparisons.

\newpage
\section{Answers to Referee 3 Comments}

1 Summary of the contents
The paper describes the software tool CVODES as a multistep ODE solver
with comprehensive capabilities for sensitivity analysis. The ODE integra-
tion itself is performed by a BDF method for stiff systems or an Adams-
Moulton formula for non-stiff systems. Arising linear systems can be solved
by either a dense direct method, a banded direct method or an iterative
Krylov subspace method.
The sensitivity analysis features of CVODES comprise forward sensitiv-
ity analysis by the simultaneous corrector method as described in [5], or
two versions of the staggered corrector method as introduced in [1]. The
right-hand sides of the arising sensitivity systems can be currently obtained
through analytical user input, finite differences or directional derivatives,
and an automatic differentiation option is envisaged in the conclusion sec-
tion.Furthermore, CVODES also allows to compute sensitivities via adjoint
sensitivity analysis by implementing a check-pointing strategy.
In Section 3, the paper describes details of the software setup and code
organization, as part of a larger family of solvers, the SUNDIALS package.
Section 4 of the paper is concerned with giving a detailed overview about
the usage of the tool.

2 Comments on the contents and assessment of relevance for a publication in TOMS
The features of the tool mentioned in this paper are apparently not new. A
BDF solver for DAE systems with dense direct, banded direct or Krylov sub-
space based linear solvers is available as the well-established code DDASPK
[2]. DDASPK also contains the three common methods for forward sensi-
tivity analysis, i.e. the staggered direct, the simultaneous corrector and the
staggered corrector method. The code DSL48S [1] contains a BDF solver
for DAE systems with the staggered corrector sensitivity method using the
sparse direct linear algebra solver MA48. 

The statement in Section 2.2.1 of the paper, that the staggered direct method 
is always unattractive, is not correct. As shown in [4], it can outperform the 
other methods in case that Ns >>  N. 

{\em Response:}
We have changed this paragraph to include the citation [4] and 
clarify that the staggered direct method is not attractive for problems that 
have many states (such as those arising from semidiscretization of PDES).
For the cases in which the number of parameters is very large (comparable
to the problem size) adjoint sensitivity is (generally) a better choice than 
any forward sensitivity method.

In contrast to CVODES, DDASPK already contains automatic differentiation capabilities 
according to [2].

Regarding the adjoint approach, the concepts and implementation in
this paper appear to be close to what has been reported in [3]. The check-
pointing has already been described in that reference. 

{\em Response:}
We have changed this paragraph to mention that we are 
using a checkpointing scheme similar to the one in DASPKADJOINT
and a justification why truly optimal checkpointing (as described
by Griewank and Walther) cannot be employed.

DDASPKADJOINT additionally contains optional automatic differentiation facilities 
for obtaining the required derivative information.

According to the paper, the key benefit of CVODES appears to be the
modular software implementation, which collects all subelements for ODE
integration and sensitivity evaluation under one common roof. The modular
approach allows an easy exchange of low-level tools such as linear solvers
etc.The editorial charter of TOMS states that "the purpose of a TOMS
communication is the presentation of the results of novel research and de-
velopment efforts in support of significant mathematical computer applica-
tion". Taking the findings above into account, it is rarely possible to accept
the paper as "novel research". An acceptable paper should also contain "a
demonstration of superiority compared to alternative approaches", typically
carried out by an experimental evaluation of computer implementations.
The paper at hand is completely lacking this issue. Basically it is a descrip-
tion of a new software implementation of well-known numerical concepts.


\end{document}