% LaTeX source for joint paper on SUNDIALS
% Version of 18 August 2003

\documentclass[acmtoms]{acmtrans2m}
\usepackage{amsmath, amssymb}
\usepackage{epsfig}
\usepackage{tabularx}
\usepackage{showlabels}

\newcommand{\id}[1]{{\tt #1}}

\newcommand{\dfdy}{\frac{\partial f}{\partial y}}
\newcommand{\dfdyI}{\partial f / \partial y}
\newcommand{\dfdpi}{\frac{\partial f}{\partial p_i}}
\newcommand{\dfdpiI}{\partial f / \partial p_i}

\acmVolume{V}
\acmNumber{N}
\acmYear{Y}
\acmMonth{Month}

\markboth{R. Serban and A.C. Hindmarsh}
{CVODES: An ODE solver with sensitivity analysis capabilities}

\title{CVODES: An ODE solver with sensitivity analysis capabilities}

\author{RADU SERBAN and ALAN C. HINDMARSH \\
  Center for Applied Scientific Computing \\
  Lawrence Livermore National Laboratory}

\begin{abstract}
CVODES, which is part of the SUNDIALS software suite,
is a stiff and nonstiff ordinary differential equation initial 
value problem solver with sensitivity analysis capabilities. 
CVODES is written in a data-independent manner, with a highly 
modular structure to allow incorporation of different 
preconditioning and/or linear solver methods. It shares with the 
other SUNDIALS solvers several common modules, most notably the 
generic kernel of vector operations and a set of generic linear 
solvers and preconditioners.

CVODES solves the IVP by one of two methods -- backward differentiation 
formula or Adams-Moulton -- both implemented in a variable-step, 
variable-order form. The forward sensitivity module in CVODES implements 
the simultaneous corrector method, as well as two flavors of 
staggered corrector methods. Its adjoint sensitivity module 
provides a combination of checkpointing and cubic Hermite interpolation 
for the efficient generation of the forward solution during the adjoint 
system integration.

We describe the current capabilities of CVODES, its design principles 
and connection to the SUNDIALS suite, and the user interface. 
Finally, we mention current and future development efforts for CVODES, 
particularly in the direction of automatic generation of the sensitivity 
right hand sides using automatic differentiation and/or complex-step 
techniques.
\end{abstract}

\category{G.4}{Mathematical Software}{Algorithm design and analysis}

\category{G.1.7}{Numerical Analysis}{Ordinary Differential Equations}
[Initial value problems \and Multistep methods \and Stiff equations]

%\category{G.1.m}{Miscellaneous}{Sensitivity Analysis}
%[Forward methods \and Adjoint methods]

\terms{Algorithms, Design}

\keywords{ODEs, Forward Sensitivity Analysis, Adjoint Sensitivity Analysis}

%=======================================================================

\begin{document}

\setcounter{page}{1}

\begin{bottomstuff}
Authors' address: 
Radu Serban, 
Lawrence Livermore National Laboratory, P.O. Box 808, L-560,
Livermore, CA 94551; email: {\tt radu@llnl.gov};
Alan C. Hindmarsh,
Lawrence Livermore National Laboratory, P.O. Box 808, L-560,
Livermore, CA 94551; email: {\tt alanh@llnl.gov};
\newline
This work was performed under the auspices of the
U.S. Department of Energy by the University of California,
Lawrence Livermore National Laboratory, under contract No.
W-7405-Eng-48.
\end{bottomstuff}

\maketitle

%=======================================================================
% Include sections

\input{introduction.tex}
\input{algorithms.tex}
\input{organization.tex}
\input{usage.tex}
\input{conclusions.tex}

%=======================================================================
% Bibliography
\bibliographystyle{acmtrans}
\bibliography{../biblio}

%=======================================================================
% The article should end by the following lines. 
% The actual dates will be supplied by the Editor-in-Chief.

\begin{received}
Received Month Year;
revised Month Year; accepted Month Year
\end{received}

\end{document}
