%===================================================================================
\chapter{Introduction}\label{s:intro}
%===================================================================================

{\cvode} is part of a software family called {\sundials}: 
SUite of Nonlinear and DIfferential/ALgebraic equation Solvers.  
This suite consists of {\cvode}, {\kinsol}, and {\ida}, and variants of these
with sensitivity analysis capabilities.
%
%---------------------------------
\section{Historical Background}\label{ss:history}
%---------------------------------

\index{CVODE@{\cvode}!relationship to {\vode}, {\vodpk}|(}
{\F} solvers for ODE initial value problems are widespread and heavily used. 
Two solvers that have been written at LLNL in the past are {\vode} \cite{BBH:89} 
and {\vodpk} \cite{Byr:92}.
{\vode}\index{VODE@{\vode}} is a general purpose solver that includes methods for
stiff and nonstiff systems, and in the stiff case uses direct methods (full or
banded) for the solution of the linear systems that arise at each implicit
step. Externally, {\vode} is very similar to the well known solver
{\lsode}\index{LSODE@{\lsode}} \cite{RaHi:94}. {\vodpk}\index{VODPK@{\vodpk}}
is a variant of {\vode} that uses a preconditioned Krylov (iterative)
method, namely GMRES, for the solution of the linear systems. {\vodpk}
is a powerful tool for large stiff systems because it combines
established methods for stiff integration, nonlinear iteration, and
Krylov (linear) iteration with a problem-specific treatment of the
dominant source of stiffness, in the form of the user-supplied
preconditioner matrix \cite{BrHi:89}.  The capabilities of both
{\vode} and {\vodpk} have been combined in the {\C}-language package
{\cvode}\index{CVODE@{\cvode}} \cite{CoHi:96}.

At present, {\cvode} contains three Krylov methods that can be used
in conjuction with Newton iteration:
the GMRES (Generalized Minimal RESidual) \cite{SaSc:86},
Bi-CGStab (Bi-Conjugate Gradient Stabilized) \cite{Van:92}, and
TFQMR (Transpose-Free Quasi-Minimal Residual) linear iterative methods
\cite{Fre:93}.  As Krylov methods, these require almost no matrix storage
for solving the Newton equations as compared to direct methods.
However, the algorithms allow for a user-supplied preconditioner
matrix, and for most problems preconditioning is essential for an
efficient solution.
For very large stiff ODE systems, the Krylov methods are preferable over
direct linear solver methods, and are often the only feasible choice.
Among the three Krylov methods in {\cvode}, we recommend GMRES as the
best overall choice.  However, users are encouraged to compare all
three, especially if encountering convergence failures with GMRES.
Bi-CGFStab and TFQMR have an advantage in storage requirements, in
that the number of workspace vectors they require is fixed, while that
number for GMRES depends on the desired Krylov subspace size.

In the process of translating the {\vode} and {\vodpk} algorithms into
{\C}, the overall {\cvode} organization has been changed considerably.
One key feature of the {\cvode} organization is that the linear system
solvers comprise a layer of code modules that is separated from the
integration algorithm, allowing for easy modification and expansion of
the linear solver array.  A second key feature is a separate module
devoted to vector operations; this facilitated the extension to
multiprosessor environments with minimal impacts on the rest of the
solver, resulting in {\pvode}\index{PVODE@{\pvode}} \cite{ByHi:99},
the parallel variant of {\cvode}.  \index{CVODE@{\cvode}!relationship
to {\vode}, {\vodpk}|)}

\index{CVODE@{\cvode}!relationship to {\cvode}, {\pvode}|(} Recently,
the functionality of {\cvode} and {\pvode} has been combined into one
single code, simply called {\cvode}. Development of the new version of
{\cvode} was concurrent with a redesign of the vector operations
module across the {\sundials} suite. The key feature of the new
{\nvector} module is that it is written in terms of abstract vector
operations with the actual vector kernels attached by a particular
implementation (such as serial or parallel) of {\nvector}. This allows
writing the {\sundials} solvers in a manner independent of the actual
{\nvector} implementation (which can be user-supplied), as well as
allowing more than one {\nvector} module linked into an executable file.
\index{CVODE@{\cvode}!relationship to {\cvode}, {\pvode}|)}

\index{CVODE@{\cvode}!motivation for writing in C|(}
There are several motivations for choosing the {\C} language for {\cvode}.
First, a general movement away from {\F} and toward {\C} in scientific
computing is apparent.  Second, the pointer, structure, and dynamic
memory allocation features in C are extremely useful in software of
this complexity, with the great variety of method options offered.
Finally, we prefer {\C} over {\CPP} for {\cvode} because of the wider
availability of {\C} compilers, the potentially greater efficiency of {\C},
and the greater ease of interfacing the solver to applications written
in extended {\F}.
\index{CVODE@{\cvode}!motivation for writing in C|)}

\section{Changes from previous versions}

\subsection*{Changes in v2.4.0}

{\cvspbcg} and {\cvsptfqmr} modules have been added to interface with the
Scaled Preconditioned Bi-CGstab ({\spbcg}) and Scaled Preconditioned
Transpose-Free Quasi-Minimal Residual ({\sptfqmr}) linear solver modules,
respectively (for details see Chapter \ref{s:simulation}). Corresponding
additions were made to the {\F} interface module {\fcvode}.
At the same time, function type names for Scaled Preconditioned Iterative
Linear Solvers were added for the user-supplied Jacobian-times-vector and
preconditioner setup and solve functions.

\subsection*{Changes in v2.3.0}

The user interface has been further refined. Several functions used
for setting optional inputs were combined into a single one.  An optional
user-supplied routine for setting the error weight vector was added.
Additionally, to resolve potential variable scope issues, all
SUNDIALS solvers release user data right after its use. The build
systems has been further improved to make it more robust.

\subsection*{Changes in v2.2.1}

The changes in this minor {\sundials} release affect only the build system.

\subsection*{Changes in v2.2.0}

The major changes from the previous version involve a redesign of the
user interface across the entire {\sundials} suite. We have eliminated the
mechanism of providing optional inputs and extracting optional statistics 
from the solver through the \id{iopt} and \id{ropt} arrays. Instead,
{\cvode} now provides a set of routines (with prefix \id{CVodeSet})
to change the default values for various quantities controlling the
solver and a set of extraction routines (with prefix \id{CVodeGet})
to extract statistics after return from the main solver routine.
Similarly, each linear solver module provides its own set of {\id{Set}-}
and {\id{Get}-type} routines. For more details see \S\ref{ss:optional_input}
and \S\ref{ss:optional_output}.

Additionally, the interfaces to several user-supplied routines
(such as those providing Jacobians and preconditioner information) 
were simplified by reducing the number
of arguments. The same information that was previously accessible
through such arguments can now be obtained through {\id{Get}-type}
functions.

The rootfinding feature was added, whereby the roots of a set of given
functions may be computed during the integration of the ODE system.

Installation of {\cvode} (and all of {\sundials}) has been completely 
redesigned and is now based on configure scripts.


\section{Reading this User Guide}\label{ss:reading}

This user guide is a combination of general usage instructions and
specific example programs.  We expect that some readers will want to
concentrate on the general instructions, while others will refer
mostly to the examples, and the organization is intended to
accommodate both styles.

There are different possible levels of usage of {\cvode}. The most
casual user, with a small IVP problem only, can get by with reading
\S\ref{ss:ivp_sol}, then Chapter \ref{s:simulation} through
\S\ref{sss:cvode} only, and looking at examples in \cite{cvode2.4.0_ex}.
In a different direction, a more expert user with an IVP problem may want
to (a) use a package preconditioner (\S\ref{ss:preconds}), (b) supply
his/her own Jacobian or preconditioner routines (\S\ref{ss:user_fct_sim}),
(c) do multiple runs of problems of the same size (\S\ref{sss:cvreinit}), 
(d) supply a new {\nvector} module (Chapter \ref{s:nvector}), or even 
(e) supply a different linear solver module
(\S\ref{ss:cvode_org} and Chapter \ref{s:gen_linsolv}).

The structure of this document is as follows:
\begin{itemize}
\item
  In Chapter \ref{s:install} we begin with instructions for the
  installation of {\cvode}, within the structure of {\sundials}.
\item
  In Chapter \ref{s:math}, we give short descriptions of the numerical
  methods implemented by {\cvode} for the solution of initial value
  problems for systems of ODEs.
\item
  The following chapter describes the structure of the {\sundials} suite
  of solvers (\S\ref{ss:sun_org}) and the software organization of the {\cvode}
  solver (\S\ref{ss:cvode_org}). 
\item
  Chapter \ref{s:simulation} is the main usage document for {\cvode} for
  {\C} applications.  It includes a complete description of the user interface
  for the integration of ODE initial value problems.
\item
  In Chapter \ref{s:fcmix}, we describe {\fcvode}, an interface module
  for the use of {\cvode} with {\F} applications.
\item
  Chapter \ref{s:nvector} gives a brief overview of the generic
  {\nvector} module shared among the various components of
  {\sundials}, and details on the two {\nvector} implementations
  provided with {\sundials}: a serial implementation
  (\S\ref{ss:nvec_ser}) and a parallel implementation based on
  MPI\index{MPI} (\S\ref{ss:nvec_par}).
\item
  Chapter \ref{s:new_linsolv} describes the interfaces to the linear
  solver modules, so that a user can provide his/her own such module.
\item
  Chapter \ref{s:gen_linsolv} describes in detail the generic linear
  solvers shared by all {\sundials} solvers.
\item
  Finally, Chapter \ref{c:constants} lists the constants used for input to
  and output from {\cvode}.
\end{itemize}

Finally, the reader should be aware of the following notational conventions
in this user guide:  program listings and identifiers (such as \id{CVodeMalloc}) 
within textual explanations appear in typewriter type style; 
fields in {\C} structures (such as {\em content}) appear in italics;
and packages or modules, such as {\cvdense}, are written in all capitals. 
In the Index, page numbers that appear in bold indicate the main reference
for that entry.

\paragraph{Acknowledgments.}
We wish to acknowledge the contributions to previous versions of the
{\cvode} and {\pvode} codes and user guides of Scott D. Cohen \cite{CoHi:94}
and George D. Byrne \cite{ByHi:98}.

