%===================================================================================
\chapter{Description of the NVECTOR module}\label{s:nvector}
%===================================================================================
\index{NVECTOR@\texttt{NVECTOR} module}
\input{nvec_generic}

%---------------------------------------------------------------------------
\section{The NVECTOR\_SERIAL implementation}\label{ss:nvec_ser}
\input{nvec_serial}

%---------------------------------------------------------------------------
\section{The NVECTOR\_PARALLEL implementation}\label{ss:nvec_par}
\input{nvec_parallel}

%---------------------------------------------------------------------------
\section{NVECTOR functions used by KINSOL}

In Table \ref{t:nvecuse} below, we list the vector functions in the 
{\nvector} module within the {\kinsol} package.
The table also shows, for each function, which of the code modules uses
the function. The {\kinsol} column shows function usage within the main
solver module, while the remaining five columns show function
usage within each of the four {\kinsol} linear solvers ({\kinspils}
stands for any of {\kinspgmr}, {\kinspbcg}, or {\kinsptfqmr}),
the {\kinbbdpre} preconditioner module, and the {\fkinsol} module.

There is one subtlety in the {\kinspils} column hidden by the table
explained here for the case of the {\kinspgmr} module. The dot product
function \id{N\_VDotProd} is called both within the 
implementation file \id{kinsol\_spgmr.c} for the {\kinspgmr} solver and within 
the implementation files \id{sundials\_spgmr.c} and \id{sundials\_iterative.c}
for the generic {\spgmr} solver upon which the {\kinspgmr} solver is implemented.
Also, although \id{N\_VDiv} is not called within the
implementation file \id{kinsol\_spgmr.c}, it is called within the
implementation file \id{sundials\_spgmr.c} and so is required by the
{\kinspgmr} solver module. This issue does not arise for the other two
{\kinsol} linear solvers because the generic {\dense} and {\band} solvers
(used in the implementation of {\kindense} and {\kinband}) do not make calls
to any vector functions.

At this point, we should emphasize that the {\kinsol} user does not need to know 
anything about the usage of vector functions by the {\kinsol} code modules in order 
to use {\kinsol}. The information is presented as an implementation detail for the 
interested reader.

\begin{table}[htb]
\centering
\caption{List of vector functions usage by KINSOL code modules}\label{t:nvecuse}
\medskip
\begin{tabular}{|r|c|c|c|c|c|c|} \hline
                                            &
\begin{sideways}{\kinsol}    \end{sideways} &
\begin{sideways}{\kindense}  \end{sideways} &
\begin{sideways}{\kinband}   \end{sideways} &
\begin{sideways}{\kinspils}  \end{sideways} &
\begin{sideways}{\kinbbdpre} \end{sideways} &
\begin{sideways}{\fkinsol}   \end{sideways} 
\\ \hline\hline
%                        KINSOL DENSE BAND  SPILS BPRE FKINSOL
\id{N\_VClone}           & \cm &     &     & \cm & \cm &     \\ \hline
\id{N\_VCloneEmpty}      &     &     &     &     &     & \cm \\ \hline
\id{N\_VDestroy}         & \cm &     &     & \cm & \cm & \cm \\ \hline
\id{N\_VSpace}           & \cm &     &     &     &     &     \\ \hline
\id{N\_VGetArrayPointer} &     & \cm & \cm &     & \cm & \cm \\ \hline
\id{N\_VSetArrayPointer} &     & \cm &     &     &     & \cm \\ \hline
\id{N\_VLinearSum}       & \cm & \cm &     & \cm &     &     \\ \hline
\id{N\_VConst}           &     &     &     & \cm &     &     \\ \hline
\id{N\_VProd}            & \cm & \cm & \cm & \cm &     &     \\ \hline
\id{N\_VDiv}             & \cm &     &     & \cm &     &     \\ \hline
\id{N\_VScale}           & \cm & \cm & \cm & \cm & \cm &     \\ \hline
\id{N\_VAbs}             & \cm &     &     &     &     &     \\ \hline
\id{N\_VInv}             & \cm &     &     &     &     &     \\ \hline
\id{N\_VDotProd}         &     & \cm & \cm & \cm &     &     \\ \hline
\id{N\_VMaxNorm}         & \cm &     &     &     &     &     \\ \hline
\id{N\_VMin}             & \cm &     &     &     &     &     \\ \hline
\id{N\_VWL2Norm}         & \cm & \cm & \cm & \cm &     &     \\ \hline
\id{N\_VL1Norm}          &     &     &     & \cm &     &     \\ \hline
\id{N\_VConstrMask}      & \cm &     &     &     &     &     \\ \hline
\id{N\_VMinQuotient}     & \cm &     &     &     &     &     \\ \hline
\end{tabular}
\end{table}

The vector functions listed in Table \ref{t:nvecops} that are {\em not} used by
{\kinsol} are: \id{N\_VAddConst}, \id{N\_VWrmsNorm}, \id{N\_VWrmsNormMask},
\id{N\_VCompare}, and \id{N\_VInvTest}.
Therefore a user-supplied {\nvector} module for {\kinsol} could omit these five
functions.
