%===============================================================================
\section{Introduction}\label{s:ex_intro}
%===============================================================================

This report is intended to serve as a companion document to the User
Documentation of {\kinsol} \cite{kinsol2.2.0_ug}.  It provides details, with
listings, on the example programs supplied with the {\kinsol} distribution
package.

The {\kinsol} distribution contains examples of four types: serial
{\C} examples, parallel {\C} examples, and serial and parallel {\F}
examples.  The following lists summarize all of these examples.

Supplied in the \id{sundials/kinsol/examples\_ser} directory is the
following serial example (using the {\nvecs} module):

\begin{itemize}
\item \id{kinkryx}
  is an example program for {\kinsol} with the Krylov linear solver.

  This program solves a nonlinear system that arises from a system
  of partial differential equations.  The PDE system is a six-species
  food web population model, with predator-prey interaction and diffusion
  on the unit square in two dimensions.
  The preconditioner is a block-diagonal matrix based on
  the partial derivatives of the interaction terms only.
\end{itemize}
%%
Supplied in the \id{sundials/kinsol/examples\_par} directory are
the following two parallel examples (using the {\nvecp} module):
\begin{itemize}
\item \id{kinkryx\_p}
  is a parallel implementation of \id{kinkryx}.
\item \id{kinkryx\_bbd\_p}
  solves the same problem as \id{kinkryx\_p}, with a block-diagonal matrix
  with banded blocks as a preconditioner, generated by difference quotients,
  using the module {\kinbbdpre}.
\end{itemize}
%%
With the {\fkinsol} module, in the directories 
\id{sundials/kinsol/fcmix/examples\_ser} and
\id{sundials/kinsol/fcmix/examples\_par}, are the following examples for
the {\F}-{\C} interface:
\begin{itemize}
\item \id{fkinkryx}
  is a serial example, which solves a nonlinear system of the form
  $u_i^2 = i^2$ using an approximate diagonal preconditioner.
\item \id{fkinkryx\_p}
  is a parallel implementation of \id{fkinkryx}.
\end{itemize}

\vspace{0.2in}\noindent 
In the following sections, we give detailed descriptions of some (but
not all) of these examples.  The Appendices contain complete listings
of those examples described below.  We also give our output files for
each of these examples, but users should be cautioned that their
results may differ slightly from these.  Differences in solution
values may differ within the tolerances, and differences in cumulative
counters, such as numbers of Newton iterations, may differ
from one machine environment to another by as much as 10\% to 20\%.

In the descriptions below, we make frequent references to the {\kinsol}
User Document \cite{kinsol2.2.0_ug}.  All citations to specific sections
(e.g. \S\ref{s:types}) are references to parts of that User Document, unless
explicitly stated otherwise.

\vspace{0.2in}\noindent {\bf Note}. 
The examples in the {\kinsol} distribution are written in such a way as
to compile and run for any combination of configuration options during
the installation of {\sundials} (see \S\ref{s:install}). As a consequence,
they contain portions of code that will not be typically present in a
user program. For example, all {\C} example programs make use of the
variable \id{SUNDIALS\_EXTENDED\_PRECISION} to test if the solver libraries
were built in extended precision and use the appropriate conversion 
specifiers in \id{printf} functions. Similarly, the {\F} examples in
{\fkinsol} are automatically pre-processed to generate source code that
corresponds to the precision in which the {\kinsol} libraries were built
(see \S\ref{s:ex_fortran} in this document for more details).

