%===================================================================================
\chapter{Introduction}\label{s:intro}
%===================================================================================

{\kinsol} is part of a software family called {\sundials}: SUite
of Nonlinear and DIfferential/ALgebraic equation Solvers~\cite{HBGLSSW:05}.
This suite consists of {\cvode}, {\arkode}, {\kinsol}, and {\ida}, and variants
of these with sensitivity analysis capabilities.

%
{\kinsol}\index{KINSOL@{\kinsol}!brief description of} is a
general-purpose nonlinear system solver based on Newton-Krylov
solver technology.  A fixed point iteration is also included with the release
of {\kinsol} v.2.8.0 and higher.

%---------------------------------
\section{Historical Background}\label{ss:history}
%---------------------------------

\index{KINSOL@{\kinsol}!relationship to NKSOL}  The first
nonlinear solver packages based on Newton-Krylov methods were
written in {\F}.  In particular, the NKSOL package, written at
LLNL, was the first Newton-Krylov solver package written for
solution of systems arising in the solution of partial differential
equations~\cite{BrSa:90}.  This {\F} code made use of Newton's
method to solve the discrete nonlinear systems and applied a
preconditioned Krylov linear solver for solution of the Jacobian
system at each nonlinear iteration.  The key to the Newton-Krylov
method was that the matrix-vector multiplies required by the
Krylov method could effectively be approximated by a finite
difference of the nonlinear system-defining function, avoiding a
requirement for the formation of the actual Jacobian matrix.
Significantly less memory was required for the solver as a result.

In the late 1990's, there was a push at LLNL to rewrite the
nonlinear solver in {\CC} and port it to distributed memory
parallel machines.  Both Newton and Krylov methods are easily
implemented in parallel, and this effort gave rise to the
{\kinsol} package. {\kinsol} is similar to NKSOL in functionality,
except that it provides for more options in the choice of linear
system methods and tolerances, and has a more modular design to provide
flexibility for future enhancements.
\index{KINSOL@{\kinsol}!relationship to NKSOL}

At present, {\kinsol} may utilize a variety of Krylov methods provided
in {\sundials}. These methods include the
GMRES (Generalized Minimal RESidual)~\cite{SaSc:86},
FGMRES (Flexible Generalized Minimum RESidual)~\cite{Saa:93},
Bi-CGStab (Bi-Conjugate Gradient Stabilized)~\cite{Van:92}, TFQMR
(Transpose-Free Quasi-Minimal Residual)~\cite{Fre:93}, and PCG
(Preconditioned Conjugate Gradient)~\cite{HeSt:52} linear iterative
methods.  As Krylov methods, these require little
matrix storage for solving the Newton equations as compared to direct
methods. However, the algorithms allow for a user-supplied preconditioner
matrix, and, for most problems, preconditioning is essential for an
efficient solution.  For very large nonlinear algebraic systems,
the Krylov methods are preferable over direct linear solver methods,
and are often the only feasible choice. Among the Krylov
methods in {\sundials}, we recommend GMRES as the best overall
choice. However, users are encouraged to compare all three,
especially if encountering convergence failures with GMRES.
Bi-CGStab and TFQMR have an advantage in storage requirements,
in that the number of workspace vectors they require is fixed,
while that number for GMRES depends on the desired Krylov
subspace size.   FGMRES
has an advantage in that it is designed to support preconditioners
that vary between iterations (e.g.~iterative methods).  PCG exhibits
rapid convergence and minimal workspace vectors, but only works for
symmetric linear systems.

For the sake of completeness in functionality, direct linear
system solvers are included in {\kinsol}.  These include methods
for both dense and banded linear systems, with Jacobians that
are either user-supplied or generated internally by difference
quotients.  {\kinsol} also includes interfaces to the sparse
direct solvers KLU~\cite{DaPa:10,KLU_site},
and the threaded sparse direct solver,
SuperLU\_MT~\cite{Li:05,DGL:99,SuperLUMT_site}.

In the process of translating NKSOL into {\CC}, the overall
{\kinsol} organization has been changed considerably. One key
feature of the {\kinsol} organization is that a separate module
devoted to vector operations was created.  This module
facilitated extension to multiprosessor environments with minimal
impact on the rest of the solver. The vector module design is
shared across the {\sundials} suite. This {\nvector} module is
written in terms of abstract vector operations with the actual
routines attached by a particular implementation (such as serial
or parallel) of {\nvector}. This abstraction allows writing the {\sundials}
solvers in a manner independent of the actual {\nvector}
implementation (which can be user-supplied), as well as allowing
more than one {\nvector} module linked into an executable file.
{\sundials} (and thus {\kinsol}) is supplied with serial,
MPI-parallel, and both OpenMP and Pthreads thread-parallel
{\nvector} implementations.

\index{KINSOL@{\kinsol}!motivation for writing in C|(} There are
several motivations for choosing the {\CC} language for {\kinsol}.
First, a general movement away from {\F} and toward {\CC} in
scientific computing was apparent. Second, the pointer, structure,
and dynamic memory allocation features in C are extremely useful
in software of this complexity, with the great variety of method
options offered. Finally, we prefer {\CC} over {\CPP} for {\kinsol}
because of the wider availability of {\CC} compilers, the
potentially greater efficiency of {\CC}, and the greater ease of
interfacing the solver to applications written in {\F}.
\index{KINSOL@{\kinsol}!motivation for writing in C|)}


\section{Changes from previous versions}

\subsection*{Changes in v5.4.0}

A new API, \id{SUNMemoryHelper}, was added to support \textbf{GPU users} who
have complex memory management needs such as using memory pools. This is paired
with new constructors for the {\nveccuda} and {\nvecraja} modules that accept a
\id{SUNMemoryHelper} object. Refer to sections
\ref{s:gpu_model},\ref{s:sunmemory}, \ref{ss:nvec_cuda} and \ref{ss:nvec_raja}
for more information.

The \id{NVECTOR\_RAJA} module has been updated to mirror the \id{NVECTOR\_CUDA} module.
Notably, the update adds managed memory support to the \id{NVECTOR\_RAJA} module.
Users of the module will need to update any calls to the \id{N\_VMake\_Raja} function
because that signature was changed. This module remains experimental and is
subject to change from version to version.

The \id{NVECTOR\_TRILINOS} module has been updated to work with Trilinos 12.18+.
This update changes the local ordinal type to always be an \id{int}.

Added support for CUDA v11.


\subsection*{Changes in v5.3.0}

Fixed a bug in the iterative linear solver modules where an error is not
returned if the Atimes function is \id{NULL} or, if preconditioning is enabled,
the PSolve function is \id{NULL}.

Added the ability to control the {\cuda} kernel launch parameters for the
\id{NVECTOR\_CUDA} and \id{SUNMATRIX\_CUSPARSE} modules. These modules remain
experimental and are subject to change from version to version.
In addition, the \id{NVECTOR\_CUDA} kernels were rewritten to be more flexible.
Most users should see equivalent performance or some improvement, but a select
few may observe minor performance degradation with the default settings. Users
are encouraged to contact the {\sundials} team about any perfomance changes
that they notice.

Added new capabilities for monitoring the solve phase in the {\sunnonlinsolnewton}
and {\sunnonlinsolfixedpoint} modules, and the {\sundials} iterative linear solver
modules. {\sundials} must be built with the CMake option
\id{SUNDIALS\_BUILD\_WITH\_MONITORING} to use these capabilties.

Added the optional function \id{KINSetJacTimesVecSysFn} to specify an
alternative system function for computing Jacobian-vector products with the
internal difference quotient approximation.

\subsection*{Changes in v5.2.0}

Fixed a build system bug related to the Fortran 2003 interfaces when using the
IBM XL compiler. When building the Fortran 2003 interfaces with an XL compiler
it is recommended to set \id{CMAKE\_Fortran\_COMPILER} to \id{f2003},
\id{xlf2003}, or \id{xlf2003\_r}.

Fixed a linkage bug affecting Windows users that stemmed from dllimport/dllexport
attributes missing on some SUNDIALS API functions.

Added a new \id{SUNMatrix} implementation, \id{SUNMATRIX\_CUSPARSE}, that interfaces
to the sparse matrix implementation from the NVIDIA cuSPARSE library. In addition,
the \id{SUNLINSOL\_CUSOLVER\_BATCHQR} linear solver has been updated to
use this matrix, therefore, users of this module will need to update their code.
These modules are still considered to be experimental, thus they are subject to
breaking changes even in minor releases.

\subsection*{Changes in v5.1.0}

Fixed a build system bug related to finding LAPACK/BLAS.

Fixed a build system bug related to checking if the KLU library works.

Fixed a build system bug related to finding PETSc when using the CMake
variables \id{PETSC\_INCLUDES} and \id{PETSC\_LIBRARIES} instead of
\id{PETSC\_DIR}.

Added a new build system option, \id{CUDA\_ARCH}, that can be used to specify
the CUDA architecture to compile for.

Added two utility functions, \id{SUNDIALSFileOpen} and \id{SUNDIALSFileClose}
for creating/destroying file pointers that are useful when using the Fortran
2003 interfaces.

Added support for constant damping when using Anderson acceleration. See Chatper
\ref{s:math} and the description of the \id{KINSetDampingAA} function for more
details.

\subsection*{Changes in v5.0.0}

\subsubsection*{Build system changes}

\begin{itemize}
\item Increased the minimum required CMake version to 3.5 for most {\sundials}
configurations, and 3.10 when CUDA or OpenMP with device offloading are enabled.
%
\item The CMake option \id{BLAS\_ENABLE} and the variable \id{BLAS\_LIBRARIES} have
been removed to simplify builds as {\sundials} packages do not use BLAS
directly. For third party libraries that require linking to BLAS, the path to
the BLAS library should be included in the \id{\_LIBRARIES} variable for the
third party library \textit{e.g.}, \id{SUPERLUDIST\_LIBRARIES} when enabling
SuperLU\_DIST.
%
\item Fixed a bug in the build system that prevented the {\nvecpthreads} module from
being built.
\end{itemize}

\subsubsection*{NVECTOR module changes}

\begin{itemize}
\item Two new functions were added to aid in creating custom {\nvector} objects. The
constructor \id{N\_VNewEmpty} allocates an ``empty'' generic {\nvector} with the
object's content pointer and the function pointers in the operations structure
initialized to \id{NULL}. When used in the constructor for custom objects this
function will ease the introduction of any new optional operations to the
{\nvector} API by ensuring only required operations need to be set.
Additionally, the function \id{N\_VCopyOps(w, v)} has been added to copy the
operation function pointers between vector objects. When used in clone routines
for custom vector objects these functions also will ease the introduction of
any new optional operations to the {\nvector} API by ensuring all operations
are copied when cloning objects. See \S\ref{ss:nvecutils} for more details.
%
\item Two new {\nvector} implementations, {\nvecmanyvector} and
{\nvecmpimanyvector}, have been created to support flexible partitioning
of solution data among different processing elements (e.g., CPU + GPU) or for
multi-physics problems that couple distinct MPI-based simulations together. This
implementation is accompanied by additions to user documentation and {\sundials}
examples. See \S\ref{ss:nvec_manyvector} and \S\ref{ss:nvec_mpimanyvector} for
more details.
%
\item One new required vector operation and ten new optional vector operations have
been added to the {\nvector} API. The new required operation, \id{N\_VGetLength},
returns the global length of an \id{N\_Vector}. The optional operations have
been added to support the new \newline\noindent
{\nvecmpimanyvector} implementation. The
operation \id{N\_VGetCommunicator} must be implemented by subvectors that are
combined to create an {\nvecmpimanyvector}, but is not used outside of
this context. The remaining nine operations are optional local reduction
operations intended to eliminate unnecessary latency when performing vector
reduction operations (norms, etc.) on distributed memory systems. The optional
local reduction vector operations are
\id{N\_VDotProdLocal},
\id{N\_VMaxNormLocal},
\id{N\_VMinLocal},
\id{N\_VL1NormLocal},
\id{N\_VWSqrSumLocal},
\id{N\_VWSqrSumMaskLocal},
\id{N\_VInvTestLocal},
\id{N\_VConstrMaskLocal}, and
\id{N\_VMinQuotientLocal}.
If an {\nvector} implementation defines any of the local operations as
\id{NULL}, then the {\nvecmpimanyvector} will call standard {\nvector}
operations to complete the computation. See \S\ref{ss:nveclocalops} for more
details.
%
\item An additional {\nvector} implementation, {\nvecmpiplusx}, has been created to
support the MPI+X paradigm where X is a type of on-node parallelism
(\textit{e.g.}, OpenMP, CUDA). The implementation is accompanied by additions to
user documentation and {\sundials} examples. See \S\ref{ss:nvec_mpiplusx} for
more details.
%
\item The \id{*\_MPICuda} and \id{*\_MPIRaja} functions have been removed from the
{\nveccuda} and {\nvecraja} implementations respectively. Accordingly, the
\id{nvector\_mpicuda.h}, \newline\noindent
\id{nvector\_mpiraja.h},
\id{libsundials\_nvecmpicuda.lib}, and \id{libsundials\_nvecmpicudaraja.lib}
files have been removed. Users should use the {\nvecmpiplusx} module coupled
in conjunction with the {\nveccuda} or {\nvecraja} modules to replace the
functionality. The necessary changes are minimal and should require few
code modifications. See the programs in \id{examples/ida/mpicuda} and
\id{examples/ida/mpiraja} for examples of how to use the {\nvecmpiplusx}
module with the {\nveccuda} and {\nvecraja} modules respectively.
%
\item Fixed a memory leak in the {\nvecpetsc} module clone function.
%
\item Made performance improvements to the {\nveccuda} module. Users who utilize a
non-default stream should no longer see default stream synchronizations
after memory transfers.
%
\item Added a new constructor to the {\nveccuda} module that allows a user to provide
custom allocate and free functions for the vector data array and internal
reduction buffer. See \S\ref{ss:nvec_cuda_functions} for more details.
%
\item Added new Fortran 2003 interfaces for most {\nvector} modules. See Chapter
\ref{s:nvector} for more details on how to use the interfaces.
%
\item Added three new {\nvector} utility functions,
\id{FN\_VGetVecAtIndexVectorArray},\newline\noindent
\id{FN\_VSetVecAtIndexVectorArray}, and
\id{FN\_VNewVectorArray},
for working with \id{N\_Vector} arrays when using the Fortran 2003 interfaces.
See \S\ref{ss:nvecutils} for more details.
\end{itemize}

\subsubsection*{SUNMatrix module changes}

\begin{itemize}
\item Two new functions were added to aid in creating custom {\sunmatrix} objects. The
constructor \id{SUNMatNewEmpty} allocates an ``empty'' generic {\sunmatrix} with
the object's content pointer and the function pointers in the operations
structure initialized to \id{NULL}. When used in the constructor for custom
objects this function will ease the introduction of any new optional operations
to the {\sunmatrix} API by ensuring only required operations need to be set.
Additionally, the function \id{SUNMatCopyOps(A, B)} has been added to copy the
operation function pointers between matrix objects. When used in clone routines
for custom matrix objects these functions also will ease the introduction of any
new optional operations to the {\sunmatrix} API by ensuring all operations are
copied when cloning objects. See \S\ref{ss:sunmatrix_utilities} for more details.
%
\item A new operation, \id{SUNMatMatvecSetup}, was added to the {\sunmatrix} API
to perform any setup necessary for computing a matrix-vector product. This
operation is useful for {\sunmatrix} implementations which need to prepare the
matrix itself, or communication structures before performing the matrix-vector
product. Users who have implemented custom {\sunmatrix} modules will need to at
least update their code to set the corresponding \id{ops} structure member,
\id{matvecsetup}, to \id{NULL}. See \S\ref{ss:sunmatrix_functions} for more
details.
%
\item The generic {\sunmatrix} API now defines error codes to be returned by
{\sunmatrix} operations. Operations which return an integer flag indiciating
success/failure may return different values than previously. See
\S\ref{ss:sunmatrix_ReturnCodes} for more details.
%
\item A new {\sunmatrix} (and {\sunlinsol}) implementation was added to
facilitate the use of the SuperLU\_DIST library with {\sundials}. See
\S\ref{ss:sunmat_slunrloc} for more details.
%
\item Added new Fortran 2003 interfaces for most {\sunmatrix} modules. See Chapter
\ref{s:sunmatrix} for more details on how to use the interfaces.
\end{itemize}

\subsubsection*{SUNLinearSolver module changes}

\begin{itemize}
\item A new function was added to aid in creating custom {\sunlinsol} objects.
The constructor \id{SUNLinSolNewEmpty} allocates an ``empty'' generic
{\sunlinsol} with the object's content pointer and the function pointers
in the operations structure initialized to \id{NULL}. When used in the
constructor for custom objects this function will ease the introduction of any
new optional operations to the {\sunlinsol} API by ensuring only required
operations need to be set. See \S\ref{ss:sunlinsol_custom} for more details.
%
\item The return type of the {\sunlinsol} API function \id{SUNLinSolLastFlag}
has changed from \id{long int} to \id{sunindextype} to be consistent with the
type used to store row indices in dense and banded linear solver modules.
%
\item Added a new optional operation to the {\sunlinsol} API,
\id{SUNLinSolGetID}, that returns a \newline\noindent
\id{SUNLinearSolver\_ID} for identifying the
linear solver module.
%
\item The {\sunlinsol} API has been updated to make the initialize and setup
functions optional.
%
\item A new {\sunlinsol} (and {\sunmatrix}) implementation was added to
facilitate the use of the SuperLU\_DIST library with {\sundials}. See
\S\ref{ss:sunlinsol_sludist} for more details.
%
\item Added a new {\sunlinsol} implementation,
\id{SUNLinearSolver\_cuSolverSp\_batchQR}, which leverages the NVIDIA cuSOLVER
sparse batched QR method for efficiently solving block diagonal linear systems
on NVIDIA GPUs. See \S\ref{ss:sunlinsol_cuspbqr} for more details.
%
\item Added three new accessor functions to the {\sunlinsolklu} module,
\id{SUNLinSol\_KLUGetSymbolic},
\id{SUNLinSol\_KLUGetNumeric}, and
\id{SUNLinSol\_KLUGetCommon},
to provide user access to the underlying KLU solver structures. See
\S\ref{ss:sunlinsol_klu_functions} for more details.
%
\item Added new Fortran 2003 interfaces for most {\sunlinsol} modules. See
Chapter \ref{s:sunlinsol} for more details on how to use the interfaces.
\end{itemize}

\subsubsection*{KINSOL changes}

\begin{itemize}
\item Fixed a bug in the {\kinsol} linear solver interface where the auxiliary scalar
\id{sJpnorm} was not computed when necessary with the Picard iteration and the
auxiliary scalar \id{sFdotJp} was unnecessarily computed in some cases.
%
\item The KINLS interface has been updated to only zero the Jacobian matrix before
calling a user-supplied Jacobian evaluation function when the attached linear
solver has type \newline\noindent
\id{SUNLINEARSOLVER\_DIRECT}.
%
\item Added a Fortran 2003 interface to {\kinsol}. See Chapter~\ref{s:kinfort}
for more details.
\end{itemize}


\subsection*{Changes in v5.0.0-dev.0}

An additional {\nvector} implementation, {\nvecmanyvector}, was
created to support flexible partitioning of solution data among
different processing elements (e.g., CPU + GPU) or for multi-physics
problems that couple distinct MPI-based simulations together (see
Section \ref{ss:nvec_manyvector} for more details).  This
implementation is accompanied by additions to user documentation and
{\sundials} examples.

Eleven new optional vector operations have been added to the {\nvector} API to
support the new {\nvecmanyvector} implementation (see Chapter \ref{s:nvector}
for more details). Two of the operations, \id{N\_VGetCommunicator} and
\id{N\_VGetLength}, must be implemented by subvectors that are combined to
create an {\nvecmanyvector}, but are not used outside of this context. The
remaining nine operations are optional local reduction operations intended to
eliminate unnecessary latency when performing vector reduction operations
(norms, etc.) on distributed memory systems. The optional local reduction vector
operations are
\id{N\_VDotProdLocal},
\id{N\_VMaxNormLocal},
\id{N\_VMinLocal},
\id{N\_VL1NormLocal},
\id{N\_VWSqrSumLocal},
\id{N\_VWSqrSumMaskLocal},
\id{N\_VInvTestLocal},
\id{N\_VConstrMaskLocal}, and\\
\id{N\_VMinQuotientLocal}.
If an {\nvector} implementation defines any of the local operations as
\id{NULL}, then the {\nvecmanyvector} will call standard {\nvector} operations
to complete the computation.

A new {\sunmatrix} and {\sunlinsol} implementation was added to facilitate
the use of the {\superludist} library with SUNDIALS.

A new operation, \id{SUNMatMatvecSetup}, was added to the {\sunmatrix} API. Users
who have implemented custom {\sunmatrix} modules will need to at least update
their code to set the corresponding ops structure member, \id{matvecsetup}, to \id{NULL}.

The generic {\sunmatrix} API now defines error codes to be returned by {\sunmatrix} operations.
Operations which return an integer flag indiciating success/failure may return different
values than previously.


\subsection*{Changes in v4.1.0}

An additional {\nvector} implementation was added for the
{\tpetra} vector from the {\trilinos} library to facilitate interoperability
between {\sundials} and {\trilinos}. This implementation is accompanied by
additions to user documentation and {\sundials} examples.

The \id{EXAMPLES\_ENABLE\_RAJA} CMake option has been removed. The option \id{EXAMPLES\_ENABLE\_CUDA}
enables all examples that use CUDA including the RAJA examples with a CUDA back end (if the RAJA
{\nvector} is enabled).

The implementation header file \id{kin\_impl.h} is no longer installed. This means users
who are directly manipulating the \id{KINMem} structure will need to update their code
to use {\kinsol}'s public API.

Python is no longer required to run \id{make test} and \id{make test\_install}.

\subsection*{Changes in v4.0.2}

Added information on how to contribute to {\sundials} and a contributing agreement.

Moved definitions of DLS and SPILS backwards compatibility functions to a source file.
The symbols are now included in the {\kinsol} library, \id{libsundials\_kinsol}.

\subsection*{Changes in v4.0.1}

No changes were made in this release.

\subsection*{Changes in v4.0.0}

{\kinsol}'s previous direct and iterative linear solver interfaces,
{\kindls} and {\kinspils}, have been merged into a single unified linear
solver interface, {\kinls}, to support any valid {\sunlinsol} module.
This includes the ``DIRECT'' and ``ITERATIVE'' types as well as the new
``MATRIX\_ITERATIVE'' type. Details regarding how {\kinls} utilizes linear
solvers of each type as well as discussion regarding intended use cases for
user-supplied {\sunlinsol} implementations are included in
Chapter~\ref{s:sunlinsol}. All {\kinsol} example programs and the standalone
linear solver examples have been updated to use the unified linear solver
interface.

The unified interface for the new {\kinls} module is very similar to the
previous {\kindls} and {\kinspils} interfaces. To minimize challenges in user
migration to the new names, the previous {\CC} and {\F} routine names may still
be used; these will be deprecated in future releases, so we recommend that users
migrate to the new names soon. Additionally, we note that {\F} users, however,
may need to enlarge their \id{iout} array of optional integer outputs, and
update the indices that they query for certain linear-solver-related
statistics.

The names of all constructor routines for {\sundials}-provided
{\sunlinsol} implementations have been updated to follow the naming convention
\id{SUNLinSol\_*} where \id{*} is the name of the linear solver. The new names
are
\id{SUNLinSol\_Band},
\id{SUNLinSol\_Dense},
\id{SUNLinSol\_KLU},
\id{SUNLinSol\_LapackBand},\newline
\id{SUNLinSol\_LapackDense},
\id{SUNLinSol\_PCG},
\id{SUNLinSol\_SPBCGS},
\id{SUNLinSol\_SPFGMR},
\id{SUNLinSol\_SPGMR},
\id{SUNLinSol\_SPTFQMR}, and
\id{SUNLinSol\_SuperLUMT}.  Solver-specific ``set'' routine names have
been similarly standardized.  To minimize challenges in user migration
to the new names, the previous routine names may still be used; these
will be deprecated in future releases, so we recommend that users
migrate to the new names soon. All {\kinsol} example programs and the standalone
linear solver examples have been updated to use the new naming convention.

The \id{SUNBandMatrix} constructor has been simplified to remove the
storage upper bandwidth argument.

Three fused vector operations and seven vector array operations have been added
to the {\nvector} API. These \textit{optional} operations are disabled by
default and may be activated by calling vector specific routines after creating
an {\nvector} (see Chapter \ref{s:nvector} for more details). The new operations
are intended to increase data reuse in vector operations, reduce parallel
communication on distributed memory systems, and lower the number of kernel
launches on systems with accelerators. The fused operations are
\id{N\_VLinearCombination},
\id{N\_VScaleAddMulti}, and
\id{N\_VDotProdMulti}
and the vector array operations are
\id{N\_VLinearCombinationVectorArray},
\id{N\_VScaleVectorArray},
\id{N\_VConstVectorArray},
\id{N\_VWrmsNormVectorArray},
\id{N\_VWrmsNormMaskVectorArray},
\id{N\_VScaleAddMultiVectorArray}, and\\
\id{N\_VLinearCombinationVectorArray}.
If an {\nvector} implementation defines any of these operations as \id{NULL},
then standard {\nvector} operations will automatically be called as necessary to
complete the computation.
\\
\\
\noindent Multiple updates to {\nveccuda} were made:
\begin{itemize}
  \item Changed \id{N\_VGetLength\_Cuda} to return the global vector length
        instead of the local vector length.
  \item Added \id{N\_VGetLocalLength\_Cuda} to return the local vector length.
  \item Added \id{N\_VGetMPIComm\_Cuda} to return the MPI communicator used.
  \item Removed the accessor functions in the namespace suncudavec.
  \item Changed the \id{N\_VMake\_Cuda} function to take a host data pointer and a device
        data pointer instead of an \id{N\_VectorContent\_Cuda} object.
  \item Added the ability to set the \id{cudaStream\_t} used for execution of the
        {\nveccuda} kernels. See the function \id{N\_VSetCudaStreams\_Cuda}.
  \item Added \id{N\_VNewManaged\_Cuda}, \id{N\_VMakeManaged\_Cuda}, and
        \id{N\_VIsManagedMemory\_Cuda} functions to accommodate using managed
        memory with the {\nveccuda}.
\end{itemize}
%
Multiple changes to {\nvecraja} were made:
\begin{itemize}
  \item Changed \id{N\_VGetLength\_Raja} to return the global vector length
        instead of the local vector length.
  \item Added \id{N\_VGetLocalLength\_Raja} to return the local vector length.
  \item Added \id{N\_VGetMPIComm\_Raja} to return the MPI communicator used.
  \item Removed the accessor functions in the namespace suncudavec.
\end{itemize}
A new {\nvector} implementation for leveraging OpenMP 4.5+ device offloading has
been added, {\nvecopenmpdev}. See \S\ref{ss:nvec_openmpdev} for more details.


\subsection*{Changes in v3.2.1}

The changes in this minor release include the following:
\begin{itemize}
\item Fixed a bug in the {\cuda} {\nvector} where the \id{N\_VInvTest} operation
  could write beyond the allocated vector data.
\item Fixed library installation path for multiarch systems. This fix changes the default
  library installation path to \id{CMAKE\_INSTALL\_PREFIX/CMAKE\_INSTALL\_LIBDIR}
  from \id{CMAKE\_INSTALL\_PREFIX/lib}.\\
  \id{CMAKE\_INSTALL\_LIBDIR} is automatically
  set, but is available as a CMake option that can modified.
\end{itemize}

\subsection*{Changes in v3.2.0}

Fixed a problem with setting \id{sunindextype} which would occur
with some compilers (e.g. armclang) that did not define \id{\_\_STDC\_VERSION\_\_}.
\\
\\
\noindent Added hybrid MPI/CUDA and MPI/RAJA vectors to allow use of more
than one MPI rank when using a GPU system.  The vectors assume one GPU
device per MPI rank.
\\
\\
\noindent Changed the name of the {\raja} {\nvector} library to
\id{libsundials\_nveccudaraja.lib} from \newline
\id{libsundials\_nvecraja.lib} to better reflect that we only support {\cuda}
as a backend for {\raja} currently.
\\
\\
\noindent Several changes were made to the build system:
\begin{itemize}
\item CMake 3.1.3 is now the minimum required CMake version.
\item Deprecate the behavior of the \id{SUNDIALS\_INDEX\_TYPE} CMake option and
  added the \newline
  \id{SUNDIALS\_INDEX\_SIZE} CMake option to select the \id{sunindextype}
  integer size.
\item The native CMake FindMPI module is now used to locate an MPI installation.
\item If MPI is enabled and MPI compiler wrappers are not set, the build system
  will check if \id{CMAKE\_<language>\_COMPILER} can compile MPI programs before
  trying to locate and use an MPI installation.
\item The previous options for setting MPI compiler wrappers and the executable
  for running MPI programs have been have been depreated. The new options that
  align with those used in native CMake FindMPI module are
  \id{MPI\_C\_COMPILER}, \id{MPI\_CXX\_COMPILER}, \id{MPI\_Fortran\_COMPILER},
  and \id{MPIEXEC\_EXECUTABLE}.
\item When a Fortran name-mangling scheme is needed (e.g., \id{LAPACK\_ENABLE}
  is \id{ON}) the build system will infer the scheme from the Fortran
  compiler. If a Fortran compiler is not available or the inferred or default
  scheme needs to be overridden, the advanced options
  \id{SUNDIALS\_F77\_FUNC\_CASE} and \id{SUNDIALS\_F77\_FUNC\_UNDERSCORES} can
  be used to manually set the name-mangling scheme and bypass trying to infer
  the scheme.
\item Parts of the main CMakeLists.txt file were moved to new files in the
  \id{src} and \id{example} directories to make the CMake configuration file
  structure more modular.
\end{itemize}

\subsection*{Changes in v3.1.2}

The changes in this minor release include the following:
\begin{itemize}
\item Updated the minimum required version of CMake to 2.8.12 and enabled
  using rpath by default to locate shared libraries on OSX.
\item Fixed Windows specific problem where \id{sunindextype} was not correctly
  defined when using 64-bit integers for the {\sundials} index type. On Windows
  \id{sunindextype} is now defined as the MSVC basic type \id{\_\_int64}.
\item Added sparse SUNMatrix ``Reallocate'' routine to allow specification of
  the nonzero storage.
\item Updated the KLU SUNLinearSolver module to set constants for the two
  reinitialization types, and fixed a bug in the full reinitialization
  approach where the sparse SUNMatrix pointer would go out of scope on
  some architectures.
\item Updated the ``ScaleAdd'' and ``ScaleAddI'' implementations in the
  sparse SUNMatrix module to more optimally handle the case where the
  target matrix contained sufficient storage for the sum, but had the
  wrong sparsity pattern.  The sum now occurs in-place, by performing
  the sum backwards in the existing storage.  However, it is still more
  efficient if the user-supplied Jacobian routine allocates storage for
  the sum $I+\gamma J$ manually (with zero entries if needed).
\item Changed the LICENSE install path to \id{instdir/include/sundials}.
\end{itemize}

\subsection*{Changes in v3.1.1}

The changes in this minor release include the following:
\begin{itemize}
\item Fixed a potential memory leak in the {\spgmr} and {\spfgmr} linear
  solvers: if ``Initialize'' was called multiple times then the solver
  memory was reallocated (without being freed).

\item Updated KLU SUNLinearSolver module to use a \id{typedef} for the
  precision-specific solve function to be used (to avoid compiler
  warnings).

\item Added missing typecasts for some \id{(void*)} pointers (again, to
  avoid compiler warnings).

\item Bugfix in \id{sunmatrix\_sparse.c} where we had used \id{int}
  instead of \id{sunindextype} in one location.

\item Fixed a minor bug in \id{KINPrintInfo} where a case was missing for
  \id{KIN\_REPTD\_SYSFUNC\_ERR} leading to an undefined info message.

\item Added missing \id{\#include <stdio.h>} in {\nvector} and {\sunmatrix}
  header files.

\item Fixed an indexing bug in the {\cuda} {\nvector} implementation of
  \id{N\_VWrmsNormMask} and revised the {\raja} {\nvector} implementation of
  \id{N\_VWrmsNormMask} to work with mask arrays using values other than zero or
  one. Replaced \id{double} with \id{realtype} in the {\raja} vector test functions.

\item Fixed compilation issue with GCC 7.3.0 and Fortran programs that do
  not require a {\sunmatrix} or {\sunlinsol} module (e.g., iterative
  linear solvers or fixed pointer solver).
\end{itemize}
In addition to the changes above, minor corrections were also made to the
example programs, build system, and user documentation.

\subsection*{Changes in v3.1.0}

Added {\nvector} print functions that write vector data to a specified
file (e.g., \id{N\_VPrintFile\_Serial}).

Added \id{make test} and \id{make test\_install} options to the build
system for testing {\sundials} after building with \id{make} and
installing with \id{make install} respectively.

\subsection*{Changes in v3.0.0}

All interfaces to matrix structures and linear solvers
have been reworked, and all example programs have been updated.
The goal of the redesign of these interfaces was to provide more encapsulation
and ease in the interfacing of custom linear solvers and interoperability
with linear solver libraries.
Specific changes include:
\begin{itemize}
\item Added generic SUNMATRIX module with three provided implementations:
        dense, banded and sparse.  These replicate previous SUNDIALS Dls and
        Sls matrix structures in a single object-oriented API.
\item Added example problems demonstrating use of generic SUNMATRIX modules.
\item Added generic \id{SUNLinearSolver} module with eleven provided
        implementations: {\sundials} native dense, {\sundials} native banded,
        LAPACK dense, LAPACK band, KLU,
        SuperLU\_MT, SPGMR, SPBCGS, SPTFQMR, SPFGMR, and PCG.  These replicate
        previous SUNDIALS generic linear solvers in a single object-oriented
        API.
\item Added example problems demonstrating use of generic SUNLINEARSOLVER
        modules.
\item Expanded package-provided direct linear solver (Dls) interfaces and
        scaled, preconditioned, iterative linear solver (Spils) interfaces
        to utilize generic SUNMATRIX and SUNLINEARSOLVER objects.
\item Removed package-specific, linear solver-specific, solver modules
        (e.g. CVDENSE, KINBAND, IDAKLU, ARKSPGMR) since their functionality
        is entirely replicated by the generic Dls/Spils interfaces and
        SUNLINEARSOLVER/SUNMATRIX modules.  The exception is CVDIAG, a
        diagonal approximate Jacobian solver available to CVODE and CVODES.
\item Converted all SUNDIALS example problems to utilize new generic
        SUNMATRIX and SUNLINEARSOLVER objects, along with updated Dls and
        Spils linear solver interfaces.
\item Added Spils interface routines to ARKode, CVODE, CVODES, IDA and
        IDAS to allow specification of a user-provided "JTSetup" routine.
        This change supports users who wish to set up data structures for
        the user-provided Jacobian-times-vector ("JTimes") routine, and
        where the cost of one JTSetup setup per Newton iteration can be
        amortized between multiple JTimes calls.
\end{itemize}

Two additional {\nvector} implementations were added -- one for
{\cuda} and one for {\raja} vectors.
These vectors are supplied to provide very basic support for running
on GPU architectures.  Users are advised that these vectors both move all data
to the GPU device upon construction, and speedup will only be realized if the
user also conducts the right-hand-side function evaluation on the device.
In addition, these vectors assume the problem fits on one GPU.
Further information about {\raja}, users are referred to th web site,
https://software.llnl.gov/RAJA/.
These additions are accompanied by additions to various interface functions
and to user documentation.

All indices for data structures were updated to a new \id{sunindextype} that
can be configured to be a 32- or 64-bit integer data index type.
\id{sunindextype} is defined to be \id{int32\_t} or \id{int64\_t} when portable types are
supported, otherwise it is defined as \id{int} or \id{long int}.
The Fortran interfaces continue to use \id{long int} for indices, except for
their sparse matrix interface that now uses the new \id{sunindextype}.
This new flexible capability for index types includes interfaces to
PETSc, hypre, SuperLU\_MT, and KLU with
either 32-bit or 64-bit capabilities depending how the user configures
{\sundials}.

To avoid potential namespace conflicts, the macros defining \id{booleantype}
values \id{TRUE} and \id{FALSE} have been changed to \id{SUNTRUE} and
\id{SUNFALSE} respectively.

Temporary vectors were removed from preconditioner setup and solve
routines for all packages.  It is assumed that all necessary data
for user-provided preconditioner operations will be allocated and
stored in user-provided data structures.

The file \id{include/sundials\_fconfig.h} was added. This file contains
{\sundials} type information for use in Fortran programs.

The build system was expanded to support many of the xSDK-compliant keys.
The xSDK is a movement in scientific software to provide a foundation for the
rapid and efficient production of high-quality,
sustainable extreme-scale scientific applications.  More information can
be found at, https://xsdk.info.

Added functions \id{SUNDIALSGetVersion} and \id{SUNDIALSGetVersionNumber} to
get {\sundials} release version information at runtime.

In addition, numerous changes were made to the build system.
These include the addition of separate \id{BLAS\_ENABLE} and \id{BLAS\_LIBRARIES}
CMake variables, additional error checking during CMake configuration,
minor bug fixes, and renaming CMake options to enable/disable examples
for greater clarity and an added option to enable/disable Fortran 77 examples.
These changes included changing \id{EXAMPLES\_ENABLE} to \id{EXAMPLES\_ENABLE\_C},
changing \id{CXX\_ENABLE} to \id{EXAMPLES\_ENABLE\_CXX}, changing \id{F90\_ENABLE} to
\id{EXAMPLES\_ENABLE\_F90}, and adding an \id{EXAMPLES\_ENABLE\_F77} option.

A bug fix was done to correct the fcmix name translation for \id{FKIN\_SPFGMR}.

Corrections and additions were made to the examples,
to installation-related files,
and to the user documentation.

\subsection*{Changes in v2.9.0}

Two additional {\nvector} implementations were added -- one for
Hypre (parallel) vectors, and one for PETSc vectors.  These
additions are accompanied by additions to various interface functions
and to user documentation.

Each {\nvector} module now includes a function, \id{N\_VGetVectorID},
that returns the {\nvector} module name.

The Picard iteration return was chanegd to always return the newest
iterate upon success.  A minor bug in the line search was fixed
to prevent an infinite loop when the beta condition fails and lamba
is below the minimum size.

For each linear solver, the various solver performance counters are
now initialized to 0 in both the solver specification function and in
solver \id{linit} function.  This ensures that these solver counters
are initialized upon linear solver instantiation as well as at the
beginning of the problem solution.

A memory leak was fixed in the banded preconditioner interface.
In addition, updates were done to return integers from linear solver
and preconditioner 'free' functions.

Corrections were made to three Fortran interface
functions.  The Anderson acceleration scheme was enhanced by use of
QR updating.

The Krylov linear solver Bi-CGstab was enhanced by removing a redundant
dot product.  Various additions and corrections were made to the
interfaces to the sparse solvers KLU and SuperLU\_MT, including support
for CSR format when using KLU.

The functions FKINCREATE and FKININIT were added to split the FKINMALLOC
routine into two pieces.  FKINMALLOC remains for backward compatibility, but
documentation for it has been removed.

A new examples was added for use of the OpenMP vector.

Minor corrections and additions were made to the {\kinsol} solver, to the
Fortran interfaces, to the examples, to installation-related files,
and to the user documentation.

\subsection*{Changes in v2.8.0}

Two major additions were made to the globalization strategy options
(\id{KINSol} argument \id{strategy}).  One is fixed-point iteration,
and the other is Picard iteration.  Both can be accelerated by use
of the Anderson acceleration method.  See the relevant paragraphs in
Chapter \ref{s:math}.

Three additions were made to the linear system solvers that are
available for use with the {\kinsol} solver.  First, in the serial case,
an interface to the sparse direct solver KLU was added.
Second, an interface to SuperLU\_MT, the multi-threaded version of
SuperLU, was added as a thread-parallel sparse direct solver option,
to be used with the serial version of the {\nvector} module.
As part of these additions, a sparse matrix (CSC format) structure
was added to {\kinsol}.  Finally, a variation of GMRES called
Flexible GMRES was added.

Otherwise, only relatively minor modifications were made to {\kinsol}:

In function \id{KINStop}, two return values were corrected to make the values
of \id{uu} and \id{fval} consistent.

A bug involving initialization of \id{mxnewtstep} was fixed.
The error affects the case of repeated user calls to \id{KINSol} with no
intervening call to \id{KINSetMaxNewtonStep}.

A bug in the increments for difference quotient Jacobian approximations
was fixed in function \id{kinDlsBandDQJac}.

In \id{KINLapackBand}, the line \id{smu = MIN(N-1,mu+ml)} was changed to
\id{smu = mu + ml} to correct an illegal input error for \id{DGBTRF/DGBTRS}.

In order to avoid possible name conflicts, the mathematical macro
and function names \id{MIN}, \id{MAX}, \id{SQR}, \id{RAbs}, \id{RSqrt},
\id{RExp}, \id{RPowerI}, and \id{RPowerR} were changed to
\id{SUNMIN}, \id{SUNMAX}, \id{SUNSQR}, \id{SUNRabs}, \id{SUNRsqrt},
\id{SUNRexp}, \id{SRpowerI}, and \id{SUNRpowerR}, respectively.
These names occur in both the solver and in various example programs.

In the FKINSOL module, an incorrect return value \id{ier} in \id{FKINfunc}
was fixed.

In the FKINSOL optional input routines \id{FKINSETIIN}, \id{FKINSETRIN}, and
\id{FKINSETVIN}, the optional fourth argument \id{key\_length} was removed,
with hardcoded key string lengths passed to all \id{strncmp} tests.

In all FKINSOL examples, integer declarations were revised so that
those which must match a C type \id{long int} are declared \id{INTEGER*8},
and a comment was added about the type match.  All other integer
declarations are just \id{INTEGER}.  Corresponding minor corrections were
made to the user guide.

Two new {\nvector} modules have been added for thread-parallel computing
environments --- one for OpenMP, denoted \id{NVECTOR\_OPENMP},
and one for Pthreads, denoted \id{NVECTOR\_PTHREADS}.

With this version of {\sundials}, support and documentation of the
Autotools mode of installation is being dropped, in favor of the
CMake mode, which is considered more widely portable.

\subsection*{Changes in v2.7.0}

One significant design change was made with this release: The problem
size and its relatives, bandwidth parameters, related internal indices,
pivot arrays, and the optional output \id{lsflag} have all been
changed from type \id{int} to type \id{long int}, except for the
problem size and bandwidths in user calls to routines specifying
BLAS/LAPACK routines for the dense/band linear solvers.  The function
\id{NewIntArray} is replaced by a pair \id{NewIntArray}/\id{NewLintArray},
for \id{int} and \id{long int} arrays, respectively.

A large number of errors have been fixed.
Three major logic bugs were fixed -- involving updating the solution
vector, updating the linesearch parameter, and a missing error return.
Three minor errors were fixed -- involving setting \id{etachoice} in the
Matlab/{\kinsol} interface, a missing error case in \id{KINPrintInfo},
and avoiding an exponential overflow in the evaluation of \id{omega}.
In each linear solver interface function, the linear solver memory is
freed on an error return, and the \id{**Free} function now includes a
line setting to NULL the main memory pointer to the linear solver memory.
In the installation files, we modified the treatment of the macro
SUNDIALS\_USE\_GENERIC\_MATH, so that the parameter GENERIC\_MATH\_LIB is
either defined (with no value) or not defined.

\subsection*{Changes in v2.6.0}

This release introduces a new linear solver module, based on BLAS and LAPACK
for both dense and banded matrices.

The user interface has been further refined. Some of the API changes involve:
(a) a reorganization of all linear solver modules into two families (besides
the already present family of scaled preconditioned iterative linear solvers,
the direct solvers, including the new LAPACK-based ones, were also organized
into a {\em direct} family); (b) maintaining a single pointer to user data,
optionally specified through a \id{Set}-type function; (c) a general
streamlining of the band-block-diagonal preconditioner module distributed
with the solver.

\subsection*{Changes in v2.5.0}

The main changes in this release involve a rearrangement of the entire
{\sundials} source tree (see \S\ref{ss:sun_org}). At the user interface
level, the main impact is in the mechanism of including {\sundials} header
files which must now include the relative path (e.g. \id{\#include <cvode/cvode.h>}).
Additional changes were made to the build system: all exported header files are
now installed in separate subdirectories of the installation {\em include} directory.

The functions in the generic dense linear solver (\id{sundials\_dense} and
\id{sundials\_smalldense}) were modified to work for rectangular $m \times n$
matrices ($m \le n$), while the factorization and solution functions were
renamed to \id{DenseGETRF}/\id{denGETRF} and \id{DenseGETRS}/\id{denGETRS},
respectively.
The factorization and solution functions in the generic band linear solver were
renamed \id{BandGBTRF} and \id{BandGBTRS}, respectively.

\subsection*{Changes in v2.4.0}

{\kinspbcg}, {\kinsptfqmr}, {\kindense}, and {\kinband} modules have been
added to interface with the Scaled Preconditioned Bi-CGStab ({\spbcg}),
Scaled Preconditioned Transpose-Free Quasi-Minimal Residual ({\sptfqmr}),
{\dense}, and {\band} linear solver modules, respectively.  (For details
see Chapter \ref{c:usage}.) Corresponding additions were made to the {\F}
interface module {\fkinsol}. At the same time, function type names for
Scaled Preconditioned Iterative Linear Solvers were added for the
user-supplied Jacobian-times-vector and preconditioner setup and solve
functions.

Regarding the {\F} interface module {\fkinsol}, optional inputs are now
set using \id{FKINSETIIN} (integer inputs), \id{FKINSETRIN} (real inputs),
and \id{FKINSETVIN} (vector inputs). Optional outputs are still obtained
from the \id{IOUT} and \id{ROUT} arrays which are owned by the user
and passed as arguments to \id{FKINMALLOC}.

The {\kindense} and {\kinband} linear solver modules include support for
nonlinear residual monitoring which can be used to control Jacobian
updating.

To reduce the possibility of conflicts, the names of all header files have
been changed by adding unique prefixes (\id{kinsol\_} and \id{sundials\_}).
When using the default installation procedure, the header files are exported
under various subdirectories of the target \id{include} directory. For more
details see Appendix \ref{c:install}.

\subsection*{Changes in v2.3.0}

The user interface has been further refined. Several functions used
for setting optional inputs were combined into a single one.
Additionally, to resolve potential variable scope issues, all
SUNDIALS solvers release user data right after its use. The build
system has been further improved to make it more robust.

\subsection*{Changes in v2.2.1}

The changes in this minor {\sundials} release affect only the build system.

\subsection*{Changes in v2.2.0}

The major changes from the previous version involve a redesign of
the user interface across the entire {\sundials} suite. We have
eliminated the mechanism of providing optional inputs and
extracting optional statistics from the solver through the
\id{iopt} and \id{ropt} arrays. Instead, {\kinsol} now provides a
set of routines (with prefix \id{KINSet}) to change the default
values for various quantities controlling the solver and a set of
extraction routines (with prefix \id{KINGet}) to extract
statistics after return from the main solver routine. Similarly,
each linear solver module provides its own set of {\id{Set}-} and
{\id{Get}-type} routines. For more details see Chapter \ref{c:usage}.

Additionally, the interfaces to several user-supplied routines
(such as those providing Jacobian-vector products and
preconditioner information) were simplified by reducing the number
of arguments. The same information that was previously accessible
through such arguments can now be obtained through {\id{Get}-type}
functions.

Installation of {\kinsol} (and all of {\sundials}) has been completely
redesigned and is now based on configure scripts.

\section{Reading this User Guide}\label{ss:reading}

This user guide is a combination of general usage instructions and specific
examples. We expect that some readers will want to concentrate on the general
instructions, while others will refer mostly to the examples, and the
organization is intended to accommodate both styles.

There are different possible levels of usage of {\kinsol}. The most casual
user, with a small nonlinear system, can get by with reading all of
Chapter~\ref{s:math}, then Chapter~\ref{c:usage} through \S\ref{sss:kinsol} only, and
looking at examples in~\cite{kinsol_ex}. In a different direction, a
more expert user with a nonlinear system may want to (a) use a package
preconditioner (\S\ref{sss:kinbbdpre}), (b) supply his/her own Jacobian or
preconditioner routines (\S\ref{ss:user_fct_sol}), (c) supply a new
{\nvector} module (Chapter~\ref{s:nvector}), or even (d) supply a different
linear solver module (\S\ref{ss:kinsol_org} and Chapter~\ref{s:sunlinsol}).

The structure of this document is as follows:
\begin{itemize}
\item
  In Chapter \ref{s:math}, we provide short descriptions of the numerical
  methods implemented by {\kinsol} for the solution of nonlinear systems.
\item
  The following chapter describes the structure of the {\sundials} suite
  of solvers (\S\ref{ss:sun_org}) and the software organization of the {\kinsol}
  solver (\S\ref{ss:kinsol_org}).
\item
  Chapter \ref{c:usage} is the main usage document for {\kinsol} for {\CC} applications.
  It includes a complete description of the user interface for the solution
  of nonlinear algebraic systems.
\item
  In Chapter \ref{s:fcmix}, we describe {\fkinsol}, an interface module for the
  use of {\kinsol} with {\F} applications.
\item
  Chapter \ref{s:nvector} gives a brief overview of the generic {\nvector} module
  shared among the various components of {\sundials}, and details on the four
  {\nvector} implementations provided with {\sundials}.
\item
Chapter \ref{s:sunmatrix} gives a brief overview of the generic
  {\sunmatrix} module shared among the various components of
  {\sundials}, and details on the {\sunmatrix} implementations
  provided with {\sundials}:
  a dense implementation (\S\ref{ss:sunmat_dense}),
  a banded implementation (\S\ref{ss:sunmat_band}) and
  a sparse implementation (\S\ref{ss:sunmat_sparse}).
\item
  Chapter \ref{s:sunlinsol} gives a brief overview of the generic
  {\sunlinsol} module shared among the various components of
  {\sundials}.  This chapter contains details on the {\sunlinsol}
  implementations provided with {\sundials}.
  The chapter
  also contains details on the {\sunlinsol} implementations provided
  with {\sundials} that interface with external linear solver
  libraries.
%% \item
%%   Chapter \ref{s:new_linsolv} describes the interfaces to the linear solver
%%   modules, so that a user can provide his/her own such module.
%% \item
%%   Chapter \ref{s:gen_linsolv} describes in detail the generic linear solvers
%%   shared by all {\sundials} solvers.
\item
  Finally, in the appendices, we provide detailed instructions for the installation
  of {\kinsol}, within the structure of {\sundials} (Appendix \ref{c:install}), as well
  as a list of all the constants used for input to and output from {\kinsol} functions
  (Appendix \ref{c:constants}).
\end{itemize}

Finally, the reader should be aware of the following notational
conventions in this user guide:  program listings and identifiers
(such as \id{KINInit}) within textual explanations appear in
typewriter type style; fields in {\CC} structures (such as {\em
content}) appear in italics; and packages or modules are written
in all capitals.
Usage and installation instructions that constitute important warnings
are marked with a triangular symbol {\warn} in the margin.

\paragraph{Acknowledgments.}
We wish to acknowledge the contributions to previous versions of the
{\kinsol} code and user guide by Allan G. Taylor.


%---------------------------------
% License information section
\input{license_info}
%---------------------------------
