%===================================================================================
\chapter{Description of the SUNLinearSolver module}\label{s:sunlinsol}
%===================================================================================
\index{SUNLinearSolver@\texttt{SUNLinearSolver} module}
\input{sunlinsol_generic}


%---------------------------------------------------------------------------
\section{KINSOL SUNLinearSolver interface}
\label{s:sunlinsol_interface}
%---------------------------------------------------------------------------

Table \ref{t:sunlinsoluse} below lists the {\sunlinsol} module linear solver
functions used within the {\kinls} interface. As with the {\sunmatrix} module, we
emphasize that the {\kinsol} user does not need to know detailed usage of linear
solver functions by the {\kinsol} code modules in order to use {\kinsol}. The
information is presented as an implementation detail for the interested reader.

The linear solver functions listed below are marked with \cm to
indicate that they are required, or with $\dagger$ to indicate that
they are only called if they are non-\id{NULL} in the {\sunlinsol}
implementation that is being used. Note:
\begin{enumerate}
\item \id{SUNLinSolNumIters} is only used to accumulate overall
  iterative linear solver statistics.  If it is not implemented by
  the {\sunlinsol} module, then {\kinls} will consider all solves as
  requiring zero iterations.
\item Although \id{SUNLinSolResNorm} is optional, if it is not
  implemented by the {\sunlinsol} then {\kinls} will consider all 
   solves a being \emph{exact}.
\item Although {\kinls} does not call \id{SUNLinSolLastFlag}
  directly, this routine is available for users to query linear solver
  issues directly.
\item Although {\kinls} does not call \id{SUNLinSolFree}
  directly, this routine should be available for users to call when
  cleaning up from a simulation.
\end{enumerate}

\begin{table}[htb]
\centering
\caption{List of linear solver function usage in the {\kinls} interface}\label{t:sunlinsoluse}
\medskip
\begin{tabular}{|r|c|c|c|} \hline
                                                    & 
\begin{sideways}{DIRECT}             \end{sideways} & 
\begin{sideways}{ITERATIVE}          \end{sideways} & 
\begin{sideways}{MATRIX\_ITERATIVE}  \end{sideways} \\ \hline\hline
%                                  DIRECT       ITER    & MAT-ITER  
\id{SUNLinSolGetType}           &    \cm    &    \cm    &    \cm    \\ \hline
\id{SUNLinSolSetATimes}         & $\dagger$ &    \cm    & $\dagger$ \\ \hline
\id{SUNLinSolSetPreconditioner} & $\dagger$ & $\dagger$ & $\dagger$ \\ \hline
\id{SUNLinSolSetScalingVectors} & $\dagger$ & $\dagger$ & $\dagger$ \\ \hline
\id{SUNLinSolInitialize}        &    \cm    &    \cm    &    \cm    \\ \hline
\id{SUNLinSolSetup}             &    \cm    &    \cm    &    \cm    \\ \hline
\id{SUNLinSolSolve}             &    \cm    &    \cm    &    \cm    \\ \hline
$^1$\id{SUNLinSolNumIters}      &           & $\dagger$ & $\dagger$ \\ \hline
$^2$\id{SUNLinSolResNorm}       &           & $\dagger$ & $\dagger$ \\ \hline
$^3$\id{SUNLinSolLastFlag}      &           &           &           \\ \hline
$^4$\id{SUNLinSolFree}          &           &           &           \\ \hline
\id{SUNLinSolSpace}             & $\dagger$ & $\dagger$ & $\dagger$ \\ \hline
\end{tabular}
\end{table}

Since there are a wide range of potential {\sunlinsol} use cases, the following
subsections describe some details of the {\kinls} interface, in the case that
interested users wish to develop custom {\sunlinsol} modules.

%---------------------------------------------------------------------------
\subsection{Lagged matrix information}
\label{ss:sunlinsol_lagged_matrix}
%---------------------------------------------------------------------------

If the {\sunlinsol} object self-identifies as having type
\id{SUNLINEARSOLVER\_DIRECT} or \\ \noindent
\id{SUNLINEARSOLVER\_MATRIX\_ITERATIVE}, then the {\sunlinsol} object solves a
linear system \emph{defined} by a {\sunmatrix} object. As a result,
{\kinsol} can perform its optional residual monitoring scheme,
described in \S\ref{ss:ModifiedNewtonResidualMon}.

%---------------------------------------------------------------------------
\subsection{Iterative linear solver tolerance}
\label{ss:sunlinsol_iterative_tolerance}
%---------------------------------------------------------------------------

If the {\sunlinsol} object self-identifies as having type
\id{SUNLINEARSOLVER\_ITERATIVE} or \newline
\id{SUNLINEARSOLVER\_MATRIX\_ITERATIVE} then {\kinls} will adjust the linear solver
tolerance \id{delta} as described in \S\ref{ss:InexactNewtonStopCrit} 
during the course of the nonlinear solve process. However, if the
iterative linear solver does not support scaling matrices (i.e., the
\id{SUNLinSolSetScalingVectors} routine is \id{NULL}), then {\kinls} will
be unable to fully handle ill-conditioning in the nonlinear solve process
through the solution and residual scaling operators described in
\S\ref{ss:Scaling}. In this case, {\kinls} will attempt to adjust the linear
solver tolerance to account for this lack of functionality. To this end, the
following assumptions are made:
\begin{enumerate}
\item All residual components have similar magnitude; hence the
  scaling matrix $D_F$ used in computing the linear residual norm (see
  \S\ref{ss:Scaling}) should satisfy the assumption 
  \[
    (D_F)_{i,i} \approx D_{F,mean},\quad \text{for}\quad i=0,\ldots,n-1.
  \]
\item The {\sunlinsol} object uses a standard 2-norm to measure
  convergence.
\end{enumerate}

Since {\kinsol} uses $D_F$ as the left-scaling matrix, $S_1 = D_F$,
then the linear solver convergence requirement is converted as follows
(using the notation from equations
\eqref{eq:transformed_linear_system}-\eqref{eq:transformed_linear_system_components}): 
\begin{align*}
  &\left\| \tilde{b} - \tilde{A} \tilde{x} \right\|_2  <  \text{tol}\\
  \Leftrightarrow \quad & \left\| D_F P_1^{-1} b - D_F P_1^{-1} A x \right\|_2  <  \text{tol}\\
  \Leftrightarrow \quad & \sum_{i=0}^{n-1} \left[(D_F)_{i,i} \left(P_1^{-1} (b - A x)\right)_i\right]^2  <  \text{tol}^2\\
  \Leftrightarrow \quad & D_{F,mean}^2 \sum_{i=0}^{n-1} \left[\left(P_1^{-1} (b - A x)\right)_i\right]^2  <  \text{tol}^2\\
  \Leftrightarrow \quad & \sum_{i=0}^{n-1} \left[\left(P_1^{-1} (b - A x)\right)_i\right]^2  <  \left(\frac{\text{tol}}{D_{F,mean}}\right)^2\\
  \Leftrightarrow \quad & \left\| P_1^{-1} (b - A x)\right\|_2  <  \frac{\text{tol}}{D_{F,mean}}
\end{align*}
Therefore the tolerance scaling factor
\[
  D_{F,mean} = \frac{1}{\sqrt{n}}\left(\sum_{i=0}^{n-1} (D_F)_{i,i}^2\right)^{1/2}
\]
is computed and the scaled tolerance \id{delta}$= \text{tol} / D_{F,mean}$ is
supplied to the {\sunlinsol} object.

%---------------------------------------------------------------------------
% sunlinsol module sections
%---------------------------------------------------------------------------

\input{sunlinsol_dense}
\input{sunlinsol_band}
\input{sunlinsol_lapackdense}
\input{sunlinsol_lapackband}
\input{sunlinsol_klu}
\input{sunlinsol_superludist}
\input{sunlinsol_superlumt}
\input{sunlinsol_spgmr}
\input{sunlinsol_spfgmr}
\input{sunlinsol_spbcgs}
\input{sunlinsol_sptfqmr}
\input{sunlinsol_pcg}
\input{sunlinsol_examples}
