%%========================================================================
\chapter{KINSOL Constants}\label{c:constants}
%%========================================================================

Below we list all input and output constants used by the main solver and 
linear solver modules, together with their numerical values and a short
description of their meaning.

%%-------------------------------------------------------------------------
%% Supertabular setings

\newlength{\tcolone}
\settowidth{\tcolone}{\id{SPTFQMR\_PSOLVE\_FAIL\_UNREC}}
\newlength{\tcoltwo}
\settowidth{\tcoltwo}{-20}
\newlength{\tcolthree}
\setlength{\tcolthree}{\textwidth}
\addtolength{\tcolthree}{-0.5in}
\addtolength{\tcolthree}{-\tcolone}
\addtolength{\tcolthree}{-\tcoltwo}

\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}

%%-------------------------------------------------------------------------

\section{KINSOL input constants}\label{s:kinsol_in_constants}

\begin{supertabular*}{\textwidth}{p{\tcolone}@{\hspace*{2mm}\extracolsep{\fill}}rp{\tcolthree}}

\hline
\multicolumn{3}{c}{\bf {\kinsol} main solver module}\\
\hline\\

\id{KIN\_ETACHOICE1}      & 1 & Use Eisenstat and Walker Choice 1 for $\eta$. \\
\id{KIN\_ETACHOICE2}      & 2 & Use Eisenstat and Walker Choice 2 for $\eta$. \\
\id{KIN\_ETACONSTANT}     & 3 & Use constant value for $\eta$. \\
\id{KIN\_NONE}            & 0 & Use inexact Newton globalization. \\
\id{KIN\_LINESEARCH}      & 1 & Use line search globalization.

\\\hline
\multicolumn{3}{c}{\bf Iterative linear solver module}\\
\hline\\

\id{PREC\_NONE}    & 0 & No preconditioning \\
\id{PREC\_RIGHT}   & 2 & Preconditioning on the right. \\
\id{MODIFIED\_GS}  & 1 & Use modified Gram-Schmidt procedure. \\
\id{CLASSICAL\_GS} & 2 & Use classical Gram-Schmidt procedure. \\

\end{supertabular*}

%%-------------------------------------------------------------------------

\section{KINSOL output constants}\label{s:kinsol_out_constants}

\begin{supertabular*}{\textwidth}{p{\tcolone}@{\hspace*{2mm}\extracolsep{\fill}}rp{\tcolthree}}

\hline
\multicolumn{3}{c}{\bf {\kinsol} main solver module}\\
\hline\\

\id{KIN\_SUCCESS}               &  0  & Successful function return. \\
\id{KIN\_INITIAL\_GUESS\_OK}    &  1  & The initial user-supplied guess already satisfies the stopping criterion. \\
\id{KIN\_STEP\_LT\_STPTOL}      &  2  & The stopping tolerance on scaled step length was satisfied. \\
\id{KIN\_MEM\_NULL}             & -1  & The \id{kin\_mem} argument was \id{NULL}. \\
\id{KIN\_ILL\_INPUT}            & -2  & One of the function inputs is illegal. \\
\id{KIN\_NO\_MALLOC}            & -3  & The {\kinsol} memory was not allocated by a call to \id{KINMalloc}. \\
\id{KIN\_MEM\_FAIL}             & -4  & A memory allocation failed. \\
\id{KIN\_LINESEARCH\_NONCONV}   & -5  & The line search algorithm was unable to find an iterate sufficiently distinct from the current iterate. \\
\id{KIN\_MAXITER\_REACHED}      & -6  & The maximum number of nonlinear iterations has been reached. \\
\id{KIN\_MXNEWT\_5X\_EXCEEDED}  & -7  & Five consecutive steps have been taken that satisfy a scaled step length test. \\
\id{KIN\_LINESEARCH\_BCFAIL}    & -8  & The line search algorithm was unable to satisfy the $\beta$-condition for \id{nbcfails} iterations. \\
\id{KIN\_LINSOLV\_NO\_RECOVERY} & -9  & The user-supplied routine preconditioner slve function failed recoverably, but the preconditioner is already current. \\
\id{KIN\_LINIT\_FAIL}           & -10 & The linear solver's initialization function failed.  \\
\id{KIN\_LSETUP\_FAIL}          & -11 & The linear solver's setup function failed in an unrecoverable manner. \\
\id{KIN\_LSOLVE\_FAIL}          & -12 & The linear solver's solve function failed in an unrecoverable manner. \\
\id{KIN\_SYSFUNC\_FAIL}         & -13 & The system function failed in an unrecoverable manner. \\
\id{KIN\_FIRST\_SYSFUNC\_ERR}   & -14 & The system function failed recoverably at the first call. \\
\id{KIN\_REPTD\_SYSFUNC\_ERR}   & -15 & The system function had repeated recoverable errors. \\

\\\hline
\multicolumn{3}{c}{\bf {\kindense} linear solver module}\\
\hline\\

\id{KINDENSE\_SUCCESS}    &  0 & Successful function return. \\
\id{KINDENSE\_MEM\_NULL}  & -1 & The \id{kin\_mem} argument was \id{NULL}.\\
\id{KINDENSE\_LMEM\_NULL} & -2 & The {\kindense} linear solver has not been initialized.\\
\id{KINDENSE\_ILL\_INPUT} & -3 & The {\kindense} solver is not compatible with the current {\nvector} module.\\
\id{KINDENSE\_MEM\_FAIL}  & -4 & A memory allocation request failed.\\

\\\hline
\multicolumn{3}{c}{\bf {\kinband} linear solver module}\\
\hline\\

\id{KINBAND\_SUCCESS}    &  0 & Successful function return. \\
\id{KINBAND\_MEM\_NULL}  & -1 & The \id{kin\_mem} argument was \id{NULL}.\\
\id{KINBAND\_LMEM\_NULL} & -2 & The {\kinband} linear solver has not been initialized.\\
\id{KINBAND\_ILL\_INPUT} & -3 & The {\kinband} solver is not compatible with the
                        current {\nvector} module, or an input value was illegal.\\
\id{KINBAND\_MEM\_FAIL}  & -4 & A memory allocation request failed.\\

\\\hline
\multicolumn{3}{c}{\bf {\kinspils} linear solver modules}\\
\hline\\

\id{KINSPILS\_SUCCESS}    &  0 & Successful function return. \\
\id{KINSPILS\_MEM\_NULL}  & -1 & The \id{kin\_mem} argument was \id{NULL}.\\
\id{KINSPILS\_LMEM\_NULL} & -2 & The linear solver has not been initialized.\\
\id{KINSPILS\_ILL\_INPUT} & -3 & The solver is not compatible with the current {\nvector} module, or an input value was illegal.\\
\id{KINSPILS\_MEM\_FAIL}  & -4 & A memory allocation request failed.\\

\\\hline
\multicolumn{3}{c}{\bf {\spgmr} generic linear solver module}\\
\hline\\

\id{SPGMR\_SUCCESS}            &  0 & Converged. \\
\id{SPGMR\_RES\_REDUCED}       &  1 & No convergence, but the residual norm was reduced. \\
\id{SPGMR\_CONV\_FAIL}         &  2 & Failure to converge. \\
\id{SPGMR\_QRFACT\_FAIL}       &  3 & A singular matrix was found during the QR factorization. \\
\id{SPGMR\_PSOLVE\_FAIL\_REC}  &  4 & The preconditioner solve function failed recoverably.\\
\id{SPGMR\_MEM\_NULL}          & -1 & The {\spgmr} memory is \id{NULL}\\
\id{SPGMR\_ATIMES\_FAIL}       & -2 & The Jacobian tims vector function failed. \\
\id{SPGMR\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPGMR\_GS\_FAIL}           & -4 & Failure in the Gram-Schmidt procedure. \\
\id{SPGMR\_QRSOL\_FAIL}        & -5 & The matrix $R$ was found to be singular during the QR solve phase. \\

\\\hline
\multicolumn{3}{c}{\bf {\spbcg} generic linear solver module}\\
\hline\\

\id{SPBCG\_SUCCESS}            &  0 & Converged. \\
\id{SPBCG\_RES\_REDUCED}       &  1 & No convergence, but the residual norm was reduced. \\
\id{SPBCG\_CONV\_FAIL}         &  2 & Failure to converge. \\
\id{SPBCG\_PSOLVE\_FAIL\_REC}  &  3 & The preconditioner solve function failed recoverably.\\
\id{SPBCG\_MEM\_NULL}          & -1 & The {\spbcg} memory is \id{NULL}\\
\id{SPBCG\_ATIMES\_FAIL}       & -2 & The Jacobian tims vector function failed. \\
\id{SPBCG\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\

\\\hline
\multicolumn{3}{c}{\bf {\sptfqmr} generic linear solver module}\\
\hline\\

\id{SPTFQMR\_SUCCESS}            &  0 & Converged. \\
\id{SPTFQMR\_RES\_REDUCED}       &  1 & No convergence, but the residual norm was reduced. \\
\id{SPTFQMR\_CONV\_FAIL}         &  2 & Failure to converge. \\
\id{SPTFQMR\_PSOLVE\_FAIL\_REC}  &  3 & The preconditioner solve function failed recoverably.\\
\id{SPTFQMR\_MEM\_NULL}          & -1 & The {\sptfqmr} memory is \id{NULL}\\
\id{SPTFQMR\_ATIMES\_FAIL}       & -2 & The Jacobian tims vector function failed. \\
\id{SPTFQMR\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\

\\\hline
\multicolumn{3}{c}{\bf {\kinbbdpre} preconditioner module}\\
\hline\\

\id{KINBBDPRE\_SUCCESS}         &  0 & Successful function return. \\
\id{KINBBDPRE\_PDATA\_NULL}     & -11 & The preconditioner module has not been initialized. \\


\end{supertabular*} 
