%%========================================================================
\chapter{CVODES Constants}\label{c:constants}
%%========================================================================

Below we list all input and output constants used by the main solver and 
linear solver modules, together with their numerical values and a short
description of their meaning.

%%-------------------------------------------------------------------------
%% Supertabular setings

\newlength{\tcolone}
\settowidth{\tcolone}{\id{SPTFQMR\_PSOLVE\_FAIL\_UNREC}}
\newlength{\tcoltwo}
\settowidth{\tcoltwo}{-109}
\newlength{\tcolthree}
\setlength{\tcolthree}{\textwidth}
\addtolength{\tcolthree}{-0.5in}
\addtolength{\tcolthree}{-\tcolone}
\addtolength{\tcolthree}{-\tcoltwo}

\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}

%%-------------------------------------------------------------------------

\section{CVODES input constants}

\begin{supertabular*}{\textwidth}{p{\tcolone}@{\hspace*{2mm}\extracolsep{\fill}}rp{\tcolthree}}
%%
\hline
\multicolumn{3}{c}{\bf {\cvodes} main solver module}\\
\hline\\
%%
\id{CV\_ADAMS}            & 1 & Adams-Moulton linear multistep method. \\
\id{CV\_BDF}              & 2 & BDF linear multistep method. \\
\id{CV\_FUNCTIONAL}       & 1 & Nonlinear system solution through functional iterations. \\
\id{CV\_NEWTON}           & 2 & Nonlinear system solution through Newton iterations. \\
\id{CV\_NORMAL}           & 1 & Solver returns at specified output time. \\
\id{CV\_ONE\_STEP}        & 2 & Solver returns after each successful step. \\
\id{CV\_SIMULTANEOUS}     & 1 & Simultaneous corrector forward sensitivity method. \\
\id{CV\_STAGGERED}        & 2 & Staggered corrector forward sensitivity method. \\
\id{CV\_STAGGERED1}       & 3 & Staggered (variant) corrector forward sensitivity method. \\
\id{CV\_CENTERED}         & 1 & Central difference quotient approximation ($2^{nd}$ order) of the sensitivity RHS. \\
\id{CV\_FORWARD}          & 2 & Forward difference quotient approximation ($1^{st}$ order) of the sensitivity RHS. \\
%%
\\\hline
\multicolumn{3}{c}{\bf {\cvodes} adjoint solver module}\\
\hline\\
%%
\id{CV\_HERMITE} & 1 & Use Hermite interpolation. \\
\id{CV\_POLYNOMIAL} & 2 & Use variable-degree polynomial interpolation. \\
%%
\\\hline
\multicolumn{3}{c}{\bf Iterative linear solver module}\\
\hline\\
%%
\id{PREC\_NONE} & 0 & No preconditioning \\
\id{PREC\_LEFT} & 1 & Preconditioning on the left only. \\
\id{PREC\_RIGHT} & 2 & Preconditioning on the right only. \\
\id{PREC\_BOTH} & 3 & Preconditioning on both the left and the right. \\
\id{MODIFIED\_GS} & 1 & Use modified Gram-Schmidt procedure. \\
\id{CLASSICAL\_GS} & 2 & Use classical Gram-Schmidt procedure.
%%
\end{supertabular*}


%%-------------------------------------------------------------------------

\section{CVODES output constants}

\begin{supertabular*}{\textwidth}{p{\tcolone}@{\hspace*{2mm}\extracolsep{\fill}}rp{\tcolthree}}

\hline
\multicolumn{3}{c}{\bf {\cvodes} main solver module}\\
\hline\\

\id{CV\_SUCCESS}             &  0  & Successful function return. \\
\id{CV\_TSTOP\_RETURN}       &  1  & \id{CVode} succeeded by reaching the specified stopping point. \\
\id{CV\_ROOT\_RETURN}        &  2  & \id{CVode} succeeded and found one or more roots. \\
\id{CV\_TOO\_MUCH\_WORK}     & -1  & The solver took \id{mxstep} internal steps but could not reach tout.\\
\id{CV\_TOO\_MUCH\_ACC}      & -2  & The solver could not satisfy the accuracy demanded by the user for some internal step.\\
\id{CV\_ERR\_FAILURE}        & -3  & Error test failures occurred too many times during one internal time step or minimum step size was reached. \\
\id{CV\_CONV\_FAILURE}       & -4  & Convergence test failures occurred too many times during one internal time step or minimum step size was reached. \\
\id{CV\_LINIT\_FAIL}         & -5  & The linear solver's initialization function failed.  \\
\id{CV\_LSETUP\_FAIL}        & -6  & The linear solver's setup function failed in an unrecoverable manner. \\
\id{CV\_LSOLVE\_FAIL}        & -7  & The linear solver's solve function failed in an unrecoverable manner. \\
\id{CV\_RHSFUNC\_FAIL}       & -8  & The right-hand side function failed in an unrecoverable manner. \\
\id{CV\_FIRST\_RHSFUNC\_ERR} & -9  & The right-hand side function failed at the first call. \\
\id{CV\_REPTD\_RHSFUNC\_ERR} & -10 & The right-hand side function had repetead recoverable errors. \\
\id{CV\_UNREC\_RHSFUNC\_ERR} & -11 & The right-hand side function had a recoverable error, but no recovery is possible. \\
\id{CV\_RTFUNC\_FAIL}        & -12 & The rootfinding function failed in an unrecoverable manner. \\
\id{CV\_MEM\_FAIL}           & -20 & A memory allocation failed. \\
\id{CV\_MEM\_NULL}           & -21 & The \id{cvode\_mem} argument was \id{NULL}. \\
\id{CV\_ILL\_INPUT}          & -22 & One of the function inputs is illegal. \\
\id{CV\_NO\_MALLOC}          & -23 & The {\cvode} memory block was not allocated by a call to \id{CVodeMalloc}. \\
\id{CV\_BAD\_K}              & -24 & The derivative order $k$ is larger than the order used. \\
\id{CV\_BAD\_T}              & -25 & The time $t$ s outside the last step taken. \\
\id{CV\_BAD\_DKY}            & -26 & The output derivative vector is \id{NULL}. \\
\id{CV\_TOO\_CLOSE}          & -27 & The output and initial times are too close to each other. \\
\id{CV\_NO\_QUAD}             & -30 & Quadrature integration was not activated. \\
\id{CV\_QRHSFUNC\_FAIL}       & -31 & The quadrature right-hand side function failed in an unrecoverable manner. \\
\id{CV\_FIRST\_QRHSFUNC\_ERR} & -32 & The quadrature right-hand side function failed at the first call. \\
\id{CV\_REPTD\_QRHSFUNC\_ERR} & -33 & The quadrature ight-hand side function had repetead recoverable errors. \\
\id{CV\_UNREC\_QRHSFUNC\_ERR} & -34 & The quadrature right-hand side function had a recoverable error, but no recovery is possible. \\
\id{CV\_NO\_SENS}             & -40 & Forward sensitivity integration was not activated. \\
\id{CV\_SRHSFUNC\_FAIL}       & -41 & The sensitivity right-hand side function failed in an unrecoverable manner. \\
\id{CV\_FIRST\_SRHSFUNC\_ERR} & -42 & The sensitivity right-hand side function failed at the first call. \\
\id{CV\_REPTD\_SRHSFUNC\_ERR} & -43 & The sensitivity ight-hand side function had repetead recoverable errors. \\
\id{CV\_UNREC\_SRHSFUNC\_ERR} & -44 & The sensitivity right-hand side function had a recoverable error, but no recovery is possible. \\
\id{CV\_BAD\_IS}              & -45 & The sensitivity index is larger than the number of sensitivities computed.\\
\id{CV\_NO\_QUADSENS}          & -50 & Forward sensitivity integration was not activated. \\
\id{CV\_QSRHSFUNC\_FAIL}       & -51 & The sensitivity right-hand side function failed in an unrecoverable manner. \\
\id{CV\_FIRST\_QSRHSFUNC\_ERR} & -52 & The sensitivity right-hand side function failed at the first call. \\
\id{CV\_REPTD\_QSRHSFUNC\_ERR} & -53 & The sensitivity ight-hand side function had repetead recoverable errors. \\
\id{CV\_UNREC\_QSRHSFUNC\_ERR} & -54 & The sensitivity right-hand side function had a recoverable error, but no recovery is possible. \\

\\\hline
\multicolumn{3}{c}{\bf {\cvodes} adjoint solver module}\\
\hline\\

\id{CV\_NO\_ADJ}      & -101 & Adjoint module was not initialized. \\
\id{CV\_NO_FWD}       & -102 & The forward integration was not yet performed. \\
\id{CV\_NO_BCK}       & -103 & No backward problem was specified. \\
\id{CV\_BAD\_TB0}     & -104 & The final time for the adjoint problem is outside the interval over which the forward problem was solved.\\
\id{CV\_REIFWD\_FAIL} & -105 & Reinitialization of the forward problem failed at the first checkpoint. \\
\id{CV\_FWD\_FAIL}    & -106 & An error occured during the integration of the forward problem.\\
\id{CV\_GETY\_BADT}   & -107 & Wrong time in interpolation function. \\

\\\hline
\multicolumn{3}{c}{\bf {\cvdls} linear solver modules}\\
\hline\\

\id{CVDIRECT\_SUCCESS}          &  0 & Successful function return. \\
\id{CVDIRECT\_MEM\_NULL}        & -1 & The \id{cvode\_mem} argument was \id{NULL}.\\
\id{CVDIRECT\_LMEM\_NULL}       & -2 & The {\cvdense} linear solver has not been initialized.\\
\id{CVDIRECT\_ILL\_INPUT}       & -3 & The {\cvdense} solver is not compatible with the current {\nvector} module.\\
\id{CVDIRECT\_MEM\_FAIL}        & -4 & A memory allocation request failed.\\
\id{CVDIRECT\_JACFUNC\_UNRECVR} & -5 & The Jacobian function failed in an unrecoverable manner. \\
\id{CVDIRECT\_JACFUNC\_RECVR}   & -6 & The Jacobian function had a recoverable error. \\
%%
\id{CVDIRECT\_NO\_ADJ}          & -101 & The combined forward-backward problem has not been initialized.\\
\id{CVDIRECT\_LMEMB\_NULL}      & -102 & The linear solver was not initialized for the backward phase. \\

\\\hline
\multicolumn{3}{c}{\bf {\cvdiag} linear solver module}\\
\hline\\

\id{CVDIAG\_SUCCESS}      &  0 & Successful function return. \\
\id{CVDIAG\_MEM\_NULL}    & -1 & The \id{cvode\_mem} argument was \id{NULL}.\\
\id{CVDIAG\_LMEM\_NULL}   & -2 & The {\cvdiag} linear solver has not been initialized.\\
\id{CVDIAG\_ILL\_INPUT}   & -3 & The {\cvdiag} solver is not compatible with the current {\nvector} module.\\
\id{CVDIAG\_MEM\_FAIL}    & -4 & A memory allocation request failed.\\
\id{CVDIAG\_INV\_FAIL}    & -5 & A diagonal element of the Jacobian was 0. \\
\id{CVDIAG\_RHSFUNC\_UNRECVR} & -5 & The right-hand side function failed in an unrecoverable manner. \\
\id{CVDIAG\_RHSFUNC\_RECVR}   & -6 & The right-hand side function had a recoverable error. \\
%%
\id{CVDIAG\_NO\_ADJ} & -101 & The combined forward-backward problem has not been initialized.\\

\\\hline
\multicolumn{3}{c}{\bf {\cvspils} linear solver modules}\\
\hline\\

\id{CVSPILS\_SUCCESS}      &  0 & Successful function return. \\
\id{CVSPILS\_MEM\_NULL}    & -1 & The \id{cvode\_mem} argument was \id{NULL}.\\
\id{CVSPILS\_LMEM\_NULL}   & -2 & The linear solver has not been initialized.\\
\id{CVSPILS\_ILL\_INPUT}   & -3 & The solver is not compatible with the current {\nvector} module, or an input value was illegal.\\
\id{CVSPILS\_MEM\_FAIL}    & -4 & A memory allocation request failed.\\
\id{CVSPILS\_PMEM\_NULL}   & -5 & The preconditioner module has not been initialized. \\
%%
\id{CVSPILS\_NO\_ADJ}      & -101 & The combined forward-backward problem has not been initialized.\\
\id{CVSPILS\_LMEMB\_NULL}  & -102 & The linear solver was not initialized for the backward phase. \\

\\\hline
\multicolumn{3}{c}{\bf {\spgmr} generic linear solver module}\\
\hline\\

\id{SPGMR\_SUCCESS}             &  0 & Converged. \\
\id{SPGMR\_RES\_REDUCED}        &  1 & No convergence, but the residual norm was reduced. \\
\id{SPGMR\_CONV\_FAIL}          &  2 & Failure to converge. \\
\id{SPGMR\_QRFACT\_FAIL}        &  3 & A singular matrix was found during the QR factorization. \\
\id{SPGMR\_PSOLVE\_FAIL\_REC}   &  4 & The preconditioner solve function failed recoverably.\\
\id{SPGMR\_ATIMES\_FAIL\_REC}   &  5 & The Jacobian-times-vector function failed recoverably.\\
\id{SPGMR\_PSET\_FAIL\_REC}     &  6 & The preconditioner setup function failed recoverably.\\
\id{SPGMR\_MEM\_NULL}           & -1 & The {\spgmr} memory is \id{NULL}\\
\id{SPGMR\_ATIMES\_FAIL\_UNREC} & -2 & The Jacobian-times-vector function failed unrecoverably. \\
\id{SPGMR\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPGMR\_GS\_FAIL}            & -4 & Failure in the Gram-Schmidt procedure. \\
\id{SPGMR\_QRSOL\_FAIL}         & -5 & The matrix $R$ was found to be singular during the QR solve phase. \\
\id{SPGMR\_PSET\_FAIL\_UNREC}   & -6 & The preconditioner setup function failed unrecoverably. \\

\\\hline
\multicolumn{3}{c}{\bf {\spbcg} generic linear solver module}\\
\hline\\

\id{SPBCG\_SUCCESS}             &  0 & Converged. \\
\id{SPBCG\_RES\_REDUCED}        &  1 & No convergence, but the residual norm was reduced. \\
\id{SPBCG\_CONV\_FAIL}          &  2 & Failure to converge. \\
\id{SPBCG\_PSOLVE\_FAIL\_REC}   &  3 & The preconditioner solve function failed recoverably.\\
\id{SPBCG\_ATIMES\_FAIL\_REC}   &  4 & The Jacobian-times-vector function failed recoverably.\\
\id{SPBCG\_PSET\_FAIL\_REC}     &  5 & The preconditioner setup function failed recoverably.\\
\id{SPBCG\_MEM\_NULL}           & -1 & The {\spbcg} memory is \id{NULL}\\
\id{SPBCG\_ATIMES\_FAIL\_UNREC} & -2 & The Jacobian-times-vector function failed unrecoverably. \\
\id{SPBCG\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPBCG\_PSET\_FAIL\_UNREC}   & -4 & The preconditioner setup function failed unrecoverably. \\

\\\hline
\multicolumn{3}{c}{\bf {\sptfqmr} generic linear solver module}\\
\hline\\

\id{SPTFQMR\_SUCCESS}             &  0 & Converged. \\
\id{SPTFQMR\_RES\_REDUCED}        &  1 & No convergence, but the residual norm was reduced. \\
\id{SPTFQMR\_CONV\_FAIL}          &  2 & Failure to converge. \\
\id{SPTFQMR\_PSOLVE\_FAIL\_REC}   &  3 & The preconditioner solve function failed recoverably.\\
\id{SPTFQMR\_ATIMES\_FAIL\_REC}   &  4 & The Jacobian-times-vector function failed recoverably.\\
\id{SPTFQMR\_PSET\_FAIL\_REC}     &  5 & The preconditioner setup function failed recoverably.\\
\id{SPTFQMR\_MEM\_NULL}           & -1 & The {\sptfqmr} memory is \id{NULL}\\
\id{SPTFQMR\_ATIMES\_FAIL\_UNREC} & -2 & The Jacobian-times-vector function failed. \\
\id{SPTFQMR\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPTFQMR\_PSET\_FAIL\_UNREC}   & -4 & The preconditioner setup function failed unrecoverably. \\

\end{supertabular*} 

