%%========================================================================
\chapter{CVODE Constants}\label{c:constants}
%%========================================================================

Below we list all input and output constants used by the main solver and 
linear solver modules, together with their numerical values and a short
description of their meaning.

%%-------------------------------------------------------------------------
%% Supertabular setings

\newlength{\tcolone}
\settowidth{\tcolone}{\id{CVDIAG\_RHSFUNC\_UNRECVR}}
\newlength{\tcoltwo}
\settowidth{\tcoltwo}{-109}
\newlength{\tcolthree}
\setlength{\tcolthree}{\textwidth}
\addtolength{\tcolthree}{-0.5in}
\addtolength{\tcolthree}{-\tcolone}
\addtolength{\tcolthree}{-\tcoltwo}

\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}

%%-------------------------------------------------------------------------

\section{CVODE input constants}\label{s:cvode_in_constants}

\begin{xtabular*}{\textwidth}{p{\tcolone}@{\hspace*{2mm}\extracolsep{\fill}}rp{\tcolthree}}
%%
\hline
\multicolumn{3}{c}{\bf {\cvode} main solver module}\\
\hline\\
%%
\id{CV\_ADAMS}            & 1 & Adams-Moulton linear multistep method. \\
\id{CV\_BDF}              & 2 & BDF linear multistep method. \\
\id{CV\_NORMAL}           & 1 & Solver returns at specified output time. \\
\id{CV\_ONE\_STEP}        & 2 & Solver returns after each successful step. \\
%%
\\\hline
\multicolumn{3}{c}{\bf Iterative linear solver modules}\\
\hline\\
%%
\id{PREC\_NONE} & 0 & No preconditioning \\
\id{PREC\_LEFT} & 1 & Preconditioning on the left only. \\
\id{PREC\_RIGHT} & 2 & Preconditioning on the right only. \\
\id{PREC\_BOTH} & 3 & Preconditioning on both the left and the right. \\
\id{MODIFIED\_GS} & 1 & Use modified Gram-Schmidt procedure. \\
\id{CLASSICAL\_GS} & 2 & Use classical Gram-Schmidt procedure.
%%
\end{xtabular*}


%%-------------------------------------------------------------------------

\section{CVODE output constants}\label{s:cvode_out_constants}

\begin{xtabular*}{\textwidth}{p{\tcolone}@{\hspace*{2mm}\extracolsep{\fill}}rp{\tcolthree}}

\hline
\multicolumn{3}{c}{\bf {\cvode} main solver module}\\
\hline\\

\id{CV\_SUCCESS}             &  0  & Successful function return. \\
\id{CV\_TSTOP\_RETURN}       &  1  & \id{CVode} succeeded by reaching the specified stopping point. \\
\id{CV\_ROOT\_RETURN}        &  2  & \id{CVode} succeeded and found one or more roots. \\
\id{CV\_WARNING}             & 99  & \id{CVode} succeeded but an unusual situation occurred. \\
\id{CV\_TOO\_MUCH\_WORK}     & -1  & The solver took \id{mxstep} internal steps but could not reach tout.\\
\id{CV\_TOO\_MUCH\_ACC}      & -2  & The solver could not satisfy the accuracy demanded by the user for some internal step.\\
\id{CV\_ERR\_FAILURE}        & -3  & Error test failures occurred too many times during one internal time step or minimum step size was reached. \\
\id{CV\_CONV\_FAILURE}       & -4  & Convergence test failures occurred too many times during one internal time step or minimum step size was reached. \\
\id{CV\_LINIT\_FAIL}         & -5  & The linear solver's initialization function failed.  \\
\id{CV\_LSETUP\_FAIL}        & -6  & The linear solver's setup function failed in an unrecoverable manner. \\
\id{CV\_LSOLVE\_FAIL}        & -7  & The linear solver's solve function failed in an unrecoverable manner. \\
\id{CV\_RHSFUNC\_FAIL}       & -8  & The right-hand side function failed in an unrecoverable manner. \\
\id{CV\_FIRST\_RHSFUNC\_ERR} & -9  & The right-hand side function failed at the first call. \\
\id{CV\_REPTD\_RHSFUNC\_ERR} & -10 & The right-hand side function had repetead recoverable errors. \\
\id{CV\_UNREC\_RHSFUNC\_ERR} & -11 & The right-hand side function had a recoverable error, but no recovery is possible. \\
\id{CV\_RTFUNC\_FAIL}        & -12 & The rootfinding function failed in an unrecoverable manner. \\
\id{CV\_NLS\_INIT\_FAIL}     & -13 & The nonlinear solver's init routine failed. \\
\id{CV\_NLS\_SETUP\_FAIL}    & -14 & The nonlinear solver's setup routine failed. \\
\id{CV\_CONSTR\_FAIL}        & -15 & The inequality constraints were violated and the solver was unable to recover. \\
\id{CV\_MEM\_FAIL}           & -20 & A memory allocation failed. \\
\id{CV\_MEM\_NULL}           & -21 & The \id{cvode\_mem} argument was \id{NULL}. \\
\id{CV\_ILL\_INPUT}          & -22 & One of the function inputs is illegal. \\
\id{CV\_NO\_MALLOC}          & -23 & The {\cvode} memory block was not allocated by a call to \id{CVodeMalloc}. \\
\id{CV\_BAD\_K}              & -24 & The derivative order $k$ is larger than the order used. \\
\id{CV\_BAD\_T}              & -25 & The time $t$ is outside the last step taken. \\
\id{CV\_BAD\_DKY}            & -26 & The output derivative vector is \id{NULL}. \\
\id{CV\_TOO\_CLOSE}          & -27 & The output and initial times are too close to each other. \\

\\\hline
\multicolumn{3}{c}{\bf {\cvls} linear solver interface}\\
\hline\\

\id{CVLS\_SUCCESS}          &  0 & Successful function return. \\
\id{CVLS\_MEM\_NULL}        & -1 & The \id{cvode\_mem} argument was \id{NULL}.\\
\id{CVLS\_LMEM\_NULL}       & -2 & The {\cvls} linear solver has not been initialized.\\
\id{CVLS\_ILL\_INPUT}       & -3 & The {\cvls} solver is not compatible with the current {\nvector} module.\\
\id{CVLS\_MEM\_FAIL}        & -4 & A memory allocation request failed.\\
\id{CVLS\_PMEM\_NULL}       & -5 & The preconditioner module has not been initialized. \\
\id{CVLS\_JACFUNC\_UNRECVR} & -6 & The Jacobian function failed in an unrecoverable manner. \\
\id{CVLS\_JACFUNC\_RECVR}   & -7 & The Jacobian function had a recoverable error. \\
\id{CVLS\_SUNMAT\_FAIL}     & -8 & An error occurred with the current {\sunmatrix} module. \\
\id{CVLS\_SUNLS\_FAIL}      & -9 & An error occurred with the current {\sunlinsol} module. \\

\\\hline
\multicolumn{3}{c}{\bf {\cvdiag} linear solver module}\\
\hline\\

\id{CVDIAG\_SUCCESS}      &  0 & Successful function return. \\
\id{CVDIAG\_MEM\_NULL}    & -1 & The \id{cvode\_mem} argument was \id{NULL}.\\
\id{CVDIAG\_LMEM\_NULL}   & -2 & The {\cvdiag} linear solver has not been initialized.\\
\id{CVDIAG\_ILL\_INPUT}   & -3 & The {\cvdiag} solver is not compatible with the current {\nvector} module.\\
\id{CVDIAG\_MEM\_FAIL}    & -4 & A memory allocation request failed.\\
\id{CVDIAG\_INV\_FAIL}    & -5 & A diagonal element of the Jacobian was 0. \\
\id{CVDIAG\_RHSFUNC\_UNRECVR} & -6 & The right-hand side function failed in an unrecoverable manner. \\
\id{CVDIAG\_RHSFUNC\_RECVR}   & -7 & The right-hand side function had a recoverable error. \\

\end{xtabular*} 

