%===================================================================================
\chapter{Description of the NVECTOR module}\label{s:nvector}
%===================================================================================
\index{NVECTOR@\texttt{NVECTOR} module}
\input{nvec_generic}

%---------------------------------------------------------------------------
\section{The NVECTOR\_SERIAL implementation}\label{ss:nvec_ser}
\input{nvec_serial}

%---------------------------------------------------------------------------
\section{The NVECTOR\_PARALLEL implementation}\label{ss:nvec_par}
\input{nvec_parallel}

%---------------------------------------------------------------------------
\section{The NVECTOR\_OPENMP implementation}\label{ss:nvec_openmp}
\input{nvec_openmp}

%---------------------------------------------------------------------------
\section{The NVECTOR\_PTHREADS implementation}\label{ss:nvec_pthreads}
\input{nvec_pthreads}

%---------------------------------------------------------------------------
\section{NVECTOR functions used by CVODE}

In Table \ref{t:nvecuse} below, we list the vector functions in the 
{\nvector} module within the {\cvode} package.
The table also shows, for each function, which of the code modules uses
the function. The {\cvode} column shows function usage within the main
integrator module, while the remaining seven columns show function usage
within each of the eight {\cvode} linear solvers, the {\cvbandpre} and
{\cvbbdpre} preconditioner modules, and the {\fcvode} module.  Here
{\cvdls} stands for {\cvdense} and {\cvband}; {\cvspils} stands for
{\cvspgmr}, {\cvspbcg}, and {\cvsptfqmr}; and {\cvsls} stands for
{\cvklu} and {\cvsuperlumt}.

There is one subtlety in the {\cvspils} column hidden by the table, explained
here for the case of the {\cvspgmr} module. 
The \id{N\_VDotProd} function is called both within the interface file
\id{cvode\_spgmr.c} and within the implementation
files \id{sundials\_spgmr.c} and \id{sundials\_iterative.c} for the generic
{\spgmr} solver upon which the {\cvspgmr} solver is built.  Also, although
\id{N\_VDiv} and \id{N\_VProd} are not called within the interface file
\id{cvode\_spgmr.c}, they are called within the implementation file
\id{sundials\_spgmr.c}, and so are required by the {\cvspgmr} solver module.
Analogous statements apply to the {\cvspbcg} and {\cvsptfqmr} modules,
except that they do not use \id{sundials\_iterative.c}.
This issue does not arise for the other 
three {\cvode} linear solvers because the generic {\dense} and {\band} solvers 
(used in the implementation of {\cvdense} and {\cvband}) do not make calls to 
any vector functions and {\cvdiag} is not implemented using a generic diagonal solver. 

At this point, we should emphasize that the {\cvode} user does not need to know 
anything about the usage of vector functions by the {\cvode} code modules in order 
to use {\cvode}. The information is presented as an implementation detail for the 
interested reader.

\begin{table}[htb]
\centering
\caption{List of vector functions usage by {\cvode} code modules}\label{t:nvecuse}
\medskip
\begin{tabular}{|r|c|c|c|c|c|c|c|c|} \hline
                                             & 
\begin{sideways}{\cvode}      \end{sideways} & 
\begin{sideways}{\cvdls}      \end{sideways} & 
\begin{sideways}{\cvdiag}     \end{sideways} & 
\begin{sideways}{\cvspils}    \end{sideways} &
\begin{sideways}{\cvsls}      \end{sideways} &
\begin{sideways}{\cvbandpre}  \end{sideways} &
\begin{sideways}{\cvbbdpre}   \end{sideways} &
\begin{sideways}{\fcvode}     \end{sideways} \\ \hline\hline
%                         CVODE  DLS  DIAG  SPILS  SLS  BPRE   BBD   FCVODE
\id{N\_VClone}           & \cm &     & \cm & \cm &     &     &     &     \\ \hline
\id{N\_VCloneEmpty}      &     &     &     &     &     &     &     & \cm \\ \hline
\id{N\_VDestroy}         & \cm &     & \cm & \cm &     &     &     &     \\ \hline
\id{N\_VSpace}           & \cm &     &     &     &     &     &     &     \\ \hline
\id{N\_VGetArrayPointer} &     & \cm &     &     & \cm & \cm & \cm & \cm \\ \hline
\id{N\_VSetArrayPointer} &     & \cm &     &     &     &     &     & \cm \\ \hline
\id{N\_VLinearSum}       & \cm & \cm & \cm & \cm &     &     &     &     \\ \hline
\id{N\_VConst}           & \cm &     &     & \cm &     &     &     &     \\ \hline
\id{N\_VProd}            & \cm &     & \cm & \cm &     &     &     &     \\ \hline
\id{N\_VDiv}             & \cm &     & \cm & \cm &     &     &     &     \\ \hline
\id{N\_VScale}           & \cm & \cm & \cm & \cm & \cm & \cm & \cm &     \\ \hline
\id{N\_VAbs}             & \cm &     &     &     &     &     &     &     \\ \hline
\id{N\_VInv}             & \cm &     & \cm &     &     &     &     &     \\ \hline
\id{N\_VAddConst}        & \cm &     & \cm &     &     &     &     &     \\ \hline
\id{N\_VDotProd}         &     &     &     & \cm &     &     &     &     \\ \hline
\id{N\_VMaxNorm}         & \cm &     &     &     &     &     &     &     \\ \hline
\id{N\_VWrmsNorm}        & \cm & \cm &     & \cm &     & \cm & \cm &     \\ \hline
\id{N\_VMin}             & \cm &     &     &     &     &     &     &     \\ \hline
\id{N\_VCompare}         &     &     & \cm &     &     &     &     &     \\ \hline
\id{N\_VInvTest}         &     &     & \cm &     &     &     &     &     \\ \hline
\end{tabular}
\end{table}

The vector functions listed in Table \ref{t:nvecops} that are {\em not} used by
{\cvode} are: \id{N\_VWL2Norm}, \id{N\_VL1Norm}, \id{N\_VWrmsNormMask}, 
\id{N\_VConstrMask}, and \id{N\_VMinQuotient}.
Therefore a user-supplied {\nvector} module for {\cvode} could omit
these five functions.

