%%========================================================================
\chapter{IDAS Constants}\label{c:constants}
%%========================================================================

Below we list all input and output constants used by the main solver and 
linear solver modules, together with their numerical values and a short
description of their meaning.

%%-------------------------------------------------------------------------
%% Supertabular setings

\newlength{\tcolone}
\settowidth{\tcolone}{\id{SPTFQMR\_PSOLVE\_FAIL\_UNREC}}
\newlength{\tcoltwo}
\settowidth{\tcoltwo}{-20}
\newlength{\tcolthree}
\setlength{\tcolthree}{\textwidth}
\addtolength{\tcolthree}{-0.5in}
\addtolength{\tcolthree}{-\tcolone}
\addtolength{\tcolthree}{-\tcoltwo}

\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}

%%-------------------------------------------------------------------------

\section{IDAS input constants}\label{s:ida_in_constants}

\begin{supertabular*}{\textwidth}{p{\tcolone}@{\hspace*{2mm}\extracolsep{\fill}}rp{\tcolthree}}

\hline
\multicolumn{3}{c}{\bf {\idas} main solver module}\\
\hline\\

\id{IDA\_NORMAL}           & 1 & Solver returns at specified output time. \\
\id{IDA\_ONE\_STEP}        & 2 & Solver returns after each successful step. \\
\id{IDA\_SIMULTANEOUS}     & 1 & Simultaneous corrector forward sensitivity method. \\
\id{IDA\_STAGGERED}        & 2 & Staggered corrector forward sensitivity method. \\
\id{IDA\_CENTERED}         & 1 & Central difference quotient approximation ($2^{nd}$ order) of the sensitivity RHS. \\
\id{IDA\_FORWARD}          & 2 & Forward difference quotient approximation ($1^{st}$ order) of the sensitivity RHS. \\
\id{IDA\_YA\_YDP\_INIT}    & 1 & Compute $y_a$ and $y^\prime_d$, given $y_d$.\\
\id{IDA\_Y\_INIT}          & 2 & Compute $y$, given $y^\prime$.\\

\\\hline
\multicolumn{3}{c}{\bf {\idas} adjoint solver module}\\
\hline\\
%%
\id{IDA\_HERMITE} & 1 & Use Hermite interpolation. \\
\id{IDA\_POLYNOMIAL} & 2 & Use variable-degree polynomial interpolation. \\



\\\hline
\multicolumn{3}{c}{\bf Iterative linear solver module}\\
\hline\\

\id{PREC\_NONE}  &  0 & No preconditioning \\
\id{PREC\_LEFT}  &  1 & Preconditioning on the left. \\
\id{MODIFIED\_GS}  & 1 & Use modified Gram-Schmidt procedure. \\
\id{CLASSICAL\_GS} & 2 & Use classical Gram-Schmidt procedure. \\

\end{supertabular*}

%%-------------------------------------------------------------------------

\section{IDAS output constants}\label{s:ida_out_constants}

\begin{supertabular*}{\textwidth}{p{\tcolone}@{\hspace*{2mm}\extracolsep{\fill}}rp{\tcolthree}}

\hline
\multicolumn{3}{c}{\bf {\idas} main solver module}\\
\hline\\
\id{IDA\_SUCCESS}         &  0  & Successful function return. \\
\id{IDA\_TSTOP\_RETURN}   &  1  & \id{IDASolve} succeeded by reaching the specified stopping point. \\
\id{IDA\_ROOT\_RETURN}    &  2  & \id{IDASolve} succeeded and found one or more roots. \\
\id{IDA\_WARNING}         & 99  & \id{IDASolve} succeeded but an unusual situation occured. \\
\id{IDA\_TOO\_MUCH\_WORK} & -1  & The solver took \id{mxstep} internal steps but could not reach tout.\\
\id{IDA\_TOO\_MUCH\_ACC}  & -2  & The solver could not satisfy the accuracy demanded by the user for some internal step.\\
\id{IDA\_ERR\_FAIL}       & -3  & Error test failures occurred too many times during one internal time step or minimum step size was reached. \\
\id{IDA\_CONV\_FAIL}      & -4  & Convergence test failures occurred too many times during one internal time step or minimum step size was reached. \\
\id{IDA\_LINIT\_FAIL}     & -5  & The linear solver's initialization function failed.  \\
\id{IDA\_LSETUP\_FAIL}    & -6  & The linear solver's setup function failed in an unrecoverable manner. \\
\id{IDA\_LSOLVE\_FAIL}    & -7  & The linear solver's solve function failed in an unrecoverable manner. \\
\id{IDA\_RES\_FAIL}       & -8  & The user-provided residual function failed in an unrecoverable manner. \\
\id{IDA\_REP\_RES\_FAIL}  & -9  & The user-provided residual function repeatedly returned a recoverable error flag, but the solver was unable to recover. \\
\id{IDA\_RTFUNC\_FAIL}    & -10 & The rootfinding function failed in an unrecoverable manner. \\
\id{IDA\_CONSTR\_FAIL}    & -11 & The inequality constraints were violated and the solver was unable to recover. \\
\id{IDA\_FIRST\_RES\_FAIL}& -12 & The user-provided residual function failed recoverably on the first call. \\
\id{IDA\_LINESEARCH\_FAIL}& -13 & The line search failed. \\
\id{IDA\_NO\_RECOVERY}    & -14 & The residual function, linear solver setup function, or linear solver solve function had a recoverable failure, but \id{IDACalcIC} could not recover. \\
\id{IDA\_MEM\_NULL}       & -20  & The \id{ida\_mem} argument was \id{NULL}. \\
\id{IDA\_MEM\_FAIL}       & -21 & A memory allocation failed. \\
\id{IDA\_ILL\_INPUT}      & -22 & One of the function inputs is illegal. \\
\id{IDA\_NO\_MALLOC}      & -23 & The {\idas} memory was not allocated by a call to \id{IDAMalloc}. \\
\id{IDA\_BAD\_EWT}        & -24 & Zero value of some error weight component. \\
\id{IDA\_BAD\_K}          & -25 & The $k$-th derivative is not available. \\
\id{IDA\_BAD\_T}          & -26 & The time $t$ s outside the last step taken. \\
\id{IDA\_BAD\_DKY}        & -26 & The vector argument where derivative should be stored is \id{NULL}. \\
\id{IDA\_NO\_QUAD}           & -30 & Quadratures were not initialized. \\
\id{IDA\_QRHS\_FAIL}         & -31 & The user-provided right-hand side function for quadratures failed in an unrecoverable manner.\\
\id{IDA\_FIRST\_QRHS\_ERR}   & -32 & The user-provided right-hand side function for quadratures failed in an unrecoverable manner on the first call.\\
\id{IDA\_REP\_QRHS\_ERR}     & -33 & The user-provided right-hand side repeatedly returned a recoverable error flag, but the solver was unable to recover. \\
\id{IDA\_NO\_SENS}           & -40 & Sensitivities were not initialized.\\
\id{IDA\_SRES\_FAIL}         & -41 & The user-provided sensitivity residual function failed in an unrecoverable manner.\\
\id{IDA\_REP\_SRES\_ERR}     & -42 & The user-provided sensitivity residual function repeatedly returned a recoverable error flag, but the solver was unable to recover.\\
\id{IDA\_BAD\_IS}            & -43 & The sensitivity identifier is not valid.\\
\id{IDA\_NO\_QUADSENS}       & -50 & Sensitivity-dependent quadratures were not initialized.\\
\id{IDA\_QSRHS\_FAIL}        & -51 & The user-provided sensitivity-dependent quadrature right-hand side function failed in an unrecoverable manner.\\
\id{IDA\_FIRST\_QSRHS\_ERR}  & -52 & The user-provided sensitivity-dependent quadrature right-hand side function failed in an unrecoverable manner on the first call.\\\\
\id{IDA\_REP\_QSRHS\_ERR}    & -53 & The user-provided sensitivity-dependent quadrature right-hand side repeatedly returned a recoverable error flag, but the solver was unable to recover. \\

\\\hline
\multicolumn{3}{c}{\bf {\idas} adjoint solver module}\\
\hline\\

\id{IDA\_NO\_ADJ}            & -101 & The combined forward-backward problem has not been initialized.\\
\id{IDA\_NO\_FWD}            & -102 & \id{IDASolveF} has not been previously called. \\
\id{IDA\_NO\_BCK}            & -103 & No backward problem was specified. \\
\id{IDA\_BAD\_TB0}           & -104 & The desired output for backward problem is outside the interval over which the forward problem was solved. \\
\id{IDA\_REIFWD\_FAIL}       & -105 & No checkpoint is available for this hot start. \\
\id{IDA\_FWD\_FAIL}          & -106 & \id{IDASolveB} failed because \id{IDASolve} was unable to store data between two consecutive checkpoints. \\
\id{IDA\_GETY\_BADT}         & -107 & Wrong time in interpolation function. \\

\\\hline
\multicolumn{3}{c}{\bf {\idadls} linear solver modules}\\
\hline\\

\id{IDADLS\_SUCCESS}    &  0 & Successful function return. \\
\id{IDADLS\_MEM\_NULL}  & -1 & The \id{ida\_mem} argument was \id{NULL}.\\
\id{IDADLS\_LMEM\_NULL} & -2 & The {\idadls} linear solver has not been initialized.\\
\id{IDADLS\_ILL\_INPUT} & -3 & The {\idadls} solver is not compatible with the current {\nvector} module.\\
\id{IDADLS\_MEM\_FAIL}  & -4 & A memory allocation request failed.\\
\id{IDADLS\_JACFUNC\_UNRECVR} & -5 & The Jacobian function failed in an unrecoverable manner. \\
\id{IDADLS\_JACFUNC\_RECVR}   & -6 & The Jacobian function had a recoverable error. \\
%%
\id{IDADLS\_NO\_ADJ}          & -101 & The combined forward-backward problem has not been initialized.\\
\id{IDADLS\_LMEMB\_NULL}      & -102 & The linear solver was not initialized for the backward phase. \\

\\\hline
\multicolumn{3}{c}{\bf {\idaspils} linear solver modules}\\
\hline\\

\id{IDASPILS\_SUCCESS}    &  0 & Successful function return. \\
\id{IDASPILS\_MEM\_NULL}  & -1 & The \id{ida\_mem} argument was \id{NULL}.\\
\id{IDASPILS\_LMEM\_NULL} & -2 & The {\idaspils} linear solver has not been initialized.\\
\id{IDASPILS\_ILL\_INPUT} & -3 & The {\idaspils} solver is not compatible with the current {\nvector} module.\\
\id{IDASPILS\_MEM\_FAIL}  & -4 & A memory allocation request failed.\\
\id{IDASPILS\_PMEM\_NULL} & -5 & The preconditioner module has not been initialized. \\
%%
\id{IDASPILS\_NO\_ADJ}      & -101 & The combined forward-backward problem has not been initialized.\\
\id{IDASPILS\_LMEMB\_NULL}  & -102 & The linear solver was not initialized for the backward phase. \\


\\\hline
\multicolumn{3}{c}{\bf {\spgmr} generic linear solver module}\\
\hline\\

\id{SPGMR\_SUCCESS}             &  0 & Converged. \\
\id{SPGMR\_RES\_REDUCED}        &  1 & No convergence, but the residual norm was reduced. \\
\id{SPGMR\_CONV\_FAIL}          &  2 & Failure to converge. \\
\id{SPGMR\_QRFACT\_FAIL}        &  3 & A singular matrix was found during the QR factorization. \\
\id{SPGMR\_PSOLVE\_FAIL\_REC}   &  4 & The preconditioner solve function failed recoverably.\\
\id{SPGMR\_ATIMES\_FAIL\_REC}   &  5 & The Jacobian-times-vector function failed recoverably.\\
\id{SPGMR\_PSET\_FAIL\_REC}     &  6 & The preconditioner setup function failed recoverably.\\
\id{SPGMR\_MEM\_NULL}           & -1 & The {\spgmr} memory is \id{NULL}\\
\id{SPGMR\_ATIMES\_FAIL\_UNREC} & -2 & The Jacobian-times-vector function failed unrecoverably. \\
\id{SPGMR\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPGMR\_GS\_FAIL}            & -4 & Failure in the Gram-Schmidt procedure. \\
\id{SPGMR\_QRSOL\_FAIL}         & -5 & The matrix $R$ was found to be singular during the QR solve phase. \\
\id{SPGMR\_PSET\_FAIL\_UNREC}   & -6 & The preconditioner setup function failed unrecoverably. \\

\\\hline
\multicolumn{3}{c}{\bf {\spbcg} generic linear solver module}\\
\hline\\

\id{SPBCG\_SUCCESS}             &  0 & Converged. \\
\id{SPBCG\_RES\_REDUCED}        &  1 & No convergence, but the residual norm was reduced. \\
\id{SPBCG\_CONV\_FAIL}          &  2 & Failure to converge. \\
\id{SPBCG\_PSOLVE\_FAIL\_REC}   &  3 & The preconditioner solve function failed recoverably.\\
\id{SPBCG\_ATIMES\_FAIL\_REC}   &  4 & The Jacobian-times-vector function failed recoverably.\\
\id{SPBCG\_PSET\_FAIL\_REC}     &  5 & The preconditioner setup function failed recoverably.\\
\id{SPBCG\_MEM\_NULL}           & -1 & The {\spbcg} memory is \id{NULL}\\
\id{SPBCG\_ATIMES\_FAIL\_UNREC} & -2 & The Jacobian-times-vector function failed unrecoverably. \\
\id{SPBCG\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPBCG\_PSET\_FAIL\_UNREC}   & -4 & The preconditioner setup function failed unrecoverably. \\

\\\hline
\multicolumn{3}{c}{\bf {\sptfqmr} generic linear solver module}\\
\hline\\

\id{SPTFQMR\_SUCCESS}             &  0 & Converged. \\
\id{SPTFQMR\_RES\_REDUCED}        &  1 & No convergence, but the residual norm was reduced. \\
\id{SPTFQMR\_CONV\_FAIL}          &  2 & Failure to converge. \\
\id{SPTFQMR\_PSOLVE\_FAIL\_REC}   &  3 & The preconditioner solve function failed recoverably.\\
\id{SPTFQMR\_ATIMES\_FAIL\_REC}   &  4 & The Jacobian-times-vector function failed recoverably.\\
\id{SPTFQMR\_PSET\_FAIL\_REC}     &  5 & The preconditioner setup function failed recoverably.\\
\id{SPTFQMR\_MEM\_NULL}           & -1 & The {\sptfqmr} memory is \id{NULL}\\
\id{SPTFQMR\_ATIMES\_FAIL\_UNREC} & -2 & The Jacobian-times-vector function failed. \\
\id{SPTFQMR\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPTFQMR\_PSET\_FAIL\_UNREC}   & -4 & The preconditioner setup function failed unrecoverably. \\
\end{supertabular*} 
