%===================================================================================
\chapter{Description of the NVECTOR module}\label{s:nvector}
%===================================================================================
\index{NVECTOR@\texttt{NVECTOR} module}
\input{nvec_generic}

%---------------------------------------------------------------------------
\section{The NVECTOR\_SERIAL implementation}\label{ss:nvec_ser}
\input{nvec_serial}

%---------------------------------------------------------------------------
\section{The NVECTOR\_PARALLEL implementation}\label{ss:nvec_par}
\input{nvec_parallel}

%---------------------------------------------------------------------------
\section{NVECTOR kernels used by CVODES}

In Table \ref{t:nvecuse} below, we list the vector kernels in the 
{\nvector} module within the {\cvodes} package.
The table also shows, for each kernel, which of the code modules uses
the kernel. The {\cvodes} column shows kernel usage within the main
integrator module, while the remaining four columns show kernel usage
within each of the four {\cvodes} linear solvers. 

There is one subtlety in the {\cvspgmr} column hidden by the table. 
The dot product kernel \id{N\_VDotProd} is called both within the 
implementation file \id{cvspgmr.c} for the {\cvspgmr} solver and within 
the implementation files \id{spgmr.c} and \id{iterative.c} for the generic {\spgmr} 
solver upon which the {\cvspgmr} solver is implemented. 
This issue does not arise for the other 
three {\cvodes} linear solvers because the generic {\dense} and {\band} solvers 
(used in the implementation of {\cvdense} and {\cvband}) do not make calls to 
any vector kernels and {\cvdiag} is not implemented using a generic diagonal solver. 

The vector kernels \id{N\_VMake}, \id{N\_VDispose}, \id{N\_VGetData}, and
\id{N\_VSetData} are only called by the {\cvodes} direct linear solvers, 
{\cvdense} and {\cvband}, and by preconditioner solve routines that use a 
direct linear solver ({\cvbandpre}  and {\cvbbdpre}). They duplicate 
functionality provided by macros in particular implementations of {\nvector},
but are part of the specifications of the generic {\nvector} module to insure 
that the {\cvodes} package is not dependent on any particular {\nvector} implementation.

At this point, we should emphasize that the {\cvodes} user does not need to know 
anything about the usage of vector kernels by the {\cvodes} code modules in order 
to use {\cvodes}. The information is presented as an implementation detail for the 
interested reader.

\begin{table}[htb]
\centering
\caption{List of vector functions usage by {\cvodes} code modules}\label{t:nvecuse}
\medskip
\begin{tabular}{|r|c|c|c|c|c|c|c|c|} \hline
                                             & 
\begin{sideways}{\cvode}      \end{sideways} & 
\begin{sideways}{\cvdense}    \end{sideways} & 
\begin{sideways}{\cvband}     \end{sideways} & 
\begin{sideways}{\cvdiag}     \end{sideways} & 
\begin{sideways}{\cvspgmr}    \end{sideways} &
\begin{sideways}{\cvbandpre}  \end{sideways} &
\begin{sideways}{\cvbbdpre}   \end{sideways} &
\begin{sideways}{\cvodea}     \end{sideways} \\ \hline\hline
\id{N\_VNew}       & \cm &     &     & \cm & \cm &     &     & \cm \\ \hline
\id{N\_VFree}      & \cm &     &     & \cm & \cm &     &     & \cm \\ \hline
\id{N\_VSpace}     & \cm &     &     &     & \cm &     &     &     \\ \hline
\id{N\_VNew\_S}    & \cm &     &     &     &     &     &     &     \\ \hline
\id{N\_VFree\_S}   & \cm &     &     &     &     &     &     &     \\ \hline
\id{N\_VMake}      &     & \cm & \cm &     &     &     &     &     \\ \hline
\id{N\_VDispose}   &     & \cm & \cm &     &     &     &     &     \\ \hline
\id{N\_VGetData}   &     & \cm & \cm &     &     & \cm & \cm &     \\ \hline
\id{N\_VSetData}   &     & \cm & \cm &     &     & \cm & \cm &     \\ \hline
\id{N\_VLinearSum} & \cm & \cm &     & \cm & \cm &     &     & \cm \\ \hline
\id{N\_VConst}     & \cm &     &     &     & \cm &     &     &     \\ \hline
\id{N\_VProd}      & \cm &     &     & \cm & \cm &     &     &     \\ \hline
\id{N\_VDiv}       & \cm &     &     & \cm & \cm &     &     &     \\ \hline
\id{N\_VScale}     & \cm & \cm & \cm & \cm & \cm & \cm & \cm & \cm \\ \hline
\id{N\_VAbs}       & \cm &     &     &     &     &     &     &     \\ \hline
\id{N\_VInv}       & \cm &     &     & \cm &     &     &     &     \\ \hline
\id{N\_VAddConst}  & \cm &     &     & \cm &     &     &     &     \\ \hline
\id{N\_VDotProd}   &     &     &     &     & \cm &     &     &     \\ \hline
\id{N\_VMaxNorm}   & \cm &     &     &     &     &     &     &     \\ \hline
\id{N\_VWrmsNorm}  & \cm & \cm & \cm &     & \cm & \cm & \cm &     \\ \hline
\id{N\_VMin}       & \cm &     &     &     &     &     &     &     \\ \hline
\id{N\_VCompare}   &     &     &     & \cm &     &     &     &     \\ \hline
\id{N\_VInvTest}   &     &     &     & \cm &     &     &     &     \\ \hline
\end{tabular}
\end{table}


The vector kernels listed in Table \ref{t:nvecops} that are {\em not} used by
{\cvodes} are: \id{N\_VWL2Norm}, \id{N\_VL1Norm}, \id{N\_VWrmsNormMask},
\id{N\_VConstrMask}, and \id{N\_VMinQuotient}. Therefore a user-supplied {\nvector} module
for {\cvodes} could omit these five kernels. Also, recall that a user-supplied {\nvector}
module {\em need not} implement the \id{N\_VNew\_S} and \id{N\_VFree\_S} kernels.