\addcontentsline{toc}{section}{References}

\begin{thebibliography}{99}

\bibitem{BBH89}
{P.~N. Brown, G.~D. Byrne, and A.~C. Hindmarsh}, {\em {VODE}, a
Variable-Coefficient {ODE} Solver}, SIAM J. Sci. Stat. Comput., 10 (1989),
pp.~1038--1051.

\bibitem{BrHi89}
P.~N. Brown and A.~C. Hindmarsh, {\em Reduced Storage Matrix Methods
in Stiff ODE Systems}, J. Appl. Math. \& Comp. 31 (1989), pp. 40--91.

\bibitem{By92}
G.~D. Byrne, {\em Pragmatic Experiments with Krylov Methods in
the Stiff ODE Setting}, in {\em Computational Ordinary Differential
Equations}, J. R. Cash and I. Gladwell (Eds.), Oxford University
Press, Oxford, 1992, pp. 323--356.

%\bibitem{ByH98}
%G.~D. Byrne and A.~C. Hindmarsh, {\em User Documentation for PVODE,
%An ODE Solver for Parallel Computers}, LLNL Report UCRL-ID-130884, May 1998

\bibitem{ByH99}
G.~D. Byrne and A.~C. Hindmarsh, {\em PVODE, An ODE Solver for Parallel Computers}, 
Intl. J. High Perf. Comput. Apps., 1999, 13(4), pp. 254--365.

\bibitem{CLPS02}
Y. Cao, S. Li, L.~R. Petzold, and R. Serban,
{\em Adjoint Sensitivity Analysis for Differential-Algebraic
Equations: The Adjoint DAE and its Numerical Solution},
SIAM J. Sci. Comp., to appear.

\bibitem{CS85}
M. Carcotsios and W.~E. Stewart, {\em Sensitivity Analysis of Initial Value
Problems with Mixed ODEs and Algebraic Constraints}, Comput. Chem. Eng., 1985,
9, pp. 359--365.

\bibitem{CH94}
S.~D. Cohen and A.~C. Hindmarsh, {\em CVODE User Guide},
LLNL Report UCRL-MA-118618, Sept. 1994

\bibitem{CH96}
S.~D. Cohen and A.~C. Hindmarsh, {\em CVODE, a Stiff/Nonstiff ODE Solver in C},
Computers in Physics. 10(2) (1996), pp. 138-143.

\bibitem{FTB97}
W.~F. Feehery, J.~E. Tolsma, and P.~I. Barton, {\em Efficient Sensitivity
Analysis of Large-Scale Differential-Algebraic Systems}, Appl. Num. Math., 1997, 25, 
pp. 41--54.

\bibitem{Hi92}
A.~C. Hindmarsh, {\em Detecting Stability Barriers in BDF Solvers},
in Computational Ordinary Differential Equations, J. R. Cash and
I. Gladwell (Eds.), Oxford Univ. Press, 1992, pp. 87-96.  Also
available as LLNL Report UCRL-101197, June 1989.

\bibitem{Hi95}
A.~C. Hindmarsh, {\em Avoiding BDF Stability Barriers in the MOL
Solution of Advection-Dominated Problems},
Appl. Num. Math., 1995, 17, pp. 311-318.

\bibitem{HT98}
A.~C. Hindmarsh and A.~G. Taylor, {\em PVODE and KINSOL: Parallel Software
for Differential and Nonlinear Systems}, LLNL Report UCRL-ID-129739, February 1998.

\bibitem{MP96}
T. Maly and L.~R. Petzold, {\em Numerical Methods and Software for Sensitivity
Analysis of Differential-Algebraic Systems}, Appl. Numer. Math., 1996, 20, pp. 57--79.

\bibitem{KH93}
K. Radhakrishnan and A.~C. Hindmarsh, {\em Description and Use
of LSODE, the Livermore Solver for Ordinary Differential Equations},
NASA Reference Publication 1327, 1993, and LLNL Report UCRL-ID-113855, March
1994.

\bibitem{SaSc86}
Y. Saad and M.~H. Schultz, {\em GMRES: A Generalized Minimal Residual
Algorithm for Solving Nonsymmetric Linear Systems}, SIAM J. Sci. Stat.
Comp. 7 (1986), pp. 856--869.

\end{thebibliography}
