\textit{The Fortran 2003 interface modules were generated with SWIG Fortran, a
fork of SWIG \cite{Swig-Fortran}. Users who are interested in the SWIG code used
in the generation process should contact the {\sundials} development team.}

\subsection{SUNDIALS Fortran 2003 Interface Modules}

All of the generic {\sundials} modules provide Fortran 2003 interface modules.
Many of the generic module implementations provide Fortran 2003 interfaces
(a complete list of modules with Fortran 2003 interfaces is given in
Table~\ref{t:f2003interface}). A module can be accessed with the \id{use}
statement, e.g. \id{use fnvector\_openmp\_mod}, and linking to the Fortran
2003 library in addition to the {\CC} library, e.g.
\id{libsundials\_fnvecpenmp\_mod}.{\em lib} and
\id{libsundials\_nvecopenmp}.{\em lib}.

The Fortran 2003 interfaces leverage the \id{iso\_c\_binding} module and the
\id{bind(C)} attribute to closely follow the {\sundials} {\CC} API (ignoring
language differences). The generic {\sundials} structures, e.g. \id{N\_Vector},
are interfaced as Fortran derived types, and function signatures are matched
but with an \id{F} prepending the name, e.g. \id{FN\_VConst} instead of
\id{N\_VConst}. Constants are named exactly as they are in the {\CC} API.
Accordingly, using {\sundials} via the Fortran 2003 interfaces looks just like
using it in {\CC}. Some caveats stemming from the language differences are
discussed in the section \ref{ss:fortran2003_differences}. A discussion on the
topic of equivalent data types in {\CC} and Fortran 2003 is presented in
section \ref{ss:fortran2003_data_types}.

Further information on the Fortran 2003 interfaces specific to modules is given
in the {\nvector}, {\sunmatrix}, {\sunlinsol}, and {\sunnonlinsol} alongside
the {\CC} documentation (chapters \ref{s:nvector}, \ref{s:sunmatrix},
\ref{s:sunlinsol}, and \ref{c:sunnonlinsol} respectively). For details on where
the Fortran 2003 module (\id{.mod}) files and libraries are installed see Appendix
\ref{c:install}.

\begin{table}[!htb]
\centering
\caption{Summary of Fortran 2003 interfaces for shared {\sundials} modules.}
\label{t:f2003interface}
\medskip
\begin{tabular}{|l|l|c|}
\hline
{\bf Module}                & {\bf Fortran 2003 Module Name}       \\
\hline
  NVECTOR                   & \id{fsundials\_nvector\_mod}         \\
  {\nvecs}                  & \id{fnvector\_serial\_mod}           \\
  {\nvecp}                  & \id{fnvector\_parallel\_mod}         \\
  {\nvecopenmp}             & \id{fnvector\_openmp\_mod}           \\
  {\nvecpthreads}           & \id{fnvector\_pthreads\_mod}         \\
  {\nvecph}                 & Not interfaced                       \\
  {\nvecpetsc}              & Not interfaced                       \\
  {\nveccuda}               & Not interfaced                       \\
  {\nvecraja}               & Not interfaced                       \\
  {\nvecmanyvector}         & \id{fnvector\_manyvector\_mod}       \\
  {\nvecmpimanyvector}      & \id{fnvector\_mpimanyvector\_mod}    \\
  {\nvecmpiplusx}           & \id{fnvector\_mpiplusx\_mod}         \\
  SUNMatrix                 & \id{fsundials\_matrix\_mod}          \\
  {\sunmatband}             & \id{fsunmatrix\_band\_mod}           \\
  {\sunmatdense}            & \id{fsunmatrix\_dense\_mod}          \\
  {\sunmatsparse}           & \id{fsunmatrix\_sparse\_mod}         \\
  SUNLinearSolver           & \id{fsundials\_linearsolver\_mod}    \\
  {\sunlinsolband}          & \id{fsunlinsol\_band\_mod}           \\
  {\sunlinsoldense}         & \id{fsunlinsol\_dense\_mod}          \\
  {\sunlinsollapband}       & Not interfaced                       \\
  {\sunlinsollapdense}      & Not interfaced                       \\
  {\sunlinsolklu}           & \id{fsunlinsol\_klu\_mod}            \\
  {\sunlinsolslumt}         & Not interfaced                       \\
  {\sunlinsolsludist}       & Not interfaced                       \\
  {\sunlinsolspgmr}         & \id{fsunlinsol\_spgmr\_mod}          \\
  {\sunlinsolspfgmr}        & \id{fsunlinsol\_spfgmr\_mod}         \\
  {\sunlinsolspbcgs}        & \id{fsunlinsol\_spbcgs\_mod}         \\
  {\sunlinsolsptfqmr}       & \id{fsunlinsol\_sptfqmr\_mod}        \\
  {\sunlinsolpcg}           & \id{fsunlinsol\_pcg\_mod}            \\
  SUNNonlinearSolver        & \id{fsundials\_nonlinearsolver\_mod} \\
  {\sunnonlinsolnewton}     & \id{fsunnonlinsol\_newton\_mod}      \\
  {\sunnonlinsolfixedpoint} & \id{fsunnonlinsol\_fixedpoint\_mod}  \\
\hline
\end{tabular}
\end{table}


\subsection{Data Types}\label{ss:fortran2003_data_types}
Generally, the Fortran 2003 type that is equivalent to the {\CC} type is what one
would expect. Primitive types map to the \id{iso\_c\_binding} type equivalent.
{\sundials} generic types map to a Fortran derived type. However, the handling
of pointer types is not always clear as they can depend on the parameter direction.
Table \ref{t:fortran2003_data_types} presents a summary of the type equivalencies
with the parameter direction in mind.
\\
\\
\noindent{\warn}Currently, the Fortran 2003 interfaces are only compatible with
{\sundials} builds where the \id{realtype} is double precision and the \id{sunindextype}
size is 64-bits.

\begin{table}[!htb]
\centering
\caption{C/Fortran 2003 Equivalent Types}
\label{t:fortran2003_data_types}
\medskip
\begin{tabular}{|l|l|l|}
\hline
{\bf {\CC} type}         & {\bf Parameter Direction} & {\bf Fortran 2003 type}                      \\
\hline
\id{double}              & in, inout, out, return    & \id{real(c\_double)}                         \\
\id{int}                 & in, inout, out, return    & \id{integer(c\_int)}                         \\
\id{long}                & in, inout, out, return    & \id{integer(c\_long)}                        \\
\id{booleantype}         & in, inout, out, return    & \id{integer(c\_int)}                         \\
\id{realtype}            & in, inout, out, return    & \id{real(c\_double)}                         \\
\id{sunindextype}        & in, inout, out, return    & \id{integer(c\_long)}                        \\
\id{double*}             & in, inout, out            & \id{real(c\_double), dimension(*)}           \\
\id{double*}             & return                    & \id{real(c\_double), pointer, dimension(:)}  \\
\id{int*}                & in, inout, out            & \id{integer(c\_int), dimension(*)}           \\
\id{int*}                & return                    & \id{integer(c\_int), pointer, dimension(:)}  \\
\id{long*}               & in, inout, out            & \id{integer(c\_long), dimension(*)}          \\
\id{long*}               & return                    & \id{integer(c\_long), pointer, dimension(:)} \\
\id{realtype*}           & in, inout, out            & \id{real(c\_double), dimension(*)}           \\
\id{realtype*}           & return                    & \id{real(c\_double), pointer, dimension(:)}  \\
\id{sunindextype*}       & in, inout, out            & \id{integer(c\_long), dimension(*)}          \\
\id{sunindextype*}       & return                    & \id{integer(c\_long), pointer, dimension(:)} \\
\id{realtype[]}          & in, inout, out            & \id{real(c\_double), dimension(*)}           \\
\id{sunindextype[]}      & in, inout, out            & \id{integer(c\_long), dimension(*)}          \\
\id{N\_Vector}           & in, inout, out            & \id{type(N\_Vector)}                         \\
\id{N\_Vector}           & return                    & \id{type(N\_Vector), pointer}                \\
\id{SUNMatrix}           & in, inout, out            & \id{type(SUNMatrix)}                         \\
\id{SUNMatrix}           & return                    & \id{type(SUNMatrix), pointer}                \\
\id{SUNLinearSolver}     & in, inout, out            & \id{type(SUNLinearSolver)}                   \\
\id{SUNLinearSolver}     & return                    & \id{type(SUNLinearSolver), pointer}          \\
\id{SUNNonlinearSolver}  & in, inout, out            & \id{type(SUNNonlinearSolver)}                \\
\id{SUNNonlinearSolver}  & return                    & \id{type(SUNNonlinearSolver), pointer}       \\
\id{FILE*}               & in, inout, out, return    & \id{type(c\_ptr)}                            \\
\id{void*}               & in, inout, out, return    & \id{type(c\_ptr)}                            \\
\id{T**}                 & in, inout, out, return    & \id{type(c\_ptr)}                            \\
\id{T***}                & in, inout, out, return    & \id{type(c\_ptr)}                            \\
\id{T****}               & in, inout, out, return    & \id{type(c\_ptr)}                            \\
\hline
\end{tabular}
\end{table}


\subsection{Notable Fortran/C usage differences}\label{ss:fortran2003_differences}

While the Fortran 2003 interface to {\sundials} closely follows the {\CC} API,
some differences are inevitable due to the differences between Fortran and {\CC}.
In this section, we note the most critical differences. Additionally, section
\ref{ss:fortran2003_data_types} discusses equivalencies of data types in the
two languages.

\subsubsection{Creating generic {\sundials} objects}

In the {\CC} API a generic {\sundials} object, such as an \id{N\_Vector}, is actually
a pointer to an underlying {\CC} struct. However, in the Fortran 2003 interface,
the derived type is bound to the {\CC} struct, not the pointer to the struct. E.g.,
\id{type(N\_Vector)} is bound to the {\CC} struct \id{\_generic\_N\_Vector} not the
\id{N\_Vector} type. The consequence of this is that creating and declaring {\sundials}
objects in Fortran is nuanced. This is illustrated in the code snippets below:
\\
\\
\noindent \emph{{\CC} code}:
\begin{verbatim}
N_Vector x;
x = N_VNew_Serial(N);
\end{verbatim}

\noindent \emph{Fortran code}:
\begin{verbatim}
type(N_Vector), pointer :: x
x => FN_VNew_Serial(N)
\end{verbatim}

Note that in the Fortran declaration, the vector is a \id{type(N\_Vector), pointer}, and
that the pointer assignment operator is then used.

\subsubsection{Arrays and pointers}

Unlike in the {\CC} API, in the Fortran 2003 interface, arrays and pointers are
treated differently when they are return values versus arguments to a function.
Additionally, pointers which are meant to be out parameters, not arrays,
in the {\CC} API must still be declared as a rank-1 array in Fortran.
The reason for this is partially due to the Fortran 2003 standard for C bindings,
and partially due to the tool used to generate the interfaces. Regardless, the
code snippets below illustrate the differences.
\\
\\
\noindent \emph{{\CC} code}:
\begin{verbatim}
N_Vector x
realtype* xdata;
long int leniw, lenrw;

x = N_VNew_Serial(N);

/* capturing a returned array/pointer */
xdata = N_VGetArrayPointer(x)

/* passing array/pointer to a function */
N_VSetArrayPointer(xdata, x)

/* pointers that are out-parameters */
N_VSpace(x, &leniw, &lenrw);
\end{verbatim}

\noindent \emph{Fortran code}:
\begin{verbatim}
type(N_Vector), pointer :: x
real(c_double), pointer :: xdataptr(:)
real(c_double)          :: xdata(N)
integer(c_long)         :: leniw(1), lenrw(1)

x => FN_VNew_Serial(x)

! capturing a returned array/pointer
xdataptr => FN_VGetArrayPointer(x)

! passing array/pointer to a function
call FN_VSetArrayPointer(xdata, x)

! pointers that are out-parameters
call FN_VSpace(x, leniw, lenrw)
\end{verbatim}

\subsubsection{Passing procedure pointers and user data}

Since functions/subroutines passed to {\sundials} will be called from within
{\CC} code, the Fortran procedure must have the attribute \id{bind(C)}.
Additionally, when providing them as arguments to a Fortran 2003 interface
routine, it is required to convert a procedure's Fortran address to C with
the Fortran intrinsic \id{c\_funloc}.

Typically when passing user data to a {\sundials} function, a user may
simply cast some custom data structure as a \id{void*}. When using the
Fortran 2003 interfaces, the same thing can be achieved. Note, the
custom data structure \emph{does not} have to be \id{bind(C)} since
it is never accessed on the {\CC} side.
\\
\\
\noindent \emph{{\CC} code}:
\begin{verbatim}
MyUserData* udata;
void *cvode_mem;

ierr = CVodeSetUserData(cvode_mem, udata);
\end{verbatim}

\noindent \emph{Fortran code}:
\begin{verbatim}
type(MyUserData) :: udata
type(c_ptr)      :: cvode_mem

ierr = FCVodeSetUserData(cvode_mem, c_loc(udata))
\end{verbatim}

On the other hand, Fortran users may instead choose to store problem-specific data,
e.g. problem parameters, within modules, and thus do not need the SUNDIALS-provided
\id{user\_data} pointers to pass such data back to user-supplied functions. These users
should supply the \id{c\_null\_ptr} input for \id{user\_data} arguments to the relevant
{\sundials} functions.

\subsubsection{Passing NULL to optional parameters}

In the {\sundials} {\CC} API some functions have optional parameters that a
caller can pass \id{NULL} to. If the optional parameter is of a type that is
equivalent to a Fortran \id{type(c\_ptr)} (see section \ref{ss:fortran2003_data_types}),
then a Fortran user can pass the intrinsic \id{c\_null\_ptr}. However, if the
optional parameter is of a type that is not equivalent to \id{type(c\_ptr)},
then a caller must provide a Fortran pointer that is dissociated. This
is demonstrated in the code example below.
\\
\\
\noindent \emph{{\CC} code}:
\begin{verbatim}
SUNLinearSolver LS;
N_Vector x, b;

! SUNLinSolSolve expects a SUNMatrix or NULL
! as the second parameter.
ierr = SUNLinSolSolve(LS, NULL, x, b);
\end{verbatim}

\noindent \emph{Fortran code}:
\begin{verbatim}
type(SUNLinearSolver), pointer :: LS
type(SUNMatrix), pointer :: A
type(N_Vector), pointer :: x, b

A => null()

! SUNLinSolSolve expects a type(SUNMatrix), pointer
! as the second parameter. Therefore, we cannot
! pass a c_null_ptr, rather we pass a disassociated A.
ierr = FSUNLinSolSolve(LS, A, x, b)
\end{verbatim}

\subsubsection{Working with \id{N\_Vector} arrays}
\label{sss:fortran2003_nvarrays}

Arrays of \id{N\_Vector} objects are interfaced to Fortran 2003 as opaque
\id{type(c\_ptr)}.  As such, it is not possible to directly index an array of
\id{N\_Vector} objects returned by the \id{N\_Vector} ``VectorArray''
operations, or packages with sensitivity capablities.  Instead, {\sundials}
provides a utility function \id{FN\_VGetVecAtIndexVectorArray} that can be called for
accessing a vector in a vector array. The example below demonstrates this:
\\
\\
\noindent \emph{{\CC} code}:
\begin{verbatim}
N_Vector x;
N_Vector* vecs;

vecs = N_VCloneVectorArray(count, x);
for (int i=0; i < count; ++i)
  N_VConst(vecs[i]);
\end{verbatim}

\noindent \emph{Fortran code}:
\begin{verbatim}
type(N_Vector), pointer :: x, xi
type(c_ptr)             :: vecs

vecs = FN_VCloneVectorArray(count, x)
do index, count
  xi => FN_VGetVecAtIndexVectorArray(vecs, index)
  call FN_VConst(xi)
enddo
\end{verbatim}

{\sundials} also provides the functions \id{FN\_VSetVecAtIndexVectorArray} and
\id{FN\_VNewVectorArray} for working with \id{N\_Vector} arrays. These functions
are particularly useful for users of the Fortran interface to the
{\nvecmanyvector} or {\nvecmpimanyvector} when creating the subvector array.
Both of these functions along with \id{FN\_VGetVecAtIndexVectorArray} are further
described in Chapter \ref{ss:nvecutils}.

\subsubsection{Providing file pointers}

Expert {\sundials} users may notice that there are a few advanced functions in the {\sundials}
{\CC} API that take a \id{FILE *} argument. Since there is no portable way to convert between
a Fortran file descriptor and a {\CC} file pointer, {\sundials} provides two utility functions
for creating a \id{FILE *} and destroying it. These functions are defined in the module
\id{fsundials\_futils\_mod}.
\\
\\
\ucfunction{FSUNDIALSFileOpen}
{
  fp = FSUNDIALSFileOpen(filename, mode)
}
{
  The function allocates a \id{FILE *} by calling the C function \id{fopen}.
}
{
  \begin{args}[filename]
  \item[filename] (\id{character(kind=C\_CHAR, len=*)}) - the path to the file to open
  \item[mode] (\id{character(kind=C\_CHAR, len=*)}) - the mode string given to \id{fopen}
    It should begin with one of the following characters: 
    \begin{itemize}
      \item[] ``r'' - open text file for reading
      \item[] ``r+'' - open text file for reading and writing
      \item[] ``w''  - truncate text file to zero length or create it for writing
      \item[] ``w+'' - open text file for reading or writing, create it if it does not exist
      \item[] ``a''  - open for appending, see documentation of ``fopen`` for your system/compiler
      \item[] ``a+'' - open for reading and appending, see documentation for ``fopen`` for your system/compiler
    \end{itemize}
  \end{args}
}
{
  This returns a \id{type(C\_PTR)} which is a \id{FILE*} in C.
  If it is \id{NULL}, then there was an error opening the file.
}
{}
%
\ucfunction{FSUNDIALSFileClose}
{
  call FSUNDIALSFileClose(fp)
}
{
  The function deallocates a \id{FILE*} by calling the C function \id{fclose}.
}
{
  \begin{args}[fp]
  \item[fp] (\id{type(C\_PTR)}) - the file pointer (type \id{FILE*} in C)
  \end{args}
}
{}
{}

% =============================================================================
\subsection{Important notes on portability}\label{ss:f2003_portability}

The {\sundials} Fortran 2003 interface \textit{should} be compatible with any compiler
supporting the Fortran 2003 ISO standard. However, it has only been tested and confirmed
to be working with GNU Fortran 4.9+ and Intel Fortran 18.0.1+.

Upon compilation of {\sundials}, Fortran module (\id{.mod}) files are generated
for each Fortran 2003 interface. These files are highly compiler specific, and
thus it is almost always necessary to compile a consuming application with the
same compiler used to generate the modules.
