\section{NVECTOR Examples}\label{ss:nvec_examples}

There are \id{NVector} examples that may be installed for the
implementations provided with {\sundials}. Each
implementation makes use of the functions in \id{test\_nvector.c}.
These example functions show simple usage of the \id{NVector} family
of functions. The input to the examples are the vector length, number
of threads (if threaded implementation), and a print timing flag.

\noindent The following is a list of the example functions in \id{test\_nvector.c}:
\begin{itemize}
\item \id{Test\_N\_VClone}: Creates clone of vector and checks validity of clone.  
\item \id{Test\_N\_VCloneEmpty}: Creates clone of empty vector and checks validity of clone.  
\item \id{Test\_N\_VCloneVectorArray}: Creates clone of vector array and checks validity of cloned array.  
\item \id{Test\_N\_VCloneVectorArray}: Creates clone of empty vector array and checks validity of cloned array.  
\item \id{Test\_N\_VGetArrayPointer}: Get array pointer. 
\item \id{Test\_N\_VSetArrayPointer}: Allocate new vector, set pointer to new vector array, and check values. 
\item \id{Test\_N\_VLinearSum} Case 1a: Test y =  x + y 
\item \id{Test\_N\_VLinearSum} Case 1b: Test y = -x + y 
\item \id{Test\_N\_VLinearSum} Case 1c: Test y = ax + y
\item \id{Test\_N\_VLinearSum} Case 2a: Test x =  x + y
\item \id{Test\_N\_VLinearSum} Case 2b: Test x =  x - y
\item \id{Test\_N\_VLinearSum} Case 2c: Test x =  x + by
\item \id{Test\_N\_VLinearSum} Case 3:  Test z =  x + y
\item \id{Test\_N\_VLinearSum} Case 4a: Test z =  x - y
\item \id{Test\_N\_VLinearSum} Case 4b: Test z = -x + y
\item \id{Test\_N\_VLinearSum} Case 5a: Test z =  x + by
\item \id{Test\_N\_VLinearSum} Case 5b: Test z = ax + y
\item \id{Test\_N\_VLinearSum} Case 6a: Test z = -x + by
\item \id{Test\_N\_VLinearSum} Case 6b: Test z = ax - y
\item \id{Test\_N\_VLinearSum} Case 7:  Test z = a(x + y)
\item \id{Test\_N\_VLinearSum} Case 8:  Test z = a(x - y)
\item \id{Test\_N\_VLinearSum} Case 9:  Test z = ax + by
\item \id{Test\_N\_VConst}: Fill vector with constant and check result.
\item \id{Test\_N\_VProd}: Test vector multiply: z = x * y
\item \id{Test\_N\_VDiv}: Test vector division: z = x / y
\item \id{Test\_N\_VScale}: Case 1: scale: x = cx
\item \id{Test\_N\_VScale}: Case 2: copy: z = x
\item \id{Test\_N\_VScale}: Case 3: negate: z = -x
\item \id{Test\_N\_VScale}: Case 4: combination: z = cx
\item \id{Test\_N\_VAbs}: Create absolute value of vector. 
\item \id{Test\_N\_VAddConst}: add constant vector: z = c + x
\item \id{Test\_N\_VDotProd}: Calculate dot product of two vectors.
\item \id{Test\_N\_VMaxNorm}: Create vector with known values, find and validate the max norm.
\item \id{Test\_N\_VWrmsNorm}: Create vector of known values, find and validate the weighted root mean square.
\item \id{Test\_N\_VWrmsNormMask}: Create vector of known values, find and validate the weighted root mean square using all elements except one.
\item \id{Test\_N\_VMin}: Create vector, find and validate the min.
\item \id{Test\_N\_VWL2Norm}: Create vector, find and validate the weighted Euclidean L2 norm.
\item \id{Test\_N\_VL1Norm}: Create vector, find and validate the L1 norm.
\item \id{Test\_N\_VCompare}: Compare vector with constant returning and validating comparison vector.
\item \id{Test\_N\_VInvTest}: Test z[i] = 1 / x[i]
\item \id{Test\_N\_VConstrMask}: Test mask of vector x with vector c.
\item \id{Test\_N\_VMinQuotient}: Fill two vectors with known values. Calculate and validate minimum quotient.
\item \id{Test\_N\_VLinearCombination} Case 1a: Test x = a x
\item \id{Test\_N\_VLinearCombination} Case 1b: Test z = a x
\item \id{Test\_N\_VLinearCombination} Case 2a: Test x = a x + b y
\item \id{Test\_N\_VLinearCombination} Case 2b: Test z = a x + b y
\item \id{Test\_N\_VLinearCombination} Case 3a: Test x = x + a y + b z
\item \id{Test\_N\_VLinearCombination} Case 3b: Test x = a x + b y + c z
\item \id{Test\_N\_VLinearCombination} Case 3c: Test w = a x + b y + c z
\item \id{Test\_N\_VScaleAddMulti} Case 1a: y = a x + y
\item \id{Test\_N\_VScaleAddMulti} Case 1b: z = a x + y
\item \id{Test\_N\_VScaleAddMulti} Case 2a: Y[i] = c[i] x + Y[i], i = 1,2,3
\item \id{Test\_N\_VScaleAddMulti} Case 2b: Z[i] = c[i] x + Y[i], i = 1,2,3
\item \id{Test\_N\_VDotProdMulti} Case 1: Calculate the dot product of two vectors
\item \id{Test\_N\_VDotProdMulti} Case 2: Calculate the dot product of one vector with three other vectors in a vector array.
\item \id{Test\_N\_VLinearSumVectorArray} Case 1: z = a x + b y 
\item \id{Test\_N\_VLinearSumVectorArray} Case 2a: Z[i] = a X[i] + b Y[i]
\item \id{Test\_N\_VLinearSumVectorArray} Case 2b: X[i] = a X[i] + b Y[i]
\item \id{Test\_N\_VLinearSumVectorArray} Case 2c: Y[i] = a X[i] + b Y[i]
\item \id{Test\_N\_VScaleVectorArray} Case 1a: y = c y
\item \id{Test\_N\_VScaleVectorArray} Case 1b: z = c y
\item \id{Test\_N\_VScaleVectorArray} Case 2a: Y[i] = c[i] Y[i]
\item \id{Test\_N\_VScaleVectorArray} Case 2b: Z[i] = c[i] Y[i]
\item \id{Test\_N\_VScaleVectorArray} Case 1a: z = c
\item \id{Test\_N\_VScaleVectorArray} Case 1b: Z[i] = c
\item \id{Test\_N\_VWrmsNormVectorArray} Case 1a: Create a vector of know values, find and validate the weighted root mean square norm.
\item \id{Test\_N\_VWrmsNormVectorArray} Case 1b: Create a vector array of three vectors of know values, find and validate the weighted root mean square norm of each.
\item \id{Test\_N\_VWrmsNormMaskVectorArray} Case 1a: Create a vector of know values, find and validate the weighted root mean square norm using all elements except one.
\item \id{Test\_N\_VWrmsNormMaskVectorArray} Case 1b: Create a vector array of three vectors of know values, find and validate the weighted root mean square norm of each using all elements except one.
\item \id{Test\_N\_VScaleAddMultiVectorArray} Case 1a: y = a x + y 
\item \id{Test\_N\_VScaleAddMultiVectorArray} Case 1b: z = a x + y
\item \id{Test\_N\_VScaleAddMultiVectorArray} Case 2a: Y[j][0] = a[j] X[0] + Y[j][0]
\item \id{Test\_N\_VScaleAddMultiVectorArray} Case 2b: Z[j][0] = a[j] X[0] + Y[j][0]
\item \id{Test\_N\_VScaleAddMultiVectorArray} Case 3a: Y[0][i] = a[0] X[i] + Y[0][i]
\item \id{Test\_N\_VScaleAddMultiVectorArray} Case 3b: Z[0][i] = a[0] X[i] + Y[0][i]
\item \id{Test\_N\_VScaleAddMultiVectorArray} Case 4a: Y[j][i] = a[j] X[i] + Y[j][i]
\item \id{Test\_N\_VScaleAddMultiVectorArray} Case 4b: Z[j][i] = a[j] X[i] + Y[j][i]  
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 1a: x = a x
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 1b: z = a x
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 2a: x = a x + b y 
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 2b: z = a x + b y
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 3a: x = a x + b y + c z
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 3b: w = a x + b y + c z
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 4a: X[0][i] = c[0] X[0][i]
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 4b: Z[i] = c[0] X[0][i]
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 5a: X[0][i] = c[0] X[0][i] + c[1] X[1][i]
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 5b: Z[i] = c[0] X[0][i] + c[1] X[1][i]
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 6a: X[0][i] = X[0][i] + c[1] X[1][i] + c[2] X[2][i]
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 6b: X[0][i] = c[0] X[0][i] + c[1] X[1][i] + c[2] X[2][i]
\item \id{Test\_N\_VLinearCombinationVectorArray} Case 6c: Z[i] = c[0] X[0][i] + c[1] X[1][i] + c[2] X[2][i]
\end{itemize}
