\\\hline
\multicolumn{3}{c}{\bf {\spgmr} generic linear solver module}\\
\hline\\

\id{SPGMR\_SUCCESS}             &  0 & Converged. \\
\id{SPGMR\_RES\_REDUCED}        &  1 & No convergence, but the residual norm was reduced. \\
\id{SPGMR\_CONV\_FAIL}          &  2 & Failure to converge. \\
\id{SPGMR\_QRFACT\_FAIL}        &  3 & A singular matrix was found during the QR factorization. \\
\id{SPGMR\_PSOLVE\_FAIL\_REC}   &  4 & The preconditioner solve function failed recoverably.\\
\id{SPGMR\_ATIMES\_FAIL\_REC}   &  5 & The Jacobian-times-vector function failed recoverably.\\
\id{SPGMR\_PSET\_FAIL\_REC}     &  6 & The preconditioner setup routine failed recoverably.\\
\id{SPGMR\_MEM\_NULL}           & -1 & The {\spgmr} memory is \id{NULL}\\
\id{SPGMR\_ATIMES\_FAIL\_UNREC} & -2 & The Jacobian-times-vector function failed unrecoverably. \\
\id{SPGMR\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPGMR\_GS\_FAIL}            & -4 & Failure in the Gram-Schmidt procedure. \\
\id{SPGMR\_QRSOL\_FAIL}         & -5 & The matrix $R$ was found to be singular during the QR solve phase. \\
\id{SPGMR\_PSET\_FAIL\_UNREC}   & -6 & The preconditioner setup routine failed unrecoverably.\\

\\\hline
\multicolumn{3}{c}{\bf {\spfgmr} generic linear solver module (only available in {\kinsol} and {\arkode})}\\
\hline\\

\id{SPFGMR\_SUCCESS}             &  0 & Converged. \\
\id{SPFGMR\_RES\_REDUCED}        &  1 & No convergence, but the residual norm was reduced. \\
\id{SPFGMR\_CONV\_FAIL}          &  2 & Failure to converge. \\
\id{SPFGMR\_QRFACT\_FAIL}        &  3 & A singular matrix was found during the QR factorization. \\
\id{SPFGMR\_PSOLVE\_FAIL\_REC}   &  4 & The preconditioner solve function failed recoverably.\\
\id{SPFGMR\_ATIMES\_FAIL\_REC}   &  5 & The Jacobian-times-vector function failed recoverably.\\
\id{SPFGMR\_PSET\_FAIL\_REC}     &  6 & The preconditioner setup routine failed recoverably.\\
\id{SPFGMR\_MEM\_NULL}           & -1 & The {\spfgmr} memory is \id{NULL}\\
\id{SPFGMR\_ATIMES\_FAIL\_UNREC} & -2 & The Jacobian-times-vector function failed unrecoverably. \\
\id{SPFGMR\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPFGMR\_GS\_FAIL}            & -4 & Failure in the Gram-Schmidt procedure. \\
\id{SPFGMR\_QRSOL\_FAIL}         & -5 & The matrix $R$ was found to be singular during the QR solve phase. \\
\id{SPFGMR\_PSET\_FAIL\_UNREC}   & -6 & The preconditioner setup routine failed unrecoverably.\\

\\\hline
\multicolumn{3}{c}{\bf {\spbcg} generic linear solver module}\\
\hline\\

\id{SPBCG\_SUCCESS}             &  0 & Converged. \\
\id{SPBCG\_RES\_REDUCED}        &  1 & No convergence, but the residual norm was reduced. \\
\id{SPBCG\_CONV\_FAIL}          &  2 & Failure to converge. \\
\id{SPBCG\_PSOLVE\_FAIL\_REC}   &  3 & The preconditioner solve function failed recoverably.\\
\id{SPBCG\_ATIMES\_FAIL\_REC}   &  4 & The Jacobian-times-vector function failed recoverably.\\
\id{SPBCG\_PSET\_FAIL\_REC}     &  5 & The preconditioner setup routine failed recoverably.\\
\id{SPBCG\_MEM\_NULL}           & -1 & The {\spbcg} memory is \id{NULL}\\
\id{SPBCG\_ATIMES\_FAIL\_UNREC} & -2 & The Jacobian-times-vector function failed unrecoverably. \\
\id{SPBCG\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPBCG\_PSET\_FAIL\_UNREC}   & -4 & The preconditioner setup routine failed unrecoverably.\\

\\\hline
\multicolumn{3}{c}{\bf {\sptfqmr} generic linear solver module}\\
\hline\\

\id{SPTFQMR\_SUCCESS}             &  0 & Converged. \\
\id{SPTFQMR\_RES\_REDUCED}        &  1 & No convergence, but the residual norm was reduced. \\
\id{SPTFQMR\_CONV\_FAIL}          &  2 & Failure to converge. \\
\id{SPTFQMR\_PSOLVE\_FAIL\_REC}   &  3 & The preconditioner solve function failed recoverably.\\
\id{SPTFQMR\_ATIMES\_FAIL\_REC}   &  4 & The Jacobian-times-vector function failed recoverably.\\
\id{SPTFQMR\_PSET\_FAIL\_REC}     &  5 & The preconditioner setup routine failed recoverably.\\
\id{SPTFQMR\_MEM\_NULL}           & -1 & The {\sptfqmr} memory is \id{NULL}\\
\id{SPTFQMR\_ATIMES\_FAIL\_UNREC} & -2 & The Jacobian-times-vector function failed. \\
\id{SPTFQMR\_PSOLVE\_FAIL\_UNREC} & -3 & The preconditioner solve function failed unrecoverably. \\
\id{SPTFQMR\_PSET\_FAIL\_UNREC}   & -4 & The preconditioner setup routine failed unrecoverably.\\
