% This is a shared SUNDIALS TEX file with a description of functions
% for getting SUNDIALS version information and is included in each
% solver user guide as the first subsubsection under the ``Optional
% output functions'' subsection. 

%%==============================================================================
\subsubsection{SUNDIALS version information} \label{sss:sunversioninfo}
%%==============================================================================
\index{optional output!version|(}

The following functions provide a way to get {\sundials} version
information at runtime.
%%
\ucfunction{SUNDIALSGetVersion}
{
  flag = SUNDIALSGetVersion(version, len);
}
{
  The function \ID{SUNDIALSGetVersion} fills a character array with {\sundials}
  version information.
}
{
  \begin{args}[version]
  \item[version] (\id{char *}) character array to hold the {\sundials} version information.
  \item[len]     (\id{int})    allocated length of the \id{version} character array.
  \end{args}
}
{
  If successful, \id{SUNDIALSGetVersion} returns 0 and
  \id{version} contains the {\sundials} version
  information. Otherwise, it returns $-1$ and \id{version} is not
  set (the input character array is too short).
}
{
  A string of 25 characters should be sufficient to hold the version
  information. Any trailing characters in the \id{version} array are
  removed.
}
%%
%%
\ucfunction{SUNDIALSGetVersionNumber}
{
  flag = SUNDIALSGetVersionNumber(\&major, \&minor, \&patch, label, len);
}
{
  The function \ID{SUNDIALSGetVersionNumber} set integers for the
  {\sundials} major, minor, and patch release numbers and fills a
  character array with the release label if applicable.
}
{
  \begin{args}[length]
  \item[major] (\id{int})    {\sundials} release major version number.
  \item[minor] (\id{int})    {\sundials} release minor version number.
  \item[patch] (\id{int})    {\sundials} release patch version number.
  \item[label] (\id{char *}) character array to hold the {\sundials} release label.
  \item[len]   (\id{int})    allocated length of the \id{label} character array.
  \end{args}
}
{
  If successful, \id{SUNDIALSGetVersionNumber} returns 0 and the
  \id{major}, \id{minor}, \id{patch}, and \id{label} values are
  set. Otherwise, it returns $-1$ and the values are not set (the
  input character array is too short).
}
{
  A string of 10 characters should be sufficient to hold the label
  information. If a label is not used in the release version, no
  information is copied to \id{label}. Any trailing characters in the
  \id{label} array are removed. 
}
\index{optional output!version|)}

