% ====================================================================
\section{The SUNLinearSolver\_LapackBand implementation}
\label{ss:sunlinsol_lapband}
% ====================================================================

This section describes the {\sunlinsol} implementation for solving banded linear
systems with LAPACK. The {\sunlinsollapband} module is designed to be used with the
corresponding {\sunmatband} matrix type, and one of the serial or
shared-memory {\nvector} implementations ({\nvecs}, {\nvecopenmp}, or
{\nvecpthreads}).

To access the {\sunlinsollapband} module, include the header file \newline
\id{sunlinsol/sunlinsol\_lapackband.h}. The installed module library to link
to is \newline
\id{libsundials\_sunlinsollapackband.\textit{lib}} where \id{\em.lib}
is typically \id{.so} for shared libraries and \id{.a} for static libraries.

The {\sunlinsollapband} module is a {\sunlinsol} wrapper for
the LAPACK band matrix factorization and solve routines, \id{*GBTRF}
and \id{*GBTRS}, where \id{*} is either \id{D} or \id{S}, depending on
whether {\sundials} was configured to have \id{realtype} set to
\id{double} or \id{single}, respectively (see Section \ref{s:types}).
In order to use the {\sunlinsollapband} module it is assumed
that LAPACK has been installed on the system prior to installation of
{\sundials}, and that {\sundials} has been configured appropriately to
link with LAPACK (see Appendix \ref{c:install} for details).
We note that since there do not exist 128-bit floating-point
factorization and solve routines in LAPACK, this interface cannot be
compiled when using \id{extended} precision for \id{realtype}.
Similarly, since there do not exist 64-bit integer LAPACK routines,
the {\sunlinsollapband} module also cannot be compiled when using
64-bit integers for the \id{sunindextype}. {\warn}


% ====================================================================
\subsection{SUNLinearSolver\_LapackBand description}
\label{ss:sunlinsol_lapband_description}
% ====================================================================

This solver is constructed to perform the following operations:
\begin{itemize}
\item The ``setup'' call performs a $LU$ factorization with
  partial (row) pivoting, $PA=LU$, where $P$ is a permutation matrix,
  $L$ is a lower triangular matrix with 1's on the diagonal, and $U$
  is an upper triangular matrix.  This factorization is stored
  in-place on the input {\sunmatband} object $A$, with pivoting
  information encoding $P$ stored in the \id{pivots} array.
\item The ``solve'' call performs pivoting and forward and
  backward substitution using the stored \id{pivots} array and the
  $LU$ factors held in the {\sunmatband} object.
\item
  $A$ must be allocated to accommodate the increase in upper
  bandwidth that occurs during factorization.  More precisely, if $A$
  is a band matrix with upper bandwidth \id{mu} and lower bandwidth
  \id{ml}, then the upper triangular factor $U$ can have upper
  bandwidth as big as \id{smu = MIN(N-1,mu+ml)}. The lower triangular
  factor $L$ has lower bandwidth \id{ml}. {\warn}
\end{itemize}


% ====================================================================
\subsection{SUNLinearSolver\_LapackBand functions}
\label{ss:sunlinsol_lapband_functions}
% ====================================================================

The {\sunlinsollapband} module provides the following user-callable constructor
for creating a \newline \id{SUNLinearSolver} object.
%
% --------------------------------------------------------------------
%
\ucfunctiond{SUNLinSol\_LapackBand}
{
  LS = SUNLinSol\_LapackBand(y, A);
}
{
  The function \ID{SUNLinSol\_LapackBand} creates and allocates memory for
  a LAPACK-based, band \id{SUNLinearSolver} object.
}
{
  \begin{args}[y]
  \item[y] (\id{N\_Vector})
    a template for cloning vectors needed within the solver
  \item[A] (\id{SUNMatrix})
    a {\sunmatband} matrix template for cloning matrices needed
    within the solver
  \end{args}
}
{
  This returns a \id{SUNLinearSolver} object.  If either \id{A} or
  \id{y} are incompatible then this routine will return \id{NULL}.
}
{
  This routine will perform consistency checks to ensure that it is
  called with consistent {\nvector} and {\sunmatrix} implementations.
  These are currently limited to the {\sunmatband} matrix type and
  the {\nvecs}, {\nvecopenmp}, and {\nvecpthreads} vector types.  As
  additional compatible matrix and vector implementations are added to
  {\sundials}, these will be included within this compatibility check.

  Additionally, this routine will verify that the input matrix \id{A}
  is allocated with appropriate upper bandwidth storage for the $LU$
  factorization.
}
{SUNLapackBand}
%
% --------------------------------------------------------------------
%
The {\sunlinsollapband} module defines band implementations of all
``direct'' linear solver operations listed in Sections
\ref{ss:sunlinsol_CoreFn} -- \ref{ss:sunlinsol_GetFn}:
\begin{itemize}
\item \id{SUNLinSolGetType\_LapackBand}
\item \id{SUNLinSolInitialize\_LapackBand} -- this does nothing, since all
  consistency checks are performed at solver creation.
\item \id{SUNLinSolSetup\_LapackBand} -- this calls either
  \id{DGBTRF} or \id{SGBTRF} to perform the $LU$ factorization.
\item \id{SUNLinSolSolve\_LapackBand} -- this calls either
  \id{DGBTRS} or \id{SGBTRS} to use the $LU$ factors and \id{pivots}
  array to perform the solve.
\item \id{SUNLinSolLastFlag\_LapackBand}
\item \id{SUNLinSolSpace\_LapackBand} -- this only returns information for
  the storage \emph{within} the solver object, i.e.~storage
  for \id{N}, \id{last\_flag}, and \id{pivots}.
\item \id{SUNLinSolFree\_LapackBand}
\end{itemize}


% ====================================================================
\subsection{SUNLinearSolver\_LapackBand Fortran interfaces}
\label{ss:sunlinsol_lapband_fortran}
% ====================================================================

For solvers that include a {\F} 77 interface module, the {\sunlinsollapband}
module also includes a Fortran-callable function for creating a
\id{SUNLinearSolver} object.
%
% --------------------------------------------------------------------
%
\ucfunction{FSUNLAPACKDENSEINIT}
{
  FSUNLAPACKBANDINIT(code, ier)
}
{
  The function \ID{FSUNLAPACKBANDINIT} can be called for Fortran programs
  to create a LAPACK-based band \id{SUNLinearSolver} object.
}
{
  \begin{args}[code]
  \item[code] (\id{int*})
    is an integer input specifying the solver id (1 for {\cvode}, 2
    for {\ida}, 3 for {\kinsol}, and 4 for {\arkode}).
  \end{args}
}
{
  \id{ier} is a return completion flag equal to \id{0} for a success
  return and \id{-1} otherwise. See printed message for details in case
  of failure.
}
{
  This routine must be
  called \emph{after} both the {\nvector} and {\sunmatrix} objects have
  been initialized.
}
Additionally, when using {\arkode} with a non-identity
mass matrix, the {\sunlinsollapband} module includes a Fortran-callable
function for creating a \id{SUNLinearSolver} mass matrix solver
object.
%
% --------------------------------------------------------------------
%
\ucfunction{FSUNMASSLAPACKBANDINIT}
{
  FSUNMASSLAPACKBANDINIT(ier)
}
{
  The function \ID{FSUNMASSLAPACKBANDINIT} can be called for Fortran programs
  to create a LAPACK-based, band \id{SUNLinearSolver} object for mass matrix linear
  systems.
}
{}
{
  \id{ier} is a \id{int} return completion flag equal to \id{0} for a success
  return and \id{-1} otherwise. See printed message for details in case
  of failure.
}
{
  This routine must be
  called \emph{after} both the {\nvector} and {\sunmatrix} mass-matrix
  objects have been initialized.
}


% ====================================================================
\subsection{SUNLinearSolver\_LapackBand content}
\label{ss:sunlinsol_lapband_content}
% ====================================================================

The {\sunlinsollapband} module defines the \textit{content} field of a
\id{SUNLinearSolver} as the following structure:
%%
\begin{verbatim}
struct _SUNLinearSolverContent_Band {
  sunindextype N;
  sunindextype *pivots;
  sunindextype last_flag;
};
\end{verbatim}
%%
These entries of the \emph{content} field contain the following
information:
\begin{args}[last\_flag]
  \item[N] - size of the linear system,
  \item[pivots] - index array for partial pivoting in LU factorization,
  \item[last\_flag] - last error return flag from internal function evaluations.
\end{args}
