%% This is a shared SUNDIALS TEX file with a description of the
%% spfgmr sunlinsol implementation
%%

The {\spfgmr} (Scaled, Preconditioned, Flexible, Generalized Minimum
Residual \cite{Saa:93}) implementation of the {\sunlinsol} module
provided with {\sundials}, {\sunlinsolspfgmr}, is an iterative linear
solver that is designed to be compatible with any {\nvector}
implementation (serial, threaded, parallel, and user-supplied) that
supports a minimal subset of operations (\id{N\_VClone},
\id{N\_VDotProd}, \id{N\_VScale}, \id{N\_VLinearSum}, \id{N\_VProd},
\id{N\_VConst}, \id{N\_VDiv}, and \id{N\_VDestroy}).  When using
Classical Gram-Schmidt, the optional function \id{N\_VDotProdMulti}
may be supplied for increased efficiency.  Unlike the other
Krylov iterative linear solvers supplied with {\sundials}, FGMRES is
specifically designed to work with a changing preconditioner
(e.g.~from an iterative method).

%---------------------------------------------------------------------------
\subsection{{\sunlinsolspfgmr} usage}\label{ss:sunlinsol_spfgmr_usage}

The header file to include when using this module
is \id{sunlinsol/sunlinsol\_spfgmr.h}. The {\sunlinsolspfgmr} module
is accessible from all {\sundials} solvers \textit{without}
linking to the \\ \noindent
\id{libsundials\_sunlinsolspfgmr} module library.

The module {\sunlinsolspfgmr} provides the following user-callable routines:
% --------------------------------------------------------------------
\ucfunction{SUNLinSol\_SPFGMR}
{
  LS = SUNLinSol\_SPFGMR(y, pretype, maxl);
}
{
  The function \ID{SUNLinSol\_SPFGMR} creates and allocates memory for
  a {\spfgmr} \id{SUNLinearSolver}.
}
{
  \begin{args}[pretype]
  \item[y] (\id{N\_Vector})
    a template for cloning vectors needed within the solver
  \item[pretype] (\id{int})
    flag indicating the desired type of preconditioning, allowed
    values are:
    \begin{itemize}
    \item \id{PREC\_NONE} (0)
    \item \id{PREC\_LEFT} (1)
    \item \id{PREC\_RIGHT} (2)
    \item \id{PREC\_BOTH} (3)
    \end{itemize}
    Any other integer input will result in the default (no
    preconditioning).
  \item[maxl] (\id{int})
    the number of Krylov basis vectors to use.  values $\le0$ will
    result in the default value (5).
  \end{args}
}
{
  This returns a \id{SUNLinearSolver} object.  If either \id{y} is
  incompatible then this routine will return \id{NULL}.
}
{
  This routine will perform consistency checks to ensure that it is
  called with a consistent {\nvector} implementation (i.e.~that it
  supplies the requisite vector operations).  If \id{y} is
  incompatible, then this routine will return \id{NULL}.

  We note that some {\sundials} solvers are designed to only work with
  left preconditioning ({\ida} and {\idas}). While it is possible to
  use a right-preconditioned {\sunlinsolspfgmr} object for these
  packages, this use mode is not supported and may result in inferior
  performance.
}
% --------------------------------------------------------------------
\ucfunction{SUNLinSol\_SPFGMRSetPrecType}
{
  retval = SUNLinSol\_SPFGMRSetPrecType(LS, pretype);
}
{
  The function \ID{SUNLinSol\_SPFGMRSetPrecType} updates the type of
  preconditioning to use in the {\sunlinsolspfgmr} object.
}
{
  \begin{args}[pretype]
  \item[LS] (\id{SUNLinearSolver})
    the {\sunlinsolspfgmr} object to update
  \item[pretype] (\id{int})
    flag indicating the desired type of preconditioning, allowed
    values match those discussed in \id{SUNLinSol\_SPFGMR}.
  \end{args}
}
{
  This routine will return with one of the error codes
  \id{SUNLS\_ILL\_INPUT} (illegal \id{pretype}), \id{SUNLS\_MEM\_NULL}
  (\id{S} is \id{NULL}) or \id{SUNLS\_SUCCESS}.
}
{
}
% --------------------------------------------------------------------
\ucfunction{SUNLinSol\_SPFGMRSetGSType}
{
  retval = SUNLinSol\_SPFGMRSetGSType(LS, gstype);
}
{
  The function \ID{SUNLinSol\_SPFGMRSetPrecType} sets the type of
  Gram-Schmidt orthogonalization to use in the {\sunlinsolspfgmr}
  object.
}
{
  \begin{args}[gstype]
  \item[LS] (\id{SUNLinearSolver})
    the {\sunlinsolspfgmr} object to update
  \item[gstype] (\id{int})
    flag indicating the desired orthogonalization algorithm; allowed
    values are:
    \begin{itemize}
    \item \id{MODIFIED\_GS} (1)
    \item \id{CLASSICAL\_GS} (2)
    \end{itemize}
    Any other integer input will result in a
    failure, returning error code \id{SUNLS\_ILL\_INPUT}.
  \end{args}
}
{
  This routine will return with one of the error codes
  \id{SUNLS\_ILL\_INPUT} (illegal \id{pretype}), \id{SUNLS\_MEM\_NULL}
  (\id{S} is \id{NULL}) or \id{SUNLS\_SUCCESS}.
}
{
}
% --------------------------------------------------------------------
\ucfunction{SUNLinSol\_SPFGMRSetMaxRestarts}
{
  retval = SUNLinSol\_SPFGMRSetMaxRestarts(LS, maxrs);
}
{
  The function \ID{SUNLinSol\_SPFGMRSetMaxRestarts} sets the number of
  GMRES restarts to allow in the {\sunlinsolspfgmr} object.
}
{
  \begin{args}[maxrs]
  \item[LS] (\id{SUNLinearSolver})
    the {\sunlinsolspfgmr} object to update
  \item[maxrs] (\id{int})
    integer indicating number of restarts to allow.  A negative input
    will result in the default of 0.
  \end{args}
}
{
  This routine will return with one of the error codes
  \id{SUNLS\_MEM\_NULL} (\id{S} is \id{NULL}) or \id{SUNLS\_SUCCESS}.
}
{
}
% --------------------------------------------------------------------
%%
For backwards compatibility, we also provide the wrapper functions,
each with identical input and output arguments to the routines that
they wrap:
\begin{itemize}

\item \ID{SUNSPFGMR}

  Wrapper function for \ID{SUNLinSol\_SPFGMR}

\item \ID{SUNSPFGMRSetPrecType}

  Wrapper function for \ID{SUNLinSol\_SPFGMRSetPrecType}

\item \ID{SUNSPFGMRSetGSType}

  Wrapper function for \ID{SUNLinSol\_SPFGMRSetGSType}

\item \ID{SUNSPFGMRSetMaxRestarts}

  Wrapper function for \ID{SUNLinSol\_SPFGMRSetMaxRestarts}

\end{itemize}
%%
%%------------------------------------
%%
For solvers that include a Fortran interface module, the
{\sunlinsolspfgmr} module also includes a Fortran-callable function
for creating a \id{SUNLinearSolver} object.
% --------------------------------------------------------------------
\ucfunction{FSUNSPFGMRINIT}
{
  FSUNSPFGMRINIT(code, pretype, maxl, ier)
}
{
  The function \ID{FSUNSPFGMRINIT} can be called for Fortran programs
  to create a {\sunlinsolspfgmr} object.
}
{
  \begin{args}[pretype]
  \item[code] (\id{int*})
    is an integer input specifying the solver id (1 for {\cvode}, 2
    for {\ida}, 3 for {\kinsol}, and 4 for {\arkode}).
  \item[pretype] (\id{int*})
    flag indicating desired preconditioning type
  \item[maxl] (\id{int*})
    flag indicating Krylov subspace size
  \end{args}
}
{
  \id{ier} is a return completion flag equal to \id{0} for a success
  return and \id{-1} otherwise. See printed message for details in case
  of failure.
}
{
  This routine must be called \emph{after} the {\nvector} object has
  been initialized.

  Allowable values for \id{pretype} and \id{maxl} are the same as for
  the C function \ID{SUNLinSol\_SPFGMR}.
}
% --------------------------------------------------------------------
Additionally, when using
{\arkode} with a non-identity mass matrix, the {\sunlinsolspfgmr} module
includes a Fortran-callable function for creating a
\id{SUNLinearSolver} mass matrix solver object.
% --------------------------------------------------------------------
\ucfunction{FSUNMASSSPFGMRINIT}
{
  FSUNMASSSPFGMRINIT(pretype, maxl, ier)
}
{
  The function \ID{FSUNMASSSPFGMRINIT} can be called for Fortran programs
  to create a {\sunlinsolspfgmr} object for mass matrix linear systems.
}
{
  \begin{args}[pretype]
  \item[pretype] (\id{int*})
    flag indicating desired preconditioning type
  \item[maxl] (\id{int*})
    flag indicating Krylov subspace size
  \end{args}
}
{
  \id{ier} is a \id{int} return completion flag equal to \id{0} for a success
  return and \id{-1} otherwise. See printed message for details in case
  of failure.
}
{
  This routine must be called \emph{after} the {\nvector} object has
  been initialized.

  Allowable values for \id{pretype} and \id{maxl} are the same as for
  the C function \ID{SUNLinSol\_SPFGMR}.
}
% --------------------------------------------------------------------
The \id{SUNLinSol\_SPFGMRSetPrecType}, \id{SUNLinSol\_SPFGMRSetGSType}
and \id{SUNLinSol\_SPFGMRSetMaxRestarts} routines also support Fortran
interfaces for the system and mass matrix solvers

% --------------------------------------------------------------------
\ucfunction{FSUNSPFGMRSETGSTYPE}
{
  FSUNSPFGMRSETGSTYPE(code, gstype, ier)
}
{
  The function \ID{FSUNSPFGMRSETGSTYPE} can be called for Fortran
  programs to change the Gram-Schmidt orthogonaliation algorithm.
}
{
  \begin{args}[gstype]
  \item[code] (\id{int*})
    is an integer input specifying the solver id (1 for {\cvode}, 2
    for {\ida}, 3 for {\kinsol}, and 4 for {\arkode}).
  \item[gstype] (\id{int*})
    flag indicating the desired orthogonalization algorithm.
  \end{args}
}
{
  \id{ier} is a \id{int} return completion flag equal to \id{0} for a success
  return and \id{-1} otherwise. See printed message for details in case
  of failure.
}
{
  See \id{SUNLinSol\_SPFGMRSetGSType} for complete further documentation of
  this routine.
}
% --------------------------------------------------------------------
\ucfunction{FSUNMASSSPFGMRSETGSTYPE}
{
  FSUNMASSSPFGMRSETGSTYPE(gstype, ier)
}
{
  The function \ID{FSUNMASSSPFGMRSETGSTYPE} can be called for Fortran
  programs to change the Gram-Schmidt orthogonaliation algorithm for
  mass matrix linear systems.
}
{
  The arguments are identical to \id{FSUNSPFGMRSETGSTYPE} above, except that
  \id{code} is not needed since mass matrix linear systems only arise
  in {\arkode}.
}
{
  \id{ier} is a \id{int} return completion flag equal to \id{0} for a success
  return and \id{-1} otherwise. See printed message for details in case
  of failure.
}
{
  See \id{SUNLinSol\_SPFGMRSetGSType} for complete further documentation of
  this routine.
}
% --------------------------------------------------------------------
\ucfunction{FSUNSPFGMRSETPRECTYPE}
{
  FSUNSPFGMRSETPRECTYPE(code, pretype, ier)
}
{
  The function \ID{FSUNSPFGMRSETPRECTYPE} can be called for Fortran
  programs to change the type of preconditioning to use.
}
{
  \begin{args}[pretype]
  \item[code] (\id{int*})
    is an integer input specifying the solver id (1 for {\cvode}, 2
    for {\ida}, 3 for {\kinsol}, and 4 for {\arkode}).
  \item[pretype] (\id{int*})
    flag indication the type of preconditioning to use.
  \end{args}
}
{
  \id{ier} is a \id{int} return completion flag equal to \id{0} for a success
  return and \id{-1} otherwise. See printed message for details in case
  of failure.
}
{
  See \id{SUNLinSol\_SPFGMRSetPrecType} for complete further documentation of
  this routine.
}
% --------------------------------------------------------------------
\ucfunction{FSUNMASSSPFGMRSETPRECTYPE}
{
  FSUNMASSSPFGMRSETPRECTYPE(pretype, ier)
}
{
  The function \ID{FSUNMASSSPFGMRSETPRECTYPE} can be called for Fortran
  programs to change the type of preconditioning for mass matrix
  linear systems.
}
{
  The arguments are identical to \id{FSUNSPFGMRSETPRECTYPE} above, except that
  \id{code} is not needed since mass matrix linear systems only arise
  in {\arkode}.
}
{
  \id{ier} is a \id{int} return completion flag equal to \id{0} for a success
  return and \id{-1} otherwise. See printed message for details in case
  of failure.
}
{
  See \id{SUNLinSol\_SPFGMRSetPrecType} for complete further documentation of
  this routine.
}
% --------------------------------------------------------------------
\ucfunction{FSUNSPFGMRSETMAXRS}
{
  FSUNSPFGMRSETMAXRS(code, maxrs, ier)
}
{
  The function \ID{FSUNSPFGMRSETMAXRS} can be called for Fortran programs
  to change the maximum number of restarts allowed for {\spfgmr}.
}
{
  \begin{args}[maxrs]
  \item[code] (\id{int*})
    is an integer input specifying the solver id (1 for {\cvode}, 2
    for {\ida}, 3 for {\kinsol}, and 4 for {\arkode}).
  \item[maxrs] (\id{int*})
    maximum allowed number of restarts.
  \end{args}
}
{
  \id{ier} is a \id{int} return completion flag equal to \id{0} for a success
  return and \id{-1} otherwise. See printed message for details in case
  of failure.
}
{
  See \id{SUNLinSol\_SPFGMRSetMaxRestarts} for complete further
  documentation of this routine.
}
% --------------------------------------------------------------------
\ucfunction{FSUNMASSSPFGMRSETMAXRS}
{
  FSUNMASSSPFGMRSETMAXRS(maxrs, ier)
}
{
  The function \ID{FSUNMASSSPFGMRSETMAXRS} can be called for Fortran
  programs to change the maximum number of restarts allowed for
  {\spfgmr} for mass matrix linear systems.
}
{
  The arguments are identical to \id{FSUNSPFGMRSETMAXRS} above, except that
  \id{code} is not needed since mass matrix linear systems only arise
  in {\arkode}.
}
{
  \id{ier} is a \id{int} return completion flag equal to \id{0} for a success
  return and \id{-1} otherwise. See printed message for details in case
  of failure.
}
{
  See \id{SUNLinSol\_SPFGMRSetMaxRestarts} for complete further
  documentation of this routine.
}
% --------------------------------------------------------------------


%---------------------------------------------------------------------------
\subsection{{\sunlinsolspfgmr} description}\label{ss:sunlinsol_spfgmr_description}


The {\sunlinsolspfgmr} module defines the {\em content} field of a
\id{SUNLinearSolver} to be the following structure:
%%
\begin{verbatim}
struct _SUNLinearSolverContent_SPFGMR {
  int maxl;
  int pretype;
  int gstype;
  int max_restarts;
  int numiters;
  realtype resnorm;
  long int last_flag;
  ATimesFn ATimes;
  void* ATData;
  PSetupFn Psetup;
  PSolveFn Psolve;
  void* PData;
  N_Vector s1;
  N_Vector s2;
  N_Vector *V;
  N_Vector *Z;
  realtype **Hes;
  realtype *givens;
  N_Vector xcor;
  realtype *yg;
  N_Vector vtemp;
};
\end{verbatim}
%%
These entries of the \emph{content} field contain the following
information:
\begin{description}
  \item[maxl] - number of FGMRES basis vectors to use (default is 5),
  \item[pretype] - flag for use of preconditioning (default is none),
  \item[gstype] - flag for type of Gram-Schmidt orthogonalization
    (default is modified Gram-Schmidt),
  \item[max\_restarts] - number of FGMRES restarts to allow
    (default is 0),
  \item[numiters] - number of iterations from the most-recent solve,
  \item[resnorm] - final linear residual norm from the most-recent solve,
  \item[last\_flag] - last error return flag from an internal function,
  \item[ATimes] - function pointer to perform $Av$ product,
  \item[ATData] - pointer to structure for \id{ATimes},
  \item[Psetup] - function pointer to preconditioner setup routine,
  \item[Psolve] - function pointer to preconditioner solve routine,
  \item[PData] - pointer to structure for \id{Psetup} and \id{Psolve},
  \item[s1, s2] - vector pointers for supplied scaling matrices
    (default is \id{NULL}),
  \item[V] - the array of Krylov basis vectors
    $v_1, \ldots, v_{\text{\id{maxl}}+1}$, stored in \id{V[0]},
    \ldots, \id{V[maxl]}. Each $v_i$ is a vector of type {\nvector}.,
  \item[Z] - the array of preconditioned Krylov basis vectors
    $z_1, \ldots, z_{\text{\id{maxl}}+1}$, stored in \id{Z[0]},
    \ldots, \id{Z[maxl]}. Each $z_i$ is a vector of type {\nvector}.,
  \item[Hes] - the $(\text{\id{maxl}}+1)\times\text{\id{maxl}}$
    Hessenberg matrix. It is stored row-wise so that the (i,j)th
    element is given by \id{Hes[i][j]}.,
  \item[givens] - a length \id{2*maxl} array which represents the
    Givens rotation matrices that arise in the FGMRES algorithm. These
    matrices are $F_0, F_1, \ldots, F_j$, where
    $F_i = \begin{bmatrix}
      1 &        &   &     &      &   &        &   \\
        & \ddots &   &     &      &   &        &   \\
        &        & 1 &     &      &   &        &   \\
        &        &   & c_i & -s_i &   &        &   \\
        &        &   & s_i &  c_i &   &        &   \\
        &        &   &     &      & 1 &        &   \\
        &        &   &     &      &   & \ddots &   \\
        &        &   &     &      &   &        & 1\end{bmatrix}$,
    are represented in the \id{givens} vector as \id{givens[0] =}
    $c_0$, \id{givens[1] = } $s_0$, \id{givens[2] = } $c_1$,
    \id{givens[3] = } $s_1$, \ldots \id{givens[2j] = } $c_j$,
    \id{givens[2j+1] = } $s_j$.,
  \item[xcor] - a vector which holds the scaled, preconditioned
    correction to the initial guess,
  \item[yg] - a length \id{(maxl+1)} array of \id{realtype} values
    used to hold ``short'' vectors (e.g. $y$ and $g$),
  \item[vtemp] - temporary vector storage.
\end{description}

This solver is constructed to perform the following operations:
\begin{itemize}
\item During construction, the \id{xcor} and \id{vtemp} arrays are
  cloned from a template {\nvector} that is input, and default solver
  parameters are set.
\item User-facing ``set'' routines may be called to modify default
  solver parameters.
\item Additional ``set'' routines are called by the {\sundials} solver
  that interfaces with {\sunlinsolspfgmr} to supply the
  \id{ATimes}, \id{PSetup}, and \id{Psolve} function pointers and
  \id{s1} and \id{s2} scaling vectors.
\item In the ``initialize'' call, the remaining solver data is
  allocated (\id{V}, \id{Hes}, \id{givens}, and \id{yg} )
\item In the ``setup'' call, any non-\id{NULL}
  \id{PSetup} function is called.  Typically, this is provided by
  the {\sundials} solver itself, that translates between the
  generic \id{PSetup} function and the
  solver-specific routine (solver-supplied or user-supplied).
\item In the ``solve'' call, the FGMRES iteration is performed.  This
  will include scaling, preconditioning, and restarts if those options
  have been supplied.
\end{itemize}

%%
%%----------------------------------------------
%%

\noindent The {\sunlinsolspfgmr} module defines implementations of all
``iterative'' linear solver operations listed in Sections
\ref{ss:sunlinsol_CoreFn}-\ref{ss:sunlinsol_GetFn}:
\begin{itemize}
\item \id{SUNLinSolGetType\_SPFGMR}
\item \id{SUNLinSolInitialize\_SPFGMR}
\item \id{SUNLinSolSetATimes\_SPFGMR}
\item \id{SUNLinSolSetPreconditioner\_SPFGMR}
\item \id{SUNLinSolSetScalingVectors\_SPFGMR}
\item \id{SUNLinSolSetup\_SPFGMR}
\item \id{SUNLinSolSolve\_SPFGMR}
\item \id{SUNLinSolNumIters\_SPFGMR}
\item \id{SUNLinSolResNorm\_SPFGMR}
\item \id{SUNLinSolResid\_SPFGMR}
\item \id{SUNLinSolLastFlag\_SPFGMR}
\item \id{SUNLinSolSpace\_SPFGMR}
\item \id{SUNLinSolFree\_SPFGMR}
\end{itemize}
