\subsection{KINSOL}\label{ss:kinsol}

KINSOL solves nonlinear algebraic systems in real space, which we write as
\begin{equation}\label{nonlinear_system}
  F(u) = 0 \, , \quad F:\mbox{\bf R}^N \rightarrow \mbox{\bf R}^N \, , 
\end{equation}
given an initial guess $u_0$.  It is a rewrite in C of the Fortran 77
code NKSOL of Brown and Saad \cite{BrSa:90}.

KINSOL employs the Inexact Newton method developed in 
\cite{BrSa:90,Bro:87,DES:82} 
and further described in \cite{DeSc:96,Kel:95}, 
resulting in the following iteration:

\vspace{1ex}
\noindent{\bf Inexact Newton iteration}
\begin{enumerate}
   \item Set $u_0 = $ an initial guess
   \item For $n = 0, 1, 2,...$ until convergence do:
      \begin{itemize}
          \item[(a)] Solve $J(u_n)\delta_n = - F(u_n)$ 
          \item[(b)] Set $u_{n+1} = u_n + \delta_n$
          \item[(c)] Test for convergence
      \end{itemize}
\end{enumerate}
Here, $J(u) = F'(u)$ is the system Jacobian. As this code module is
anticipated for use on large systems, only iterative methods are
provided to solve the system in step 2(a). These solutions are only
approximate.  At each stage in the iteration process, a scalar
multiple of the approximate solution, $\delta_n$, is added to the
previous approximate solution, $u_n$, to produce a new approximate
solution. A test for convergence is made before the iteration
continues.

The linear iterative method currently implemented is one of the class of 
Krylov methods, GMRES \cite{BrHi:89,SaSc:86}, 
provided through the SPGMR module common to all SUNDIALS codes. 
Use of SPGMR provides a linear solver which, by default, is applied in a 
matrix-free manner, with matrix-vector products $Jv$ obtained by either 
finite difference quotients or a user-supplied routine.
In the case where finite differences are used, 
the matrix-vector product $J(u)v$ is approximated by a quotient of the form
given in (\ref{jacobv}),
where $f(t, y) = F(y)$ for our nonlinear system, 
$u$ is the current approximation to a root of (\ref{nonlinear_system}), 
and $\sigma$ is a scalar.  The choice of $\sigma$ is taken from 
\cite{BrSa:90} and is given by
\begin{equation}\label{sigma_comp}
  \sigma = \frac{\max \{|u^T v|, \mbox{typ}u^T |v|\}}{\|v\|_2} 
  \mbox{sign}(u^T v) \sqrt{U} \, ,
\end{equation}
where $\mbox{typ}u$ is a vector of typical values for the absolute
values of the solution (and can be taken to be inverses of the scale
factors given for $u$ as described below), and $U$ is unit roundoff.
Convergence of the Newton method is maintained as long as the value of
$\sigma$ remains appropriately small as shown in \cite{Bro:87}.

To the above methods are added scaling and preconditioning. Scaling is allowed
for both the approximate solution vector and the system function vector.
For scaling to be used, the user should supply values $D_u$, which
are diagonal elements of the scaling matrix such that $D_u u_n$ has all
components roughly the same magnitude when $u_n$ is close to a solution,
and $D_F F$ has all components roughly the same magnitude when $u_n$ is not
too close to a solution.
In the text below, we use the following scaled norms:
\begin{equation}
\|z\|_{D_u} = \|D_u z\|, \;\; {\rm and} \;\; \|z\|_{D_F} = \|D_F z\|.
\end{equation}
When scaling values are provided for the solution vector, these values are 
automatically incorporated into the calculation of $\sigma$ in
(\ref{sigma_comp}).
Additionally, right preconditioning is provided if the preconditioning 
setup and solve routines are supplied by the user.  In this case,
GMRES is applied to the linear systems $(JP^{-1})(P\delta) = -F$.

Two methods of applying a computed step $\delta_n$ to the previously computed
approximate solution vector are implemented.  The 
first and simplest is the Inexact Newton strategy which applies step 2(b) 
as above.  
The other method is a global strategy,
which attempts to use the direction implied by $\delta_n$ 
in the most efficient way for furthering convergence of the nonlinear problem. 
This technique 
is implemented in the second strategy, called Linesearch.  This option
employs both the $\alpha$ and $\beta$ conditions of the Goldstein-Armijo
linesearch given in \cite{DeSc:96} and changes step 2(b) to
\begin{enumerate}
  \addtocounter{enumi}{1}
\item{(b)$\;$Set $u_{n+1} = u_n + \lambda \delta_n$},
\end{enumerate}
where $\lambda$ is chosen to guarantee a sufficient decrease in $F$ 
relative to the step length as well as a minimum step length relative 
to the initial rate of decrease of $F$.  The full Newton step is taken
close to the solution.

Stopping criteria for the Newton method can be required for either or 
both of the nonlinear residual and the step length.  For the former, 
the Newton iteration must pass a stopping test
\[ \|F(u_n)\|_{D_F,\mbox{\scriptsize max}} < \mbox{\sc ftol} \, , \]
where {\sc ftol} is an input scalar tolerance with a default value of
$U^{1/3}$. 
For the latter, the Newton method will terminate when the maximum scaled step
is below a given tolerance
\[ \|\delta_n\|_{D_u,\mbox{\scriptsize max}} < \mbox{\sc steptol} \, , \]
where {\sc steptol} is an input scalar tolerance with a default value of 
$U^{2/3}$.

Three options for stopping criteria for the linear system solve are
implemented, including the two 
algorithms of Eisenstat and Walker \cite{EiWa:96}.
The Krylov iteration must pass a stopping test
\[ \|J \delta_n + F\|_{D_F} < (\eta_n + U) \|F\|_{D_F} \, , \]
where $\eta_n$ is one of:
\begin{itemize}
\item Eisenstat and Walker Choice 1
  \[ 
  \eta_n = \frac{\left|\; \|F(u_n)\|_{D_F}  
      - \|F(u_{n-1}) + J(u_{n-1}) \delta_n \|_{D_F} 
      \; \right|}
  {\|F(u_{n-1})\|_{D_F}} \, ,  
  \]
\item Eisenstat and Walker Choice 2
  \[ 
  \eta_n = \gamma 
  \left( \frac{ \|F(u_n)\|_{D_F}}{\|F(u_{n-1})\|_{D_F}} \right)^{\alpha} \, ,
  \] 
where default values of $\gamma$ and $\alpha$ are $0.9$ and $2$,
 respectively. 
\item  $\eta_n$ = constant with 0.1 as the default.
\end{itemize}
The default is Eisenstat and Walker Choice 1. For both options 1 and 2, 
appropriate safeguards are incorporated to ensure that $\eta$ does not 
decrease too fast~\cite{EiWa:96}.

As a user option, KINSOL permits the application of inequality
constraints, either $u^i > 0$ or $u^i < 0$.  Either constraint, or
no constraint, may be imposed on each component.  KINSOL will reduce 
step lengths in order to ensure that no constraint is violated.


