% LaTeX source for joint paper on SUNDIALS
% Version of 7 March 2003

\documentclass[11pt]{article}

\usepackage{epsfig}

\parskip=12pt
\font\bftitle=cmbx10 scaled\magstep1            %title headings
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9.0in}
\setlength{\topmargin}{-0.5in}


\title{SUNDIALS: SUite of Nonlinear and DIfferential/ALgebraic\\
       equation Solvers}
\author{Peter N.~Brown \and Keith E.~Grant \and Alan C.~Hindmarsh 
\and Steven L.~Lee \and Radu Serban \and Dan E.~Shumaker 
\and Carol S.~Woodward}


\begin{document}


% Start of page 1 of paper

\maketitle

\setcounter{page}{1}

\input{abstract.tex}

\noindent {\bf Keywords:} ODEs, DAEs, Nonlinear Systems,
Sensitivity Analysis


\input{introduction.tex}
\input{solvers_section.tex}
   \input{CVODE_subsection.tex}
   \input{KINSOL_subsection.tex}
   \input{IDA_subsection.tex}
\input{preconditioning.tex}
\input{sensitivity_analysis.tex}
\input{organization.tex}
\input{usage.tex}
\input{applications.tex}
\input{availability.tex}


\section{Acknowledgements}

The authors wish to acknowledge the contributions of Scott Cohen
in the development of CVODE and Allan Taylor in development of
KINSOL and IDA.  This work was performed under the auspices of the
U.S. Department of Energy by the University of California,
Lawrence Livermore National Laboratory, under contract No.
W-7405-Eng-48.


\begin{thebibliography}{99}

\bibitem{BrCaPe96}
K. E. Brenan, S. L. Campbell and L. R. Petzold, {\em Numerical Solution
of Initial-Value Problems in Differential-Algebraic Equations}, second
edition, SIAM, 1996.

\bibitem{VODE}  P. N. Brown, G. D. Byrne, and A.C. Hindmarsh, {\it VODE, a
Variable-Coefficient ODE Solver}, SIAM J. Sci. Stat. Comput., {\bf 10}
(1989), 1038-1051.

\bibitem{DASPK}
P. N. Brown, A. C. Hindmarsh, and L. R. Petzold, Using Krylov
Methods in the Solution of Large-Scale Differential-Algebraic
Systems, SIAM J. Sci. Comp., 15 (1994), pp. 1467-1488.

\bibitem{DASPKIC}
P. N. Brown, A. C. Hindmarsh, and L. R. Petzold, Consistent
Initial Condition Calculation for Differential-Algebraic
Systems, SIAM J. Sci. Comp., 19 (1998), pp. 1495-1512.

\bibitem{NKSOL}
P. N. Brown and Y. Saad, {\it Hybrid Krylov methods for nonlinear
systems of equations}, SIAM J. Sci. Stat. Comput., 11 (1990),
pp. 450-481.

\bibitem{VODPK}  George D. Byrne, {\it Pragmatic Experiments with Krylov
Methods in the Stiff ODE Setting}, in {\it Computational Ordinary
Differential Equations}, J. R. Cash and I. Gladwell (Eds.), Oxford
University Press, Oxford, 1992, pp. 323-356.

\bibitem{PVUdoc}
George D. Byrne and Alan C. Hindmarsh, ``User Documentation for PVODE,
An ODE Solver for Parallel Computers," LLNL Report UCRL-ID-130884,
May 1998.

\bibitem{PVODE}
George D. Byrne and Alan C. Hindmarsh, ``PVODE, An ODE Solver for
Parallel Computers," in Int. J. High Perf. Comput. Applic., vol. 13,
no. 4 (1999), pp. 354-365.

\bibitem{CVUdoc}
S. D. Cohen and A. C. Hindmarsh, ``CVODE User Guide," LLNL Report
UCRL-MA-118618, Sept. 1994.

\bibitem{CVODE}
Scott D. Cohen and Alan C. Hindmarsh, ``CVODE, a Stiff/Nonstiff ODE
Solver in C," Computers in Physics, vol. 10, no. 2 (March/April 1996),
pp. 138-143.

\bibitem{EiWa96}
S. C. Eisenstat and H. F. Walker, {\it Choosing the Forcing Terms in
an Inexact Newton Method}, SIAM J. Sci. Comput., 17 (1996), pp. 16-32.

\bibitem{SensKIN}
Keith E. Grant, Alan C. Hindmarsh, and Allan G. Taylor, ``User
Documentation for SensKINSOL, a Variant of KINSOL for Sensitivity
Analysis,'' in process.

\bibitem{PVIAlgs}
Alan C. Hindmarsh, ``The PVODE and IDA Algorithms", LLNL Report
UCRL-ID-141558, December 2000.

\bibitem{CVSUdoc}
Alan C. Hindmarsh and Radu Serban, ``User Documentation for
CVODES, An ODE Solver with Sensitivity Analysis Capabilities,''
LLNL Report UCRL-MA-148813, July 2002.

\bibitem{IDAUdoc}
Alan C. Hindmarsh and Allan G. Taylor, ``User Documentation for IDA,
a Differential-Algebraic Equation Solver for Sequential and Parallel
Computers," LLNL Report UCRL-MA-136910, December 1999.

\bibitem{SensPV}
S. L. Lee, A. C. Hindmarsh, and P. N. Brown, ``User Documentation for
SensPVODE, A Variant of PVODE for Sensitivity Analysis,'' LLNL
Report UCRL-MA-140211, August 22, 2000.

\bibitem{KINUdoc}
Allan G. Taylor and Alan C. Hindmarsh, ``User Documentation for KINSOL,
A Nonlinear Solver for Sequential and Parallel Computers," LLNL Report
UCRL-ID-131185, July 1998.

\bibitem{PVTests}
Michael R. Wittman, ``Testing of PVODE, a Parallel ODE Solver,"
LLNL Report UCRL-ID-125562, August 1996.  Compares parallel CVODE with
SHMEM and two versions of MPI.

\end{thebibliography}


\end{document}
